### 2019-08-24 abstract.tex
first commit

This system was \#1 in the world for a very long time, and has remained
a critically important computer system. It defied scale with 18,688
individual compute nodes, and delivered tens of billions of computing
hours to DOE mission-critical programs for nearly 7 years.

### 2019-08-24 acknowledgments.tex
first commit

\section{Acknowledgments}
\label{sec:ack}
\fix{Standard and all}

### 2019-08-24 dataclean.tex
first commit

\section{Viewing and Cleaning the Data}
\label{sec:clean}
\fix{George's section: The two views of GPU life: by SN and by
  location, its use in cleaning data, etc.}

A typical GPU data record produced by the methods in the last section
is of the following form
{\small
\begin{verbatim}
0323812007945 | c17-4c1s3n1 | 01/10/2013 02:54:45 | 02/02/2013 11:32:29
 | c13-1c1s3n3 | 01/21/2014 21:10:42 | 08/01/2017 02:15:02
 | c0-1c1s3n3 | 10/11/2013 15:57:33 | 10/12/2013 22:09:31
 | c21-1c2s5n0 | 03/19/2013 15:48:11 | 05/29/2013 11:54:11
0325216047736 | c18-4c1s5n1 | 03/07/2017 02:15:01 | 03/12/2019 03:19:11
0323812008856 | c5-4c0s7n0 | 01/10/2013 02:54:45 | 01/25/2013 15:29:58
 | c0-6c1s7n2 | 10/21/2013 14:28:19 | 10/28/2013 17:52:44
 | c3-3c1s5n0 | 05/29/2013 11:54:11 | 05/29/2013 11:54:11
 | c23-6c1s7n2 | 01/21/2014 21:10:42 | 11/02/2018 14:42:34
 | | DBE | 11/02/2018 14:42:34
\end{verbatim}
}
\noindent where records for three GPUs are shown. Each record starts
with a serial number and locations are coded with {\tt c{\it
    col-row}c{\it cage}s{\it slot}n{\it node}}. The first GPU record
shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},
{\tt c0-1c1s3n3}, with periods off the system, and finally in {\tt
  c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
not seen again. The second is intalled in location {\tt c18-4c1s5n1},
where it stays until the last data collection date on March 3, 2019,
at 3:19:11 AM. The third GPU is first installed in locations {\tt
  c5-4c0s7n0}, {\tt c3-3c1s5n0},{\tt c0-6c1s7n2}, {\tt c23-6c1s7n2},
where a ``DBE'' is observed on November 2, 2018, and it is not seen
again.

As we need to recover durations from this data, correct processing
involves time adjustments for switching between daylight saving time
and standard time. We perform this by setting a reference time zone
(Eastern time) and converting all date-times from strings into
date-time variables \fix{(is there a name for this?)} with the R
\pkg{lubridate} package \cite{lubridate}, which enables appropriate
date arithmetic.

\fix{Other processing:} enable missing values, and create some new
variables. The event is “life” when both insert and remove are
present, “life0” when insert = remove, otherwise it is whatever string
is in insert, which is “DBE” or “Off The BUS”. Also separate location
into its components and fill in (repeat) serial numbers and locations
for all records.

Reduce to only “life” records and mark them with ending DBE, OTB,
“out”, or none events. The result is just insert and remove with event
marks to indicate DBE or OTB or “out” (last seen). bad indicates that
a DBE or OTB occurred yet the GPU was not taken out. For a first cut,
keep the bad ones in. Also keep duration to later aggregate into full
life times.

Next, aggregate into one record per serial number with total life time
and first insert time. For proper censoring treatment: Indicate if
event occurred or still in service: out = fail, dbe = fail, otb =
fail, NA = right censored (still in service). Add some other
variables, such as GPU in single location or moved one or more times,
mark new\_batch group, etc. to use in modeling later.

A plot of the life of 100 random GPUs on Titan after cleaning the data:
\begin{description}
\item[starts:] black dots (gray for zero lifetime)
\item[lifetimes:] black lines
\item[DBEs:] red triangles
\item[Off The BUS:] blue triangles
\item[Last seen:] black ]
\item[overlaps:] red square on left when they exist
\end{description}

\begin{figure}[ht]
  \includegraphics[width=6in]{sn.pdf}
  \caption{GPU serial number view of life and failures.}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=6in]{location.pdf}
  \caption{GPU location view of life and failures.}
\end{figure}

### 2019-08-24 dataprep.tex
first commit

\section{Data Collection and Preprocessing}
\label{sec:data}
\fix{Don's section}

### 2019-08-24 discussion.tex
first commit

\section{Discussion}
\label{sec:discussion}
\fix{Jim and everyone}

### 2019-08-24 intro.tex
first commit

\section{Introduction}
\label{sec:intro}
\fix{Dropped Jim's initial email here. Mostly Jim and Christian}
The Cray XK7 Titan is rapidly nearing its decommissioning date for
DOE/ASCR. The last production runs will be in July of this year, and
then we will stop services, clean up a few things and turn it off.

This system was \#1 in the world for a very long time, and has remained
a critically important computer system. It defied scale with 18,688
individual compute nodes, and delivered tens of billions of computing
hours to DOE mission-critical programs for nearly 7 years.

This was quite an interesting machine from a reliability perspective.
We were forced to execute three very significant rework cycles, two on
the mechanical assembly affecting the PCIe connector from the GPU
daughtercard to the motherboard, and more recently, a replacement of
about 11,000 GPU assemblies because of a failing resistor on a printed
circuit board.

We have conducted a good bit of statistical analysis on the failure
rates of these GPU assemblies, but I want to propose a different topic
that I have not seen for other/similar systems.  I want to complete
the statistical analysis of the failure rates as we near end of life,
looking for the opposite side of the “bathtub curve”.  There is
specific decay associated with the electronic components where, as
these products age, they will eventually fail thresholds for guard
band margin and other things. In a lot of cases, they will just fail
as well.

This machine is quite large, was very important to many people, and I
think this analysis could be very intriguing.

I am looking for help, very specifically with the statistics portion
of this, i.e. what do the failure rates for specific components look
like? What distributions do those failure rates follow? How could
appropriate models extrapolate failure rates beyond the end of service
date (the dotted lines…) ?

I have submitted an abstract to a Cray-focused user group and workshop
on this topic, and it has been accepted.  I would like to execute this
statistical analysis and generate a short presentation (30 minutes)
and a short paper (4-8 pages is a likely target).  Presentation is
targeted for May 7-9 (TBD). An initial/working version of the paper
would be due earlier. If we miss that target (April 12), no big
deal. We can finish on a similar schedule as the presentation.  I have
all of the raw data for the component failures, and need the
statistical help to complete the analysis.

Because there is a huge swing in the failure rates of the GPUS from
2016/17 when the resistor problem escalated rapidly out of control, it
is statistically interesting, where we inject fresh material in to an
existing machine. We’d likely want to categorically separate new GPU
assemblies from old and could track the early life failures from those
new 11,000 parts versus the continued failure of the remaining 6000.

George, Christian-  any interest in this? Barney gave it a thumbs
up. I think this is a timely and very interesting topic.

### 2019-08-24 related.tex
first commit

\section{Related Work}
\label{sec:related}
\fix{Christian's section}

### 2019-08-24 survival.tex
first commit

\section{Survival Analysis}
\label{sec:survival}
\fix{George's survival analysis}

Kaplan-Meier survival analysis gives the probability of survival beyond a given time. It is a nonparametric technique that makes no specific failure model assumptions, such as Weibull, Exponential, etc. It can also split the data into subpopulations and compute separate survival curves.

If $T$ is taken to be the random variable of a GPU fail time, then its probability density function is $f(t)$ and its cumulative distribution function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails by duration $t$. The survival function is then
\begin{displaymath}
  S(t) = Pr\{T \geq t\} = 1 - F(t) = \int_t^\infty f(x)dx,
\end{displaymath}
which is the probability of being operational just before duration $t$.

A sister concept to the survival function is the hazard function:
\begin{displaymath}
  H(t) = \frac{f(t)}{S(t)},
\end{displaymath}
which is what is sometimes called the "bathtub" curve in reliability.

The Cox proportional hazards (CPH) regression analys takes the hazard function above and a set of covariates $x$ and models $k$s GPU hazard as
\begin{displaymath}
  H_k(t) = H_0(t)e^{\sum\limits_{i=1}^n{\beta x}}
\end{displaymath}
That is, a base hazard rate multiplied by a function of
covariates. The key assumption in the CPH model is that the "hazards"
are multipliers on the "baseline hazard" but the estimate of the
baseline hazard is nonparametric, meaning it makes no specific
distributional assumption and just learns from the data (unlike
analyses that assume for example a Weibul or an exponential
model). The multiplicative assumption is that the hazard curves do not
cross and are multiples of each other. The multiplier is the *hazard
coefficient*. For example, if the baseline is *node0* and the *node2*
hazard is 2, then *node2* sees twice as many failures as *node0* on
average.

For our comparison of hazards, we use as *covariates* all the unpacked
location variables, indicators whether the GPU has been moved, and our
notion of old and new GPUs based on first install date. Since many
GPUs experience time at more than one location, we use the longest
period location in this analysis.

First, we do a univariate analysis for each covariate separately and
then we look at a multivariate view of this. Having multiple
covariates is called "multiple regression" as opposed to "multivariate
regression" reserved for cases with several response variables. The
univariate view is comparable to what would one see in a simple time
between failures analysis for a given population split. It can be
affected by sampling differences for other covariates. For example, if
the strongest effect is seen in a variable not used in a split and its
values are unevenly distributed across the split, its effect can leak
into the split. This is why we also do a multivariate analysis, where
the effect of each covariate is adjusted for all the other
covariates.

\begin{figure}
  \includegraphics[width=5in]{cox.pdf}
  \caption{GPU hazard ratios from Cox regression model.}
\end{figure}

Higher numbered columns appear to have a higher (up to about 2x)
failure rate. There actually seems to be a hint of three groups that
have similar rates: 0-6 baseline, 7-16 are 1.5x baseline, and 17-24
are 2x baseline.

Higher numbered rows tend to have a lower (up to about 0.5x) failure rate.

This seems surprisingly strong! Higher numbered cages have failure
rates 5x and 16x higher than cage = 0. In fact, this seems so strong
that its effect could be leaking into all other comparisons under
uneven sampling!

Not much difference between slots. They are all close to the baseline
slot0.

Interestingly, nodes 2 and 3 have half the failure rate of nodes 0 and
1. Is this related to the cage variable or is there a real location
effect?

Both loc and move are roughly the same, indicating that GPUs that were
moved to another location had slighly (.75x) lower failure rate. This
may just reflect a “repair effect” that extends the life of a GPU. It
may have been removed and reinstalled for a reason that was not
recorded.

Oddly, the older batch has a very slighly lower hazard, although there
is a lot of uncertainty here, indicating that something more
complicated is at work and needs deeper investigation.

Now the multivariate Cox regression analysis, where each factor is
adjusted for all the other factors. This can fix some uneven sampling
and multivariate dependence.

The col groups persist and are even stronger. The row effect is also
stronger. The cage effect is out of this world! slot is still without
any effect. node effect also persists and is a bit stronger.

It is strange that move and loc effects are opposite, whereas in the
univariate analysis they were the same. They are also much
stronger. The batch effect is opposite of what it was under univariate
analysys and a bit stronger as well. This is a concern and it may be
related to violation of the “proportionality assumption” of hazards
not crossing. We see some crossing in the survival curves above
already.

Let’s separate the “old” and “new” batches of GPUs and run separate
Cox models.

\begin{figure}
  \includegraphics[width=5in]{glmnet.pdf}
  \caption{GPU variable importance from glmnet penalized regression analysis.}
\end{figure}

### 2019-08-24 tbf.tex
first commit

\section{Time Between Failures Analysis}
\label{sec:tbf}
\fix{Rizwan's section}

### 2019-12-24 survival.tex
revised assignments

@@
 \label{sec:survival}
-\fix{George's survival analysis}
+\fix{George's survival analysis, check \protect\pkg{finalfit}}
 
@@
 
-The Cox proportional hazards (CPH) regression analys takes the hazard function above and a set of covariates $x$ and models $k$s GPU hazard as
+The Cox proportional hazards (CPH) regression analys \cite{Cox1972,Harrell2015} takes the hazard function above and a set of covariates $x$ and models $k$s GPU hazard as
 \begin{displaymath}

### 2020-01-03 related.tex
Started adding related work and references

@@
 \fix{Christian's section}
+
+Prior work in investigating Titan's reliability includes a significant amount of work that covers different aspects.
+%
+Tiwari et al~\cite{7056044} analyzed the types of GPU errors on Titan and used neutron beam experiments to further understand GPU soft errors.
+%
+Nie et al.~\cite{nie17characterizing,nie18machine} characterized the soft error behavior of Titan's GPUs in relation to temperature and power consumption to predict their increased occurrence by correlating data in temporal and spatial domains and using machine learning models.
+%
+Tiwari et al. and Nie et al. did not address the failure mode discussed in this paper.
+%
+Zimmer et al.~\cite{8665764} developed a new job scheduling strategy for Titan to counter the GOU failures discussed in this paper and to improve system utilization and productivity.
+
+Work by Kumar et al.~\cite{kumar18understanding} analyzed Titan's interconnect faults, errors and congestion events to improve interconnects resilience and congestion resolution methods.
+%
+Gupta et al.~\cite{gupta17failures} performed a study covering five supercomputers at Oak Ridge National Laboratory, including Titan, that concentrated on developing an understanding of which and how many errors and failure occur in supercomputers over multiple generations and over their years of operation.
+%
+Bautista-Gomez et al.~\cite{bautista-gomez16reducing} used failure data from Titan to adapt the checkpoint frequency to the current reliability of the system using dynamic checkpointing.
+%
+Tiwari et al.~\cite{6903564} also used failure data from Titan and exploited temporal locality in the data to adapt the checkpoint frequency to the current reliability of the system using lazy checkpointing.

### 2020-01-06 related.tex
Added slightly more related work and references

@@
 %
-Zimmer et al.~\cite{8665764} developed a new job scheduling strategy for Titan to counter the GOU failures discussed in this paper and to improve system utilization and productivity.
+Zimmer et al.~\cite{8665764} developed a new job scheduling strategy for Titan to counter the GPU failures discussed in this paper and to improve system utilization and productivity.
 
@@
 Tiwari et al.~\cite{6903564} also used failure data from Titan and exploited temporal locality in the data to adapt the checkpoint frequency to the current reliability of the system using lazy checkpointing.
+
+Other work in characterizing supercomputer faults, errors and failures focused on the various systems deployed around the world, primarily at US Department of Energy laboratories.
+%
+De et al.~\cite{8809553} developed an in-depth understanding of the failure characteristics of the IBM Blue Gene/Q Mira system at Argonne National Laboratory.

### 2020-01-09 dataprep.tex
data prep draft

@@
 \fix{Don's section}
+Data on GPU life times is constructed from two sources: inventory runs
+and failure event records (DBX and Off the Bus).
+
+An inventory of GPU serial numbers and their locations is recorded
+each time the system boots. This typically occurs once every few days but
+can be as frequent as a few times per day. An inventory at boot time
+typically occur-rs over a span of \fix{??} minutes. A separate file is
+produced at each boot time.
+
+The DBE and Off The BUS events are also recorded separately from boot
+time inventories. This is recorded as . . . files.
+
+Initial processing updates an inventory by checking the Serial Number
+(SN) and Location of each GPU and creating or updating a separate
+record for each contiguously observed SN-Location combination.
+
+Figure~\ref{fig:inventory} shows inventory times that appear in the
+data grouped into three categories. Only the time axis is relevant
+here as the vertical scatter within each category is added only to
+control overplotting. It appears that the early period before mid 2015
+has relatively few inventories taken, inventory frequency increases
+through the end of 2016, then it decreases slightly and becomes
+steady. There is also a nearly year-long gap in Off The BUS
+occurrences. After removing the OTB and DBE dates, the time
+distribution becomes fairly uniform. This suggests that the OTB and
+DBE times are independent of the boot time inventories.
+
+\fix{Don, the figures bring more questions: Are the OTB and DBE events
+recorded independently of the boot time inventories? The figures are
+the distributions of referenced times in the data. Separate graphs are
+done for OTB and DBE events, for general life span insert and remove
+dates, and again after removing the OTB and DBE times. After the
+removal, the times are fairly uniformly distributed. Does that mean
+that boot time inventories are done fairly uniformly?}
+
+\begin{figure}[bth]
+  \centering
+  \includegraphics[width=\textwidth]{inventory_times.pdf}
+  \caption{Inventory times in all of the data separated into ``DBE''
+    and ``Off The BUS'' failure events, and observed insert-remove
+    times as ``life'' and again after removing the failure times as
+    ``life - otb - dbe''. Vertical point jitter is applied to each to
+    mitigate overplotting effects.}
+  \label{fig:inventory_dates}
+\end{figure}
+\begin{figure}[bth]
+  \centering
+  \includegraphics[width=\textwidth]{inventory_density.pdf}
+  \caption{Inventory times density in all of the data separated into
+    ``DBE'' and ``Off The BUS'' failure events, and observed
+    insert-remove times as ``life'' and again after removing the
+    failure times as ``life - otb - dbe''. Compare to jittered dates
+    in Fig.~\ref{fig:inventory_dates}.}
+  \label{fig:inventory_density}
+\end{figure}

### 2020-01-15 related.tex
Added more explanations to the existing related work.

@@
 %
-Tiwari et al~\cite{7056044} analyzed the types of GPU errors on Titan and used neutron beam experiments to further understand GPU soft errors.
+Tiwari et al~\cite{7056044} analyzed the types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
+It also used neutron beam experiments at the Los Alamos Neutron Science Center and at Rutherford Appleton Laboratories to further understand GPU soft errors caused by neutron radiation triggered upsets (bit flips and timing errors).
 %
 Nie et al.~\cite{nie17characterizing,nie18machine} characterized the soft error behavior of Titan's GPUs in relation to temperature and power consumption to predict their increased occurrence by correlating data in temporal and spatial domains and using machine learning models.
+The work primarily focused on correctable single-bit errors.
 %
-Tiwari et al. and Nie et al. did not address the failure mode discussed in this paper.
-%
+Tiwari et al. and Nie et al. did not address the atypical failure mode discussed in this paper.
+
 Zimmer et al.~\cite{8665764} developed a new job scheduling strategy for Titan to counter the GPU failures discussed in this paper and to improve system utilization and productivity.
+The solution used reordering of compute nodes for resource allocation, scheduling larger jobs on more reliable and smaller jobs on less reliable nodes.
 
 Work by Kumar et al.~\cite{kumar18understanding} analyzed Titan's interconnect faults, errors and congestion events to improve interconnects resilience and congestion resolution methods.
+The results show that the magnitude of interconnect errors is very high with an uneven distribution across different types of links is uneven.
+They also show that congestion is highly frequent and bursty.
 %
 Gupta et al.~\cite{gupta17failures} performed a study covering five supercomputers at Oak Ridge National Laboratory, including Titan, that concentrated on developing an understanding of which and how many errors and failure occur in supercomputers over multiple generations and over their years of operation.
-%
+This work resulted in a lot of lessons learned, including that the mean-time-to-failure (MTTF) can change drastically and non-monotonically over time.
+
 Bautista-Gomez et al.~\cite{bautista-gomez16reducing} used failure data from Titan to adapt the checkpoint frequency to the current reliability of the system using dynamic checkpointing.
@@
 Tiwari et al.~\cite{6903564} also used failure data from Titan and exploited temporal locality in the data to adapt the checkpoint frequency to the current reliability of the system using lazy checkpointing.
+%
+Both approaches, Bautista-Gomez et al.
+and Tiwari et al., aim at dealing with the drastically and non-monotonically changing MTBF of Titan to match current system reliability with an efficient recovery strategy.
 
@@
 %
-De et al.~\cite{8809553} developed an in-depth understanding of the failure characteristics of the IBM Blue Gene/Q Mira system at Argonne National Laboratory.
+Di et al.~\cite{8809553} developed an in-depth understanding of the failure characteristics of the IBM Blue Gene/Q Mira system at Argonne National Laboratory.

### 2020-01-16 related.tex
Added more related work.

@@
 Di et al.~\cite{8809553} developed an in-depth understanding of the failure characteristics of the IBM Blue Gene/Q Mira system at Argonne National Laboratory.
+This work showed that 99.4\% of job failures were due to user behavior, such as errors in the code or misconfiguration.
+Martino et al.~\cite{6903615} characterized the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications (NCSA), at the University of Illinois at Urbana-Champaign.
+Blue Waters's XK7 partition had the same architecture as Titan.
+The results showed that 74.4\% of system-wide outages was caused by software.
+
+\fix{More references regarding GPU reliability}

### 2020-01-28 related.tex
Modified BlueWaters reference.

@@
 Martino et al.~\cite{6903615} characterized the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications (NCSA), at the University of Illinois at Urbana-Champaign.
-Blue Waters's XK7 partition had the same architecture as Titan.
-The results showed that 74.4\% of system-wide outages was caused by software.
+Blue Waters's XK7 partition had the same architecture as Titan, but did not experience the same GPU failure mode as Titan that is detailed in this paper.
+The results showed that 74.4\% of system-wide outages in Blue Waters was caused by software.
 

### 2020-04-06 discussion.tex
miscellaneous ...

@@
 \label{sec:discussion}
-\fix{Jim and everyone}
+\fix{Christian, Jim and everyone}
+
+\begin{itemize}
+\item Cray and NVIDIA take care of the hardware and some systems and driver
+  software so moving and troubleshooting hardware is their cost.
+\item Checkpoint strategies, allocation strategies, job resource
+  management, are all under OLCF control. What can OLCF do/change to
+  mitigate fallout in terms of delivering computing to customers when
+  facing machine disasters like the two episodes (2013 and 2016) on Titan?
+\end{itemize}

### 2020-04-06 survival.tex
miscellaneous ...

@@
 
-Kaplan-Meier survival analysis gives the probability of survival beyond a given time.
-It is a nonparametric technique that makes no specific failure model assumptions, such as Weibull, Exponential, etc.
-It can also split the data into subpopulations and compute separate survival curves.
+Kaplan-Meier survival analysis gives the probability of survival
+beyond a given time. It is a nonparametric technique that makes no
+specific failure model assumptions, such as Weibull, Exponential,
+etc. It can also split the data into subpopulations and compute
+separate survival curves.
 
-If $T$ is taken to be the random variable of a GPU fail time, then its probability density function is $f(t)$ and its cumulative distribution function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails by duration $t$.
-The survival function is then
+If $T$ is taken to be the random variable of a GPU fail time, then its
+probability density function is $f(t)$ and its cumulative distribution
+function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails
+by duration $t$. The survival function is then 
 \begin{displaymath}
@@
 
-The Cox proportional hazards (CPH) regression analys \cite{Cox1972,Harrell2015} takes the hazard function above and a set of covariates $x$ and models $k$s GPU hazard as
+The Cox proportional hazards (CPH) regression analys
+\cite{Cox1972,Harrell2015} takes the hazard function above and a set
+of covariates $x$ and models $k$s GPU hazard as 
 \begin{displaymath}

### 2020-04-13 dataclean.tex
miscellaneous ...
mostly km model revamped

@@
   location, its use in cleaning data, etc.}
-
-A typical GPU data record produced by the methods in the last section
-is of the following form
-{\small
-\begin{verbatim}
-0323812007945 | c17-4c1s3n1 | 01/10/2013 02:54:45 | 02/02/2013 11:32:29
- | c13-1c1s3n3 | 01/21/2014 21:10:42 | 08/01/2017 02:15:02
- | c0-1c1s3n3 | 10/11/2013 15:57:33 | 10/12/2013 22:09:31
- | c21-1c2s5n0 | 03/19/2013 15:48:11 | 05/29/2013 11:54:11
-0325216047736 | c18-4c1s5n1 | 03/07/2017 02:15:01 | 03/12/2019 03:19:11
-0323812008856 | c5-4c0s7n0 | 01/10/2013 02:54:45 | 01/25/2013 15:29:58
- | c0-6c1s7n2 | 10/21/2013 14:28:19 | 10/28/2013 17:52:44
- | c3-3c1s5n0 | 05/29/2013 11:54:11 | 05/29/2013 11:54:11
- | c23-6c1s7n2 | 01/21/2014 21:10:42 | 11/02/2018 14:42:34
- | | DBE | 11/02/2018 14:42:34
-\end{verbatim}
-}
-\noindent where records for three GPUs are shown. Each record starts
-with a serial number and locations are coded with {\tt c{\it
-    col-row}c{\it cage}s{\it slot}n{\it node}}. The first GPU record
-shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},
-{\tt c0-1c1s3n3}, with periods off the system, and finally in {\tt
-  c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
-not seen again. The second is intalled in location {\tt c18-4c1s5n1},
-where it stays until the last data collection date on March 3, 2019,
-at 3:19:11 AM. The third GPU is first installed in locations {\tt
-  c5-4c0s7n0}, {\tt c3-3c1s5n0},{\tt c0-6c1s7n2}, {\tt c23-6c1s7n2},
-where a ``DBE'' is observed on November 2, 2018, and it is not seen
-again.
 
@@
 
-A plot of the life of 100 random GPUs on Titan after cleaning the data:
-\begin{description}
-\item[starts:] black dots (gray for zero lifetime)
-\item[lifetimes:] black lines
-\item[DBEs:] red triangles
-\item[Off The BUS:] blue triangles
-\item[Last seen:] black ]
-\item[overlaps:] red square on left when they exist
-\end{description}
+To get some intuition for the GPU lifetimes on Titan, we give two views of
+a hundred GPUs (Figure~\ref{fig:gpuview}) and a hundred locations
+(Figure~\ref{fig:locview}). The GPU view is ``by serial number'' that
+shows when each GPU was installed and removed at various locations as
+well as its DBE and OTB events. The second view by location, shows
+when different GPUs were installed and removed at a location, and
+their OTB and DBE events.
+\begin{figure}[ht]
+  \includegraphics[width=6in]{figs/sample_sn.pdf}
+  \caption{GPU serial number view of life and failures. Legend: black
+    dots are installs, black lines are lifetimes at installed
+    location, blue squares are OTB events, red triangles are DBE
+    events, and black ] are ``last seen'' events.}
+  \label{fig:gpuview}
+\end{figure}
+\begin{figure}[ht]
+  \includegraphics[width=6in]{figs/sample_loc.pdf}
+  \caption{GPU location view of life and failures.  Legend: black
+    dots are installs, black lines are lifetimes at installed
+    location, blue squares are OTB events, red triangles are DBE
+    events, and black ] are ``last seen'' events.}
+  \label{fig:locview}
+\end{figure}
+The serial number view clearly shows the two batches and the more
+frequent OTB and DBE events in the old batch. It is also clear from
+this view that most new GPUs stay at the initial install location
+whereas the old GPUs were occasionally reinstalled at new locations.
+The location view shows that each location was operational almost all
+the time with small gaps when GPUs were changed out.
 
-\begin{figure}[ht]
-  \includegraphics[width=6in]{sn.pdf}
-  \caption{GPU serial number view of life and failures.}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics[width=6in]{location.pdf}
-  \caption{GPU location view of life and failures.}
-\end{figure}
+\fix{Were the GPUs swapped out as a blade or were individual GPUs
+  swapped on a blade?}

### 2020-04-13 dataprep.tex
mostly km model revamped

--- 2019-08-24 dataprep.tex
@@
 \fix{Don's section}
+Data on GPU life times is constructed from two sources: inventory runs
+and failure event records. Two types of failure events were collected:
+Double Bit Error (DBE) and Off the Bus (OTB). DBE was found to be the
+``signature'' error of eventual GPU failures.
+
+An inventory of GPU serial numbers and their locations is recorded
+each time the system boots. \fix{Don, you mentioned that this was more
+regular later - daily?} This typically occurs once every few days but
+can be as frequent as a few times per day. An inventory at boot time
+typically occurs over a span of \fix{??} minutes. A separate file is
+produced at each boot time.
+
+The DBE and OTB events are also recorded separately from boot
+time inventories. This is recorded as . . . files.
+
+Inventories are taken at boot time, which can occur anywhere from
+several times per day to more than a month
+apart. Figure~\ref{fig:inventory} shows inter-inventory times that
+appear in the data.
+\begin{figure}[bth]
+  \begin{center}
+    \includegraphics[width=\textwidth]{figs/attention_intervals001.pdf}
+  \end{center}
+  \caption{Inter-inventory times, obrained by taking intervals between
+    unique dates in the data. Note square root scale on $y$-axis.}
+  \label{fig:inventory}
+\end{figure}
+\fix{George: Add color for another variable? (new-old, date)??}
+Notably, there was a period of over a month that Titan operated
+without a single reboot.
+
+Initial processing updates an inventory by checking the Serial Number
+(SN) and Location of each GPU and creating or updating a separate
+record for each contiguously observed SN-Location combination.
+
+A typical GPU data record is of the form shown in Fig.~\ref{fig:dataraw},
+\begin{figure}
+{\small
+\begin{verbatim}
+0323812007945 | c17-4c1s3n1 | 01/10/2013 02:54:45 | 02/02/2013 11:32:29
+ | c13-1c1s3n3 | 01/21/2014 21:10:42 | 08/01/2017 02:15:02
+ | c0-1c1s3n3 | 10/11/2013 15:57:33 | 10/12/2013 22:09:31
+ | c21-1c2s5n0 | 03/19/2013 15:48:11 | 05/29/2013 11:54:11
+0325216047736 | c18-4c1s5n1 | 03/07/2017 02:15:01 | 03/12/2019 03:19:11
+0323812008856 | c5-4c0s7n0 | 01/10/2013 02:54:45 | 01/25/2013 15:29:58
+ | c0-6c1s7n2 | 10/21/2013 14:28:19 | 10/28/2013 17:52:44
+ | c3-3c1s5n0 | 05/29/2013 11:54:11 | 05/29/2013 11:54:11
+ | c23-6c1s7n2 | 01/21/2014 21:10:42 | 11/02/2018 14:42:34
+ | | DBE | 11/02/2018 14:42:34
+\end{verbatim}
+}
+\caption{A few records of raw data produced from inventories that is
+  processed further in our analysis.}
+\label{fig:dataraw}
+\end{figure}
+\noindent where records for three GPUs are shown. Each record starts
+with a serial number and locations are coded with {\tt c{\it
+    col-row}c{\it cage}s{\it slot}n{\it node}}. The first GPU record
+shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},
+{\tt c0-1c1s3n3}, with periods off the system, and finally in {\tt
+  c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
+not seen again. The second is intalled in location {\tt c18-4c1s5n1},
+where it stays until the last data collection date on March 3, 2019,
+at 3:19:11 AM. The third GPU is first installed in locations {\tt
+  c5-4c0s7n0}, {\tt c3-3c1s5n0},{\tt c0-6c1s7n2}, {\tt c23-6c1s7n2},
+where a ``DBE'' is observed on November 2, 2018, and it is not seen
+again.

### 2020-04-13 intro.tex
revised assignments
mostly km model revamped

@@
 \label{sec:intro}
-\fix{Dropped Jim's initial email here. Mostly Jim and Christian}
-The Cray XK7 Titan is rapidly nearing its decommissioning date for
-DOE/ASCR. The last production runs will be in July of this year, and
-then we will stop services, clean up a few things and turn it off.
+The Cray XK7 Titan supercomputer was the \#1 system in the world for a
+very long time, and has remained a critically important computer
+system through the end of its life in the Summer of 2019. It defied
+scale with 18,688 individual compute nodes and delivered tens of
+billions of computing hours to the U.S. Oepartment of Energy
+mission-critical programs for nearly 7 years.
 
-This system was \#1 in the world for a very long time, and has remained
-a critically important computer system. It defied scale with 18,688
-individual compute nodes, and delivered tens of billions of computing
-hours to DOE mission-critical programs for nearly 7 years.
+Titan was quite an interesting machine from a reliability perspective.
+Its operation was forced to execute three very significant rework cycles, two on
+the mechanical assembly affecting the PCIe connector from the GPU
+daughtercard to the motherboard, and a replacement of about 11,000 GPU
+assemblies because of a failing resistor on a printed circuit board.
 
-This was quite an interesting machine from a reliability perspective.
-We were forced to execute three very significant rework cycles, two on
-the mechanical assembly affecting the PCIe connector from the GPU
-daughtercard to the motherboard, and more recently, a replacement of
-about 11,000 GPU assemblies because of a failing resistor on a printed
-circuit board.
+Figure~\ref{fig:chronology} illustrates the chronology of the rework
+cycles. It shows the number of GPU changes at periodic inventories.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=\textwidth]{figs/chronology.pdf}
+  \caption{Daily volume of GPU swaps with labels on most frequent dates.}
+  \label{fig:chronology}
+\end{figure}
 
@@
 appropriate models extrapolate failure rates beyond the end of service
-date (the dotted lines…) ?
+date (the dotted lines) ?
 
@@
 up. I think this is a timely and very interesting topic.
+
+\fix{Need a description of Titan GPU events timeline. Jim or Don?}

### 2020-04-13 survival.tex
mostly km model revamped

@@
 \label{sec:survival}
-\fix{George's survival analysis, check \protect\pkg{finalfit}}
-
-Kaplan-Meier survival analysis gives the probability of survival
+ Kaplan-Meier survival analysis (KM) gives the probability of survival
 beyond a given time. It is a nonparametric technique that makes no
@@
 \end{displaymath}
-which is the probability of being operational just before duration $t$.
+which is the probability of being operational at duration
+$t$.
+
+We use reliability censoring concepts to use information from units
+that were proactively removed before failing as well as those still in
+operation till the end. We use the R package {\tt survival} for the KM
+analysis.
+\begin{figure}
+  \centering
+  \includegraphics[width=6in]{figs/km_cage-node_o001.pdf}
+  \caption{Survival probability from Kaplan-Meier analysis for {\tt
+      old} batch GPUs, stratified by {\tt cage} and {\tt node}.}
+  \label{fig:km-old-cage-node}
+\end{figure}
+\begin{figure}
+  \centering
+  \includegraphics[width=6in]{figs/km_cage-node_c001.pdf}
+  \caption{Here we consider the effect of cage on survival among the
+    old batch.}
+  \label{fig:km-all-cage-node}
+\end{figure}
 
@@
 \begin{figure}
-  \includegraphics[width=5in]{cox.pdf}
-  \caption{GPU hazard ratios from Cox regression model.}
+  \includegraphics[width=5in]{figs/cox_o001.pdf}
+  \caption{GPU hazard ratios from Cox regression model on {\tt old}
+    batch.}
+\end{figure}
+\begin{figure}
+  \includegraphics[width=5in]{figs/cox_n001.pdf}
+  \caption{GPU hazard ratios from Cox regression model on {\tt new}
+    batch.}
 \end{figure}
@@
 \begin{figure}
-  \includegraphics[width=5in]{glmnet.pdf}
+  \includegraphics[width=5in]{figs/glmnet001.pdf}
   \caption{GPU variable importance from glmnet penalized regression analysis.}

### 2020-04-14 survival.tex
mostly figures

@@
   \includegraphics[width=6in]{figs/km_cage-node_c001.pdf}
-  \caption{Here we consider the effect of cage on survival among the
-    old batch.}
+  \caption{Comparison of the old and new batches, focusing only on
+    three years of operation of the new batch so far.}
   \label{fig:km-all-cage-node}
@@
 
-Both loc and move are roughly the same, indicating that GPUs that were
-moved to another location had slighly (.75x) lower failure rate. This
-may just reflect a “repair effect” that extends the life of a GPU. It
-may have been removed and reinstalled for a reason that was not
-recorded.
-
-Oddly, the older batch has a very slighly lower hazard, although there
-is a lot of uncertainty here, indicating that something more
-complicated is at work and needs deeper investigation.
-
 Now the multivariate Cox regression analysis, where each factor is
@@
 
-It is strange that move and loc effects are opposite, whereas in the
-univariate analysis they were the same. They are also much
-stronger. The batch effect is opposite of what it was under univariate
-analysys and a bit stronger as well. This is a concern and it may be
-related to violation of the “proportionality assumption” of hazards
-not crossing. We see some crossing in the survival curves above
-already.
-
 Let’s separate the “old” and “new” batches of GPUs and run separate
 Cox models.
-
-\begin{figure}
-  \includegraphics[width=5in]{figs/glmnet001.pdf}
-  \caption{GPU variable importance from glmnet penalized regression analysis.}
-\end{figure}

### 2020-04-14 dataclean.tex
readded sections with modified pdfs

@@
 their OTB and DBE events.
-\begin{figure}[ht]
-  \includegraphics[width=6in]{figs/sample_sn.pdf}
+\begin{figure*}[tb]
+  \includegraphics[width=6.5in]{figs/sample_sn.pdf}
   \caption{GPU serial number view of life and failures. Legend: black
@@
   \label{fig:gpuview}
-\end{figure}
-\begin{figure}[ht]
-  \includegraphics[width=6in]{figs/sample_loc.pdf}
+\end{figure*}
+\begin{figure*}[tb]
+  \includegraphics[width=6.5in]{figs/sample_loc.pdf}
   \caption{GPU location view of life and failures.  Legend: black
@@
   \label{fig:locview}
-\end{figure}
+\end{figure*}
 The serial number view clearly shows the two batches and the more

### 2020-04-14 intro.tex
readded sections with modified pdfs
Cleaning up.

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Introduction
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
-\label{sec:intro}
+\label{section:introduction}
+
 The Cray XK7 Titan supercomputer was the \#1 system in the world for a
@@
 cycles. It shows the number of GPU changes at periodic inventories.
-\begin{figure}[htbp]
+\begin{figure*}[tbp]
   \centering
-  \includegraphics[width=\textwidth]{figs/chronology.pdf}
+  \includegraphics[width=\textwidth]{figs/chronology001.pdf}
   \caption{Daily volume of GPU swaps with labels on most frequent dates.}
   \label{fig:chronology}
-\end{figure}
+\end{figure*}
 

### 2020-04-15 abstract.tex
Cleaning up.
new title and abstract

@@
-This system was \#1 in the world for a very long time, and has remained
-a critically important computer system. It defied scale with 18,688
-individual compute nodes, and delivered tens of billions of computing
-hours to DOE mission-critical programs for nearly 7 years.
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Abstract
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The Cray XK7 Titan was the top supercomputer system in the world for a
+very long time and remained critically important throughout its nearly
+seven year life. It was also a very interesting machine from a
+reliability viewpoint as most of its power came from 18,688 GPUs whose
+operation was forced to execute three very significant rework cycles,
+two on the mechanical assembly and one on the GPUs. We write about the
+last rework cycle and a reliability analysis of the GPUs. Using
+statistical survival analysis techniques, we find that GPU reliability
+is dependent on heat dissipation to an extent that strongly correlates
+with detailed nuances of the system cooling architecture. In addition
+to describing some of the system history, the data collection, data
+cleaning, and our analysis of the data, we provide reliability
+recommendations for designing future state of the art supercomputing
+systems and their operation.

### 2020-04-15 survival.tex
readded sections with modified pdfs
Cleaning up.
km plots into one

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Survival Analysis
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Survival Analysis}
-\label{sec:survival}
+\label{section:survival}
+
  Kaplan-Meier survival analysis (KM) gives the probability of survival
@@
 analysis.
-\begin{figure}
+\begin{figure*}
   \centering
-  \includegraphics[width=6in]{figs/km_cage-node_o001.pdf}
-  \caption{Survival probability from Kaplan-Meier analysis for {\tt
-      old} batch GPUs, stratified by {\tt cage} and {\tt node}.}
-  \label{fig:km-old-cage-node}
-\end{figure}
-\begin{figure}
-  \centering
-  \includegraphics[width=6in]{figs/km_cage-node_c001.pdf}
-  \caption{Comparison of the old and new batches, focusing only on
-    three years of operation of the new batch so far.}
+  \includegraphics[width=7in]{figs/km_cage-node_a001.pdf}
+  \caption{Comparison of the old and new batches, including survival
+    differences based on {\tt cage} and {\tt node} GPU locations.}
   \label{fig:km-all-cage-node}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
-  \includegraphics[width=5in]{figs/cox_o001.pdf}
+\begin{figure*}
+  \includegraphics[width=6.5in]{figs/cox_o001.pdf}
   \caption{GPU hazard ratios from Cox regression model on {\tt old}
     batch.}
-\end{figure}
-\begin{figure}
-  \includegraphics[width=5in]{figs/cox_n001.pdf}
+\end{figure*}
+\begin{figure*}
+  \includegraphics[width=6.5in]{figs/cox_n001.pdf}
   \caption{GPU hazard ratios from Cox regression model on {\tt new}
     batch.}
-\end{figure}
+\end{figure*}
 

### 2020-04-15 dataprep.tex
new figures
readded sections with modified pdfs
Cleaning up.
smaller data figure
Layout figs

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Data Collection and Preprocessing
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Data Collection and Preprocessing}
-\label{sec:data}
+\label{section:dataprep}
+
 \fix{Don's section}
@@
 appear in the data.
-\begin{figure}[bth]
+\begin{figure}[bt]
   \begin{center}
-    \includegraphics[width=\textwidth]{figs/attention_intervals001.pdf}
+    \includegraphics[width=\columnwidth]{figs/attention_intervals001.pdf}
   \end{center}
-  \caption{Inter-inventory times, obrained by taking intervals between
-    unique dates in the data. Note square root scale on $y$-axis.}
+  \caption{Inter-inventory times, taken as time intervals between
+    unique dates in the data. Note square root scale on the vertical
+    axis.}
   \label{fig:inventory}
@@
 A typical GPU data record is of the form shown in Fig.~\ref{fig:dataraw},
-\begin{figure}
-{\small
+\begin{figure}[tb]
+{\tiny
 \begin{verbatim}
@@
 again.
+
+The location references are cabinet row (0-24), column (0-7), cage
+(0-3), slot (0-7), and node (0-3) with respect to the layout shown in
+Figure~\ref{fig:layout}.
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/TitanCabinet.png}
+  \end{minipage}
+  \begin{minipage}{0.49\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/TitanBlade.png}
+
+    \vspace{5em}
+    \includegraphics[width=\columnwidth]{figs/TitanLayout.png}
+  \end{minipage}
+  \caption{Titan cabinet, blade, and floor layout. Note cooling
+    airflow is bottom to top within cabinet.}
+  \label{fig:layout}
+\end{figure}
+
+\fix{The logical 3d torus layout is $25\times 16\times 24$. How does
+  the 16 dimension map onto this? Are the 4 nodes actually very far
+  apart in terms of the network layout?}
+
+\fix{Where does cool air arrive? Is it just dumped out at the top of
+  the heat exchanger and circulate down? I noticed airholes in some
+  floor tiles (more on the right than on the left). Is there cool air
+  under the floor?}

### 2020-04-15 dataprep.tex
more questions

@@
 
-\fix{The logical 3d torus layout is $25\times 16\times 24$. How does
-  the 16 dimension map onto this? Are the 4 nodes actually very far
-  apart in terms of the network layout?}
+\fix{In the decomissioning video, I noticed there are actually 12+2+12
+  cabinets, the middle two being different. Are the compute nodes the
+  pairs of outside 12s and the two middle are service nodes?}
+
+\fix{The logical 3d torus layout is $25\times 16\times 24$. How do
+  the 25 and 16 dimensions map onto this? Are the 4 nodes actually very far
+  apart in terms of the network layout via the two routers?}
 
@@
   under the floor?}
+
+\fix{How does the cooling water circulate?}

### 2020-04-15 intro.tex
Layout figs
fixed intro

@@
 
-Titan was quite an interesting machine from a reliability perspective.
-Its operation was forced to execute three very significant rework cycles, two on
-the mechanical assembly affecting the PCIe connector from the GPU
-daughtercard to the motherboard, and a replacement of about 11,000 GPU
-assemblies because of a failing resistor on a printed circuit board.
+From a realiability perspective, Titan was a very interesting
+machine. Its operation was forced to execute three very significant
+rework cycles, two on the mechanical assembly affecting the PCIe
+connector from the GPU daughtercard to the motherboard, and a
+replacement of about 11,000 GPU assemblies because of a failing
+resistor on a printed circuit board. We write primarily about the epoch
+that includes this last rework cycle. This includes Titan's most
+stable and failure free period and contains the most reliable data on
+GPU operation.
 
 Figure~\ref{fig:chronology} illustrates the chronology of the rework
-cycles. It shows the number of GPU changes at periodic inventories.
+cycles and stable periods as indicated by the number of GPU changes at
+periodic inventories.
 \begin{figure*}[tbp]
@@
 \end{figure*}
+After an initial fairly stable break-in period, the first rework cycle
+begins in early 2013 and ends by May. \fix{Details?}. The second
+rework cycle begins in late August of 2013 and ends by mid
+December. This is followed by a very long and stable period until
+about mid 2016, when GPU failures begin to rise. This results in the
+final rework cycle, replacing 11,000 GPUs throughout all of 2017.
 
-We have conducted a good bit of statistical analysis on the failure
-rates of these GPU assemblies, but I want to propose a different topic
-that I have not seen for other/similar systems.  I want to complete
-the statistical analysis of the failure rates as we near end of life,
-looking for the opposite side of the “bathtub curve”.  There is
-specific decay associated with the electronic components where, as
-these products age, they will eventually fail thresholds for guard
-band margin and other things. In a lot of cases, they will just fail
-as well.
+While the 2016 resistor failures were a surprise, there is a known
+phenomenon \cite{referenceneeded} that describes a two phase process
+that leads to failure. It also explains why there was a very stable
+period preceeding the failures. There is specific decay associated
+with the electronic components where, as these products age, they will
+eventually fail thresholds for guard band margin and other things. In
+a lot of cases, they will just fail as well. \fix{Give a reference and
+  expand?}
 
-This machine is quite large, was very important to many people, and I
-think this analysis could be very intriguing.
+Our analysis focuses on GPUs that were installed in the second rework
+cycle and later, essentialy units installed near the beginning of 2014
+and later.
 
-I am looking for help, very specifically with the statistics portion
-of this, i.e. what do the failure rates for specific components look
-like? What distributions do those failure rates follow? How could
-appropriate models extrapolate failure rates beyond the end of service
-date (the dotted lines) ?
-
-I have submitted an abstract to a Cray-focused user group and workshop
-on this topic, and it has been accepted.  I would like to execute this
-statistical analysis and generate a short presentation (30 minutes)
-and a short paper (4-8 pages is a likely target).  Presentation is
-targeted for May 7-9 (TBD). An initial/working version of the paper
-would be due earlier. If we miss that target (April 12), no big
-deal. We can finish on a similar schedule as the presentation.  I have
-all of the raw data for the component failures, and need the
-statistical help to complete the analysis.
-
-Because there is a huge swing in the failure rates of the GPUS from
-2016/17 when the resistor problem escalated rapidly out of control, it
-is statistically interesting, where we inject fresh material in to an
-existing machine. We’d likely want to categorically separate new GPU
-assemblies from old and could track the early life failures from those
-new 11,000 parts versus the continued failure of the remaining 6000.
-
-George, Christian-  any interest in this? Barney gave it a thumbs
-up. I think this is a timely and very interesting topic.
-
-\fix{Need a description of Titan GPU events timeline. Jim or Don?}
+We begin by describing related work in Sec.~\ref{sec:related}. Our
+data collection is desctibed in Sec.~\ref{sec:dataprep} and the data
+cleaning, checking, and lifetime summarization process in
+Sec.~\ref{sec:dataclean}. A timme between failures analysis is given
+in Sec.~\ref{sec:tbf}. Survival analysis based on Kaplan-Meier
+modeling and Cox regression modeling of relative hazard rates are in
+Sec.~\ref{sec:survival}. Finally, we discuss our main conclusions and
+recommendations in Sec.~\ref{sec:discussion}.

### 2020-04-15 dataprep.tex
pipes into

@@
 
-The location references are cabinet row (0-24), column (0-7), cage
+The location references are cabinet column (0-24), row (0-7), cage
 (0-3), slot (0-7), and node (0-3) with respect to the layout shown in
@@
 
-\fix{How does the cooling water circulate?}
+\fix{How does the coolant circulate? I notice three sets of pipes on
+  each side of each row: (2 x 3) x 8 = 48 sets.}

### 2020-04-15 dataclean.tex
Cleaning up.
fixed section refs

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Viewing and Cleaning the Data
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Viewing and Cleaning the Data}
-\label{sec:clean}
-\fix{George's section: The two views of GPU life: by SN and by
-  location, its use in cleaning data, etc.}
-
+\label{section:dataclean}
 As we need to recover durations from this data, correct processing

### 2020-04-15 intro.tex
fixed section refs

@@
 
-We begin by describing related work in Sec.~\ref{sec:related}. Our
-data collection is desctibed in Sec.~\ref{sec:dataprep} and the data
+We begin by describing related work in Sec.~\ref{section:related}. Our
+data collection is desctibed in Sec.~\ref{section:dataprep} and the data
 cleaning, checking, and lifetime summarization process in
-Sec.~\ref{sec:dataclean}. A timme between failures analysis is given
-in Sec.~\ref{sec:tbf}. Survival analysis based on Kaplan-Meier
+Sec.~\ref{section:dataclean}. A timme between failures analysis is given
+in Sec.~\ref{section:tbf}. Survival analysis based on Kaplan-Meier
 modeling and Cox regression modeling of relative hazard rates are in
-Sec.~\ref{sec:survival}. Finally, we discuss our main conclusions and
-recommendations in Sec.~\ref{sec:discussion}.
+Sec.~\ref{section:survival}. Finally, we discuss our main conclusions and
+recommendations in Sec.~\ref{section:discussion}.

### 2020-04-15 intro.tex
more fix section refs

@@
 We begin by describing related work in Sec.~\ref{section:related}. Our
-data collection is desctibed in Sec.~\ref{section:dataprep} and the data
-cleaning, checking, and lifetime summarization process in
-Sec.~\ref{section:dataclean}. A timme between failures analysis is given
-in Sec.~\ref{section:tbf}. Survival analysis based on Kaplan-Meier
-modeling and Cox regression modeling of relative hazard rates are in
-Sec.~\ref{section:survival}. Finally, we discuss our main conclusions and
-recommendations in Sec.~\ref{section:discussion}.
+data collection is desctibed in Sec.~\ref{section:dataprep} and the
+data cleaning, checking, and lifetime summarization process in
+Sec.~\ref{section:dataclean}. A time between failures (TBF) analysis
+is given in Sec.~\ref{section:tbf}. Survival analysis based on
+Kaplan-Meier modeling (KM) and Cox regression modeling (CR) of
+relative hazard rates in various locations are in
+Sec.~\ref{section:survival}. Finally, we discuss our main conclusions
+and recommendations in Sec.~\ref{section:discussion}.

### 2020-04-15 dataprep.tex
fixed section refs
tweaks

@@
 \label{section:dataprep}
-
-\fix{Don's section}
 Data on GPU life times is constructed from two sources: inventory runs
@@
 Double Bit Error (DBE) and Off the Bus (OTB). DBE was found to be the
-``signature'' error of eventual GPU failures.
+``signature'' error of eventual GPU failures and a trigger for their
+replacement.
 
@@
 The DBE and OTB events are also recorded separately from boot
-time inventories. This is recorded as . . . files.
+time inventories. This was recorded as \fix{ . . . } files.
 
@@
 \fix{George: Add color for another variable? (new-old, date)??}
-Notably, there was a period of over a month that Titan operated
-without a single reboot.
+Notably, there was a period of over a month and many periods of over a
+week that Titan operated without a single reboot.
 
@@
 at 3:19:11 AM. The third GPU is first installed in locations {\tt
-  c5-4c0s7n0}, {\tt c3-3c1s5n0},{\tt c0-6c1s7n2}, {\tt c23-6c1s7n2},
-where a ``DBE'' is observed on November 2, 2018, and it is not seen
-again.
+  c5-4c0s7n0}, {\tt c3-3c1s5n0}, {\tt c0-6c1s7n2}, and {\tt
+  c23-6c1s7n2}, where a ``DBE'' is observed on November 2, 2018, and
+it is not seen again.
 

### 2020-04-15 dataclean.tex
updated dataclean

@@
 \label{section:dataclean}
-As we need to recover durations from this data, correct processing
-involves time adjustments for switching between daylight saving time
-and standard time. We perform this by setting a reference time zone
-(Eastern time) and converting all date-times from strings into
-date-time variables \fix{(is there a name for this?)} with the R
+As we need to recover durations in operation from this data, correct
+processing involves time adjustments for switching between daylight
+saving time and standard time and leap years. We perform this by
+setting a reference time zone (Eastern time) and converting all
+date-times from strings into POSIX date-time variables with the R
 \pkg{lubridate} package \cite{lubridate}, which enables appropriate
-date arithmetic.
+date arithmetic and sensible date constructs for graphs.
 
-\fix{Other processing:} enable missing values, and create some new
-variables. The event is “life” when both insert and remove are
-present, “life0” when insert = remove, otherwise it is whatever string
-is in insert, which is “DBE” or “Off The BUS”. Also separate location
-into its components and fill in (repeat) serial numbers and locations
-for all records.
+As most analysis software uses table-like data, we fill the needed
+repeats of values missing in the raw data (see
+Fig.~\ref{fig:dataraw}. First, we reduce the data to the GPU that were
+removed later than 2013 to focus our analysis epoch on units installed
+in the second rework cycle. After this reduction, there were stil six
+older units remaining, which we also removed to have a clean third
+epoch set of units for the analysis. We do some further processing to
+handle time overlaps in a tiny fraction (under 0.0007 in GPU life and
+0.0002 in location life) of recorded life in the raw records by simply
+dropping the overlapping life times.
 
-Reduce to only “life” records and mark them with ending DBE, OTB,
-“out”, or none events. The result is just insert and remove with event
-marks to indicate DBE or OTB or “out” (last seen). bad indicates that
-a DBE or OTB occurred yet the GPU was not taken out. For a first cut,
-keep the bad ones in. Also keep duration to later aggregate into full
-life times.
+Next, we aggregate into one record per serial number with a total life
+time the first insert time, the last remove time, and a number of
+other quantities such as location where the longest time is spent, the
+proportion of time at the longest location, and the number of DBE or
+OTB events.
 
-Next, aggregate into one record per serial number with total life time
-and first insert time. For proper censoring treatment: Indicate if
-event occurred or still in service: out = fail, dbe = fail, otb =
-fail, NA = right censored (still in service). Add some other
-variables, such as GPU in single location or moved one or more times,
-mark new\_batch group, etc. to use in modeling later.
-
-To get some intuition for the GPU lifetimes on Titan, we give two views of
-a hundred GPUs (Figure~\ref{fig:gpuview}) and a hundred locations
-(Figure~\ref{fig:locview}). The GPU view is ``by serial number'' that
-shows when each GPU was installed and removed at various locations as
-well as its DBE and OTB events. The second view by location, shows
-when different GPUs were installed and removed at a location, and
-their OTB and DBE events.
+To get some intuition for the GPU lifetimes on Titan, we give two
+views of 80 randomly selected GPUs (Figure~\ref{fig:gpuview}) and 80
+randomly selected locations (Figure~\ref{fig:locview}). The GPU view
+visually documents the life of the unit: when it was installed and
+removed at various locations, its DBE and OTB events, and the last
+time it was seen. The location view documents the life of a location:
+when different GPUs were installed and removed, their OTB and DBE
+events, and whether a removal was the last time the unit was seen on
+the system. These views were critical to understanding the data and to
+verifying various processing decisions.
 \begin{figure*}[tb]
@@
 \end{figure*}
-The serial number view clearly shows the two batches and the more
-frequent OTB and DBE events in the old batch. It is also clear from
-this view that most new GPUs stay at the initial install location
-whereas the old GPUs were occasionally reinstalled at new locations.
+
+The third rework cycle was used to label GPUs as old batch and new
+batch. The two batches are clearly identifiable in the GPU view. Also,
+more frequent OTB and DBE events are apparent in the old batch. It is
+also clear from this view that most new GPUs stay at the initial
+install location whereas the old GPUs were occasionally reinstalled at
+new locations. Nevertheless, a separate analysis determined that the
+vast majority of time of the vast majority of units is spent at one
+location by both the new and the old units.
+
 The location view shows that each location was operational almost all
@@
 \fix{Were the GPUs swapped out as a blade or were individual GPUs
-  swapped on a blade?}
+  swapped on a blade? Seems it was the full blade - this may further
+  refine censoring. A DBE on one GPU triggers the replacement of the
+  blade with the other three good gpus?}

### 2020-04-15 dataclean.tex
figure placement

@@
 \label{section:dataclean}
+\begin{figure*}[bt]
+  \includegraphics[width=6.5in]{figs/sample_sn.pdf}
+  \caption{GPU serial number view of life and failures. Legend: black
+    dots are installs, black lines are lifetimes at installed
+    location, blue squares are OTB events, red triangles are DBE
+    events, and black ] are ``last seen'' events.}
+  \label{fig:gpuview}
+\end{figure*}
 As we need to recover durations in operation from this data, correct
@@
 views of 80 randomly selected GPUs (Figure~\ref{fig:gpuview}) and 80
-randomly selected locations (Figure~\ref{fig:locview}). The GPU view
-visually documents the life of the unit: when it was installed and
-removed at various locations, its DBE and OTB events, and the last
-time it was seen. The location view documents the life of a location:
-when different GPUs were installed and removed, their OTB and DBE
-events, and whether a removal was the last time the unit was seen on
-the system. These views were critical to understanding the data and to
-verifying various processing decisions.
-\begin{figure*}[tb]
-  \includegraphics[width=6.5in]{figs/sample_sn.pdf}
-  \caption{GPU serial number view of life and failures. Legend: black
-    dots are installs, black lines are lifetimes at installed
-    location, blue squares are OTB events, red triangles are DBE
-    events, and black ] are ``last seen'' events.}
-  \label{fig:gpuview}
-\end{figure*}
+randomly selected locations (Figure~\ref{fig:locview}).
 \begin{figure*}[tb]
@@
 \end{figure*}
+The GPU view
+visually documents the life of the unit: when it was installed and
+removed at various locations, its DBE and OTB events, and the last
+time it was seen. The location view documents the life of a location:
+when different GPUs were installed and removed, their OTB and DBE
+events, and whether a removal was the last time the unit was seen on
+the system. These views were critical to understanding the data and to
+verifying various data processing decisions.
 
 The third rework cycle was used to label GPUs as old batch and new
-batch. The two batches are clearly identifiable in the GPU view. Also,
-more frequent OTB and DBE events are apparent in the old batch. It is
-also clear from this view that most new GPUs stay at the initial
-install location whereas the old GPUs were occasionally reinstalled at
-new locations. Nevertheless, a separate analysis determined that the
-vast majority of time of the vast majority of units is spent at one
-location by both the new and the old units.
+batch. The two batches are clearly identifiable in the GPU view. More
+frequent OTB and DBE events are apparent in the old batch. It is also
+clear from this view that practically all new GPUs stayed at their
+initial install location whereas the old GPUs were occasionally
+reinstalled at new locations. Nevertheless, a separate analysis
+determined that the vast majority of time of the vast majority of
+units is spent at one location by both the new and the old units.
 

### 2020-04-16 survival.tex
reduced cox fig size
described failure and censoring

@@
 \label{section:survival}
+The construction of a GPU life time is more complex than it initially
+appears because the units are observed only at reboot time and because
+most units were proactively replaced to prevent failure.
+
+We consider both OTB and DBE as failures for units that are removed
+from the system and we record the inspection time as the failure
+time. Most of the time, this is the last time such a unit is seen on
+the system. In some cases, the last time seen is at a different time
+but we still record its removal time as the failure time.
+
+A key concept in survival analysis and in reliability is censoring,
+which is about using information from study subjects whose exact
+failure times are not available. This applies to our study because of
+proactive GPU replacement before failure but also because of
+observation only at inspection times (reboots). We use censoring
+concepts on the proactive replacements. This allows us to use all of
+the GPU life time data, including units that did not fail. But we
+ignore the inspection time censoring, treating removal times as exact
+failure times to reduce the complexity of this analysis. Because of
+the weight of so much data, this would not make any difference in our
+conclusions.
 
@@
 
-\begin{figure*}
-  \includegraphics[width=6.5in]{figs/cox_o001.pdf}
+\begin{figure}
+  \includegraphics[width=\columnwidth]{figs/cox_o001.pdf}
   \caption{GPU hazard ratios from Cox regression model on {\tt old}
     batch.}
-\end{figure*}
-\begin{figure*}
-  \includegraphics[width=6.5in]{figs/cox_n001.pdf}
+\end{figure}
+\begin{figure}
+  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
   \caption{GPU hazard ratios from Cox regression model on {\tt new}
     batch.}
-\end{figure*}
+\end{figure}
 

### 2020-04-16 survival.tex
removed censors from pdf plot

--- 2020-04-15 survival.tex
@@
 \label{section:survival}
+The construction of a GPU life time is more complex than it initially
+appears because the units are observed only at reboot time, because
+most units were proactively replaced to prevent failure, and because
+some units continue in operation after OTB and DBE events (a second
+reboot may be successful).
+
+A unit that experiences at least one OTB or DBE event and is removed
+from the system is considered failed and its operation time until the
+last seen time is taken as its life time. Although most failed units
+experience one of these events at the last seen time, this is
+imperfect because some units experience OTB or DBE events at a time
+different from its last seen time. We found this definition of life
+time the most pragmatic for the study with few imperfect
+observations. One exception is if the last seen time is the final
+Titan shutdown time, in which case such a unit is not considered
+failed.
+
+A key concept in survival analysis and in reliability analysis is
+censoring, which is about using information from study subjects whose
+exact failure times are not available or that have not failed. This
+applies to our study because of proactive GPU replacement before
+failure, because most units were still in operation when the system
+was shut down, and also because of observation only at inspection
+times (reboots). We use censoring concepts on the proactive
+replacements and units still in operation at the end. This allows us
+to use all of the GPU life time data, including units that did not
+fail. But we ignore the inspection time censoring, treating removal
+times as exact failure times to reduce the complexity of this
+analysis. We expect that because of the volume of data, this would
+hardly make a difference in our conclusions.
 

### 2020-04-16 dataclean.tex
tweak

@@
 repeats of values missing in the raw data (see
-Fig.~\ref{fig:dataraw}. First, we reduce the data to the GPU that were
-removed later than 2013 to focus our analysis epoch on units installed
-in the second rework cycle. After this reduction, there were stil six
-older units remaining, which we also removed to have a clean third
-epoch set of units for the analysis. We do some further processing to
-handle time overlaps in a tiny fraction (under 0.0007 in GPU life and
-0.0002 in location life) of recorded life in the raw records by simply
-dropping the overlapping life times.
+Fig.~\ref{fig:dataraw}). First, we reduce the data to the GPU that
+were removed later than 2013 to focus our analysis epoch on units
+installed in the second rework cycle or later. After this reduction,
+there were stil six older units remaining, which we also removed to
+have a clean third epoch set of units for the analysis. We do some
+further processing to handle time overlaps in a tiny fraction (under
+0.0007 in GPU life and 0.0002 in location life) of recorded life in
+the raw records by simply dropping the overlapping life times.
 

### 2020-04-17 intro.tex
public data?
updated dataclean
first changes reflecting meeting - more coming

@@
 scale with 18,688 individual compute nodes and delivered tens of
-billions of computing hours to the U.S. Oepartment of Energy
+billions of computing hours to the U.S. Department of Energy
 mission-critical programs for nearly 7 years.
@@
   \includegraphics[width=\textwidth]{figs/chronology001.pdf}
-  \caption{Daily volume of GPU swaps with labels on most frequent dates.}
+  \caption{Volume of GPU swaps detected at inventories. The break-in
+    period before 2014 involved working out mechanical assembly issues
+    at PCIe connectors
+    frequent dates.}
   \label{fig:chronology}
 \end{figure*}
-After an initial fairly stable break-in period, the first rework cycle
-begins in early 2013 and ends by May. \fix{Details?}. The second
-rework cycle begins in late August of 2013 and ends by mid
-December. This is followed by a very long and stable period until
-about mid 2016, when GPU failures begin to rise. This results in the
-final rework cycle, replacing 11,000 GPUs throughout all of 2017.
+The first two rework cycles before 2014 can be characterized as a
+break-in period on a new system that is first of its kind and pushes
+many technological boundaries, in this case mainly regarding blade
+mechanical assembly. The resulting field engineering generated high
+GPU swap counts. Late 2013 begins a very long period of stable
+operation with very few issues until about mid 2016, when GPU failures
+begin to rise. This results in the final rework cycle, replacing
+11,000 GPUs throughout much of 2017.
 
-While the 2016 resistor failures were a surprise, there is a known
+The failures were traced to a failing resistor on the circuitboard
+(not the GPU iself) While the 2016 resistor failures were a surprise, there is a known
 phenomenon \cite{referenceneeded} that describes a two phase process
@@
 cycle and later, essentialy units installed near the beginning of 2014
-and later.
+and later. \fix{Will we (are we required) make the data publicly
+  available?} The data used in this study, representing over 100,000
+years of operation may be the largest publicly available GPU
+reliability data set.
 

### 2020-04-18 dataclean.tex
more changes from last meeting

@@
   \includegraphics[width=6.5in]{figs/sample_sn.pdf}
-  \caption{GPU serial number view of life and failures. Legend: black
+  \caption{Serial number view of GPU life and failures. Legend: black
     dots are installs, black lines are lifetimes at installed
@@
   \includegraphics[width=6.5in]{figs/sample_loc.pdf}
-  \caption{GPU location view of life and failures.  Legend: black
-    dots are installs, black lines are lifetimes at installed
-    location, blue squares are OTB events, red triangles are DBE
-    events, and black ] are ``last seen'' events.}
+  \caption{Location view of GPU life and failures.  Legend: black dots
+    are installs, black lines are lifetimes at installed location,
+    blue squares are OTB events, red triangles are DBE events, and
+    black ] are ``last seen'' events.}
   \label{fig:locview}
@@
 The GPU view
-visually documents the life of the unit: when it was installed and
+visually documents the life of each GPU unit: when it was installed and
 removed at various locations, its DBE and OTB events, and the last
@@
 The location view shows that each location was operational almost all
-the time with small gaps when GPUs were changed out.
-
-\fix{Were the GPUs swapped out as a blade or were individual GPUs
-  swapped on a blade? Seems it was the full blade - this may further
-  refine censoring. A DBE on one GPU triggers the replacement of the
-  blade with the other three good gpus?}
+the time with small gaps when GPUs were changed out. It is also
+notable that OTB and DBE events are associated with a single GPU
+although four GPUs are together on a blade. The proactive replacements
+in the rework cycle were done by full blades. Events on single GPUs
+were usually first swapped for a new blade, the failed GPU was
+replaced on the blade, and the fixed blade then reused elsewhere. The
+survival analysis of Sec.~\ref{section:survival} handles these nuances
+by appropriate censoring.

### 2020-04-18 dataprep.tex
first changes reflecting meeting - more coming
more changes from last meeting

@@
   \end{minipage}
-  \caption{Titan cabinet, blade, and floor layout. Note cooling
-    airflow is bottom to top within cabinet.}
+  \caption{Titan cabinet, blade, and floor layout. Cooling airflow is
+    bottom to top within a cabinet, taking ambient air in at the foot
+    and relasing it back out at the top. Cooling fluid is passed
+    through two heat exchangers, one below Cage 0 and one above Cage
+    2.}
   \label{fig:layout}
@@
 
-\fix{In the decomissioning video, I noticed there are actually 12+2+12
-  cabinets, the middle two being different. Are the compute nodes the
-  pairs of outside 12s and the two middle are service nodes?}
+% \fix{Where does cool air arrive? Is it just dumped out at the top of
+%   the heat exchanger and circulate down? I noticed airholes in some
+%   floor tiles (more on the right than on the left). Is there cool air
+%   under the floor?}
 
-\fix{The logical 3d torus layout is $25\times 16\times 24$. How do
-  the 25 and 16 dimensions map onto this? Are the 4 nodes actually very far
-  apart in terms of the network layout via the two routers?}
-
-\fix{Where does cool air arrive? Is it just dumped out at the top of
-  the heat exchanger and circulate down? I noticed airholes in some
-  floor tiles (more on the right than on the left). Is there cool air
-  under the floor?}
-
-\fix{How does the coolant circulate? I notice three sets of pipes on
-  each side of each row: (2 x 3) x 8 = 48 sets.}
+% \fix{How does the coolant circulate? I notice three sets of pipes on
+%   each side of each row: (2 x 3) x 8 = 48 sets.}

### 2020-04-20 tbf.tex
Cleaning up.
adding writeup and figures for time between failure analysis

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Time Between Failures Analysis
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Time Between Failures Analysis}
-\label{sec:tbf}
-\fix{Rizwan's section}
+\label{section:tbf}
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/MTBF_GPUwise.pdf}
+  \end{center}
+  \caption{Distribution of MTBFs due to DBEs and OTBs across all GPUs.}
+  \label{fig:Device_MTBFs}
+\end{figure}
+
+Using the processed data, we analyze the inter-arrival times between the DBE and OTB events. 
+This analysis is done at the device level and the system level. 
+It provides important insights into the reliability of large-scale machines, where the 
+failure rate of an individual device is significantly different from the overall reliability
+of the machine.
+
+A histogram of MTBFs measured across GPUs which had at least one failure event is shown 
+in Figure~\ref{fig:Device_MTBFs}. This is a practical assessment of device reliabilities 
+as opposed to those provided in the device datasheet. The failures are tracked using the
+SN of the GPUs even though a GPU might have been placed at different locations in the 
+machine during its lifetime. The time to the first failure on a device is measured by taking
+the insert time as the reference point, whereas, a simple difference is taken for subsequent
+failures. It can be noticed that MTBFs due to DBE and OTB failures of most GPUs are clustered around 
+25,000 hours or 2.8 years. Apart from the center cluster, a significant portion of GPUs has a very 
+low MTBF (less than 1.5 hours) due to both DBE and OTB failures. This shows that a device 
+can see a failure immediately after it is put into service or multiple events are caused due 
+to a single root cause. On the other end of the spectrum, a noticeable portion of GPUs have very high
+MTBFs due to DBE failures. This indicates that most GPUs see a single DBE during their lifetime.
+Apart from this behavior, the distribution of MTBF due to DBE events looks very similar to that due 
+to OTB events. Overall, it can be noted from Figure~\ref{fig:Device_MTBFs} that the number of recorded 
+DBE events is significantly higher than the OTB events.
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/MTBF_CageWise.pdf}
+  \end{center}
+  \caption{MTBFs due to either DBEs or OTBs across the GPUs located in different cages.}
+  \label{fig:CageWise_MTBFs}
+\end{figure}
+
+\fix{Figure~\ref{fig:CageWise_MTBFs} shows temperature effect clearly. Should we move this 
+result to later in the paper?}
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/MTBF_quaterly_sys.pdf}
+  \end{center}
+  \caption{Variation of system-wide MTBF over the lifetime of the machine (time is segmented into quarters, 
+i.e., January through April is Q1 and so on).}
+  \label{fig:MTBF_sys}
+\end{figure}
+
+Although individual device MTBFs provide important insights into the varying failure rates
+observed across the machine. It is worthwhile to perform a system-wide MTBF analysis since 
+most high-performance computing applications use a significant number of GPUs in parallel. 
+In the following analysis, failures occurring across all GPUs are consolidated and time between failures
+is calculated. The old and new GPUs are all considered. At any given time, only a fixed number of GPUs
+are in operation so any variation in time-between-failures is due to individual device reliabilities.
+
+That said, a single MTBF number does not give an accurate picture of this machine.
+With many GPU relocations and replacements across the machine over its lifetime, it is 
+best to consider the variation of MTBF across fixed periods. Herein, we calculate 
+the mean of time-between-failures across four months. Figure~\ref{fig:MTBF_sys}
+clearly shows the considerable change in system-wide MTBF from one period to another. 
+The observation of the MTBF trend shows the different phases which the machine went through. 
+For example, when considering failure due to both DBE and OTB events more than 4X drop in MTBF is 
+noted from 2015-Q3 to 2015-Q4 (note, MTBF of earlier quarters was significantly higher and is not 
+included in the figure to better visualize various change points). This period marked the start of 
+a rapid decline in system-wide MTBF until 2016-Q4. The rapid rise of the number of failures before this 
+quarter can be seen in Figure~\ref{fig:NumFails_sys}. When GPU replacements start to take place in late 
+2016, it triggers an increase in MTBF. However, this change only lasts until 2018-Q1, when we see another 
+downward trend of system MTBF. Incidentally, 2018-Q1 also marks the completion of all GPU replacements. 
+So the upward trend noted in the period from 2016-Q4 to 2018-Q1 might be attributed to a portion of the machine 
+being unavailable while the reworks were taking place. A single replacement cycle lasted multiple weeks. 
+\fix{need to verify how long the replacement took}
+There is no definite way to incorporate this unavailability of the machine into MTBF analysis, so it has to
+be said that overall there has been an increase in MTBF of the machine due to the replacements. 
+For example, before 2017-Q1 MTBF as low as 2.7 hours is noted, whereas the lowest MTBF is 5.9 hours after this 
+period. With such huge variations in system MTBF, it is difficult to reliably run applications even while using 
+failure recovery approaches such as checkpoint restart, as discussed later on in Section~\ref{section:discussion}.
+
+The mean-time-between DBE and OTB failures are separately tracked as well in this analysis.
+One observation is the drastic increase in mean-time-between OTB failures after the GPU replacements.
+This is also evident from the reduction of the absolute number of OTB failures in Figure~\ref{fig:NumFails_sys}.
+However, the machine's MTBF is determined by the weakest link and DBE events tend to dictate it.
+Even though the replacements helped to increase the MTBF due to DBE events by a factor, the overall system 
+reliability is dictated by the components with the most age in the system. There is an upward trend towards 
+the end in the number of DBE failures in Figure~\ref{fig:NumFails_sys} which almost all of them are due to 
+older GPUs in the system. The old GPUs that occupy a minority portion of the machine after GPU replacements
+ and have a disproportionately high number of failures play a major role in determining the state of the 
+machine as noted in the next section.
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/NumFailures_Quarterly_newOld.pdf}
+  \end{center}
+  \caption{The number of DBE and OTB failures observed over the lifetime of the machine.
+A distinction is made for GPU failures happening across old GPUs to signify the number 
+of failures across the newer GPUs.}
+  \label{fig:NumFails_sys}
+\end{figure}

### 2020-04-20 related.tex
Cleaning up.
minor comments
Added more related work.

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Related Work
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Related Work}
-\label{sec:related}
-\fix{Christian's section}
+\label{section:related}
 
 Prior work in investigating Titan's reliability includes a significant amount of work that covers different aspects.
-%
+
 Tiwari et al~\cite{7056044} analyzed the types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
@@
 %
-Nie et al.~\cite{nie17characterizing,nie18machine} characterized the soft error behavior of Titan's GPUs in relation to temperature and power consumption to predict their increased occurrence by correlating data in temporal and spatial domains and using machine learning models.
+Tiwari et al~\cite{10.1145/2807591.2807666} further analyzed the types of GPU errors on Titan, including software/firmware related errors and failures, ECC double-bit errors, and GPU ``off the bus'' and ECC page retirement events.
+This work observed a ECC double-bit mean-time between errors (MTBE) of \~160 hours, or \~ one per week, with 86\% in GPU memory and the rest in the GPU register file.
+The experienced GPU ``off the bus'' events were caused by a system integration issue that was fixed and not an inherent GPU or GPU memory architecture flaw.
+%
+Nie et al.~\cite{7446091,nie17characterizing,nie18machine} characterized the soft error behavior of Titan's GPUs in relation to temperature and power consumption to predict their increased occurrence by correlating data in temporal and spatial domains and using machine learning models.
 The work primarily focused on correctable single-bit errors.
 %
-Tiwari et al. and Nie et al. did not address the atypical failure mode discussed in this paper.
+Tiwari et al. and Nie et al. did not address the atypical failure mode discussed in this paper and only used Titan error and failure data from June 2013 to February 2015.
 
@@
 
+Ezell~\cite{osti_1086655} created a general understanding of Titan's interconnect failures using an application to stress test Titan's Gemini interconnect.
+%
 Work by Kumar et al.~\cite{kumar18understanding} analyzed Titan's interconnect faults, errors and congestion events to improve interconnects resilience and congestion resolution methods.
@@
 %
-Gupta et al.~\cite{gupta17failures} performed a study covering five supercomputers at Oak Ridge National Laboratory, including Titan, that concentrated on developing an understanding of which and how many errors and failure occur in supercomputers over multiple generations and over their years of operation.
-This work resulted in a lot of lessons learned, including that the mean-time-to-failure (MTTF) can change drastically and non-monotonically over time.
+Gupta et al.~\cite{7266836} investigated the spatial and temporal properties of failures on Titan and their impact on resilience mechanisms and the implications for efficient system operation.
+%
+Gupta et al.~\cite{gupta17failures} also performed a study covering five supercomputers at Oak Ridge National Laboratory, including Titan, that concentrated on developing an understanding of which and how many errors and failure occur in supercomputers over multiple generations and over their years of operation.
+This work resulted in a lot of lessons learned, including that the mean-time between failures (MTBF) can change drastically and non-monotonically over time.
 
@@
 The results showed that 74.4\% of system-wide outages in Blue Waters was caused by software.
-
-\fix{More references regarding GPU reliability}

### 2020-04-20 related.tex
Added more related work.

@@
 
-Tiwari et al~\cite{7056044} analyzed the types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
+Tiwari et al~\cite{Tiwari15Experience,7056044} analyzed the types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
 It also used neutron beam experiments at the Los Alamos Neutron Science Center and at Rutherford Appleton Laboratories to further understand GPU soft errors caused by neutron radiation triggered upsets (bit flips and timing errors).
@@
 This work resulted in a lot of lessons learned, including that the mean-time between failures (MTBF) can change drastically and non-monotonically over time.
+%
+Meneses et al.~\cite{Meneses15Analyzing} analyzed the interplay and workload on Titan using failure and job submission logs. The results indicate that failures depend heavily on workload.
 

### 2020-04-20 intro.tex
Added references to the Introduction and fixed a few things in the Introduction.

@@
 
-The Cray XK7 Titan supercomputer was the \#1 system in the world for a
-very long time, and has remained a critically important computer
+The Cray XK7 Titan supercomputer~\cite{olcf:titan} was the \#1 system in the world for a
+very long time~\cite{top500list}, and has remained a critically important computer
 system through the end of its life in the Summer of 2019. It defied
@@
 The failures were traced to a failing resistor on the circuitboard
-(not the GPU iself) While the 2016 resistor failures were a surprise, there is a known
-phenomenon \cite{referenceneeded} that describes a two phase process
+(not the GPU iself). While the 2016 resistor failures were a surprise, there is a known
+phenomenon~\cite{referenceneeded} that describes a two phase process
 that leads to failure. It also explains why there was a very stable
-period preceeding the failures. There is specific decay associated
+period preceding the failures. There is specific decay associated
 with the electronic components where, as these products age, they will

### 2020-04-20 related.tex
Fixed related work section.

@@
 Martino et al.~\cite{6903615} characterized the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications (NCSA), at the University of Illinois at Urbana-Champaign.
-Blue Waters's XK7 partition had the same architecture as Titan, but did not experience the same GPU failure mode as Titan that is detailed in this paper.
-The results showed that 74.4\% of system-wide outages in Blue Waters was caused by software.
+The results showed that 74.4\% of system-wide outages in Blue Waters were caused by software.
+Blue Waters's XK7 partition had the same architecture as Titan, but did not experience the same GPU failure mode detailed in this paper.

### 2020-04-20 survival.tex

@@
 \label{section:survival}
+\renewcommand{\pkg}[1]{\textsf{#1}}
+\begin{figure*}
+  \centering
+  \includegraphics[width=7in]{figs/km_cage-node_a001.pdf}
+  \caption{Comparison of the old and new batches, including survival
+    differences based on {\tt cage} and {\tt node} GPU locations.}
+  \label{fig:km-all-cage-node}
+\end{figure*}
 The construction of a GPU life time is more complex than it initially
@@
 analysis. We expect that because of the volume of data, this would
-hardly make a difference in our conclusions.
+make much difference in our conclusions. However, we are making our
+data publicly available and expect that others, especially in the
+survival analysis community, will dive deeper.
 
@@
 specific failure model assumptions, such as Weibull, Exponential,
-etc. It can also split the data into subpopulations and compute
-separate survival curves.
+etc. The technique is able to use censored observations and can also
+split the data into subpopulations and compute separate survival
+curves.
 
@@
 \end{displaymath}
-which is the probability of being operational at duration
-$t$.
+which is the probability of being operational at duration $t$.  We use
+the R packages \pkg{survival} and \pkg{survminer} for the KM analysis,
+which is reported in Figure~\ref{fig:km-all-cage-node}. Within each
+batch, separate survival curves are computed for each cage by node
+combination.
 
-We use reliability censoring concepts to use information from units
-that were proactively removed before failing as well as those still in
-operation till the end. We use the R package {\tt survival} for the KM
-analysis.
-\begin{figure*}
-  \centering
-  \includegraphics[width=7in]{figs/km_cage-node_a001.pdf}
-  \caption{Comparison of the old and new batches, including survival
-    differences based on {\tt cage} and {\tt node} GPU locations.}
-  \label{fig:km-all-cage-node}
-\end{figure*}
-
-A sister concept to the survival function is the hazard function:
-\begin{displaymath}
-  H(t) = \frac{f(t)}{S(t)},
-\end{displaymath}
-which is what is sometimes called the "bathtub" curve in reliability.
-
-The Cox proportional hazards (CPH) regression analys
-\cite{Cox1972,Harrell2015} takes the hazard function above and a set
-of covariates $x$ and models $k$s GPU hazard as 
+Another standard survival data methodology is a Cox proportional
+hazards (CPH) regression analys, which can include covariates and
+estimate relative risk based on the covariates
+\cite{Cox1972,Harrell2015}. The CPH regression function takes the form
 \begin{displaymath}
@@
 \begin{figure}
-  \includegraphics[width=\columnwidth]{figs/cox_o001.pdf}
+  \centering
+  \includegraphics[width=0.8\columnwidth]{figs/cox_o001.pdf}
   \caption{GPU hazard ratios from Cox regression model on {\tt old}
@@
 \begin{figure}
-  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
+  \centering
+  \includegraphics[width=0.8\columnwidth]{figs/cox_n001.pdf}
   \caption{GPU hazard ratios from Cox regression model on {\tt new}

### 2020-04-20 tbf.tex
minor correction in tbf section

@@
   \caption{Variation of system-wide MTBF over the lifetime of the machine (time is segmented into quarters, 
-i.e., January through April is Q1 and so on).}
+i.e., January through March is Q1 and so on).}
   \label{fig:MTBF_sys}
@@
 best to consider the variation of MTBF across fixed periods. Herein, we calculate 
-the mean of time-between-failures across four months. Figure~\ref{fig:MTBF_sys}
+the mean of time-between-failures across three months. Figure~\ref{fig:MTBF_sys}
 clearly shows the considerable change in system-wide MTBF from one period to another. 

### 2020-04-20 discussion.tex
Cleaning up.
minor comments
Worked mostly on the discussion section.

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Discussion
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
-\label{sec:discussion}
-\fix{Christian, Jim and everyone}
+\label{section:discussion}
 
-\begin{itemize}
+\fix{A short summary of the observations in this paper to facilitate the discussion below.}
+
+The experienced unexpected reliability issues with the Titan supercomputer raise
+several questions and discussion points. First, there is the mitigation component
+that keeps the system operating at an acceptable level and possibly even restores
+some, if not all, of the lost capability. Second, there is the long-term planning
+component that addresses such scenarios in future-generation supercomputers.
+
+In Titan, the mitigation component included replacing about 11,000 GPUs and
+changing the job scheduling strategy~\cite{8665764}. Replacing about 59\% of
+Titan's 18,688 GPUs helped to improve productivity by restoring some of the lost
+capability, but it was a costly and time-consuming effort. Changing the job
+scheduling strategy to run larger jobs on more reliable and smaller jobs on less
+reliable nodes played a crucial role in maintaining productivity at reduced
+capability.
+
+The options of replacing failed components and employing reliability-aware
+resource management may not be available or cost effective for other
+supercomputing centers dealing with similar unexpected reliability issues in
+their systems. Replacement components may not be readily available and may have
+to be manufactured, which can be impossible for technology that is no longer
+supported by the original manufacturer. There is also a financial aspect, such
+as service contract and/or warranty expiration and the operational budget of a
+supercomputing center not having significant funds for replacement costs. There
+is also the time and financial cost for root-cause analysis and developing a
+hardware and software mitigation strategy in the first place. Reliability-aware
+resource scheduling also has its limitations, as the network architecture needs
+to be taken into account. Titan's 3D torus network creates certainly more
+challenges for efficient job allocations, than the fat tree network of Titan's
+successor, Summit~\cite{olcf:summit}. Nodes associated with the same job need to
+be close to each other on the network for performance. Alternatively, partially
+or completely replacing the aging and failing supercomputer with a new system,
+even if this solution offers less capability, may be more cost effective in the
+end.
+
+Other mitigation components for Titan included matching the checkpointing
+interval of applications with the system's current
+reliability~\cite{bautista-gomez16reducing, 6903564}. The experienced issue here
+was the lack of automated communication of system reliability information and
+the lack of flexibility in application-level checkpoint/restart implementations.
+Constantly measuring and reporting current system reliability is not an automated
+real-time task in currently deployed supercomputers. Applications are typically
+not designed to skip a checkpoint when system reliability is good or take more
+frequent checkpoints when system reliability is getting worse.
+
+The long-term planning component is one of the biggest lessons learned from the
+Titan reliability experience. Today's supercomputers are designed to deal with
+expected reliability issues. However, history has shown~\cite{geist12kill} that
+unexpected reliability issues do occur and do have a significant impact. Vendors
+and manufacturers obviously can not mitigate against all possible reliability
+threats, however, a resilience strategy is needed for future-generation systems
+that is able to deal with emerging unexpected reliability threats in a
+reasonable and cost effective way. In addition to mitigation, better support for
+automated real-time reliability monitoring and reporting is needed, including
+for root-cause analysis.
+
+\fix{\begin{itemize}
 \item Cray and NVIDIA take care of the hardware and some systems and driver
@@
   facing machine disasters like the two episodes (2013 and 2016) on Titan?
-\end{itemize}
+\end{itemize}}
+
+\fix{RA: Recommendation for system operators if they see similar 
+wide-spread failures in their system well before the expected lifetime 
+of the machine. Should they replace all components or just partially
+upgrade the machine. Tradeoff?}

### 2020-04-20 related.tex
Worked mostly on the discussion section.

@@
 Tiwari et al~\cite{10.1145/2807591.2807666} further analyzed the types of GPU errors on Titan, including software/firmware related errors and failures, ECC double-bit errors, and GPU ``off the bus'' and ECC page retirement events.
-This work observed a ECC double-bit mean-time between errors (MTBE) of \~160 hours, or \~ one per week, with 86\% in GPU memory and the rest in the GPU register file.
+This work observed a ECC double-bit mean-time between errors (MTBE) of about 160 hours, or about one per week, with 86\% in GPU memory and the rest in the GPU register file.
 The experienced GPU ``off the bus'' events were caused by a system integration issue that was fixed and not an inherent GPU or GPU memory architecture flaw.

### 2020-04-21 dataprep.tex
minor comments
added side view

@@
 ``signature'' error of eventual GPU failures and a trigger for their
-replacement.
+replacement. \fix{RA: Should we point out that this replacement was only done later on?}
 
@@
     \includegraphics[width=\columnwidth]{figs/TitanCabinet.png}
+    \includegraphics[width=0.9\columnwidth]{figs/TitanCabinetSide.png}
   \end{minipage}
@@
 
-    \vspace{5em}
+    \vspace{0.8\columnwidth}
     \includegraphics[width=\columnwidth]{figs/TitanLayout.png}
   \end{minipage}
-  \caption{Titan cabinet, blade, and floor layout. Cooling airflow is
-    bottom to top within a cabinet, taking ambient air in at the foot
-    and relasing it back out at the top. Cooling fluid is passed
-    through two heat exchangers, one below Cage 0 and one above Cage
-    2.}
+  \caption{Titan cabinet (front view and side view with air flow) on
+    the left, blade layout and floor layout on the right. Cooling
+    airflow is bottom to top within a cabinet, taking ambient air in
+    at the foot and relasing it back out at the top. Cooling fluid is
+    passed through two heat exchange evaporators, one below Cage 0 and
+    one above Cage 2.}
   \label{fig:layout}

### 2020-04-21 survival.tex
added side view

@@
 etc. The technique is able to use censored observations and can also
-split the data into subpopulations and compute separate survival
+split the data into subpopulations to compute separate survival
 curves.
@@
 function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails
-by duration $t$. The survival function is then 
+by duration $t$. The survival function is then its complement
 \begin{displaymath}
-  S(t) = Pr\{T \geq t\} = 1 - F(t) = \int_t^\infty f(x)dx,
+  S(t) = Pr\{T \geq t\} = 1 - F(t) = \int_t^\infty f(x)dx.
 \end{displaymath}
-which is the probability of being operational at duration $t$.  We use
+It is the probability of being operational at duration $t$.  We use
 the R packages \pkg{survival} and \pkg{survminer} for the KM analysis,

### 2020-04-21 dataclean.tex
moved figure

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Viewing and Cleaning the Data}
-\label{section:dataclean}
 \begin{figure*}[bt]
@@
 \end{figure*}
+\section{Viewing and Cleaning the Data}
+\label{section:dataclean}
 As we need to recover durations in operation from this data, correct

### 2020-04-21 survival.tex
getting there

@@
 \end{figure*}
-The construction of a GPU life time is more complex than it initially
-appears because the units are observed only at reboot time, because
-most units were proactively replaced to prevent failure, and because
-some units continue in operation after OTB and DBE events (a second
-reboot may be successful).
+In this section we apply survival analysis methods (time to event
+analysis), which use and combine information across the operational
+lifetimes of all GPUs \cite{survival}. For these analyses, we take
+apart the location string of each GPU into variables {\em col, row,
+  cage, slot, and node} and study the influence of the locations on
+the GPU life times. The construction of a GPU life time is more
+complex than it initially appears because the units are observed only
+at reboot time, because most units were proactively replaced to
+prevent failure, and because some units continue in operation after
+OTB and DBE events (when a second reboot may be successful).
 
@@
 last seen time is taken as its life time. Although most failed units
-experience one of these events at the last seen time, this is
-imperfect because some units experience OTB or DBE events at a time
-different from its last seen time. We found this definition of life
-time the most pragmatic for the study with few imperfect
-observations. One exception is if the last seen time is the final
-Titan shutdown time, in which case such a unit is not considered
-failed.
+experience one of these events at the last seen time, our definition
+is not perfect because some units experience OTB or DBE events at a
+time different from its last seen time. Such units were relatively few
+so we consider this definition of life time as the most pragmatic.
 
-A key concept in survival analysis and in reliability analysis is
-censoring, which is about using information from study subjects whose
-exact failure times are not available or that have not failed. This
-applies to our study because of proactive GPU replacement before
-failure, because most units were still in operation when the system
-was shut down, and also because of observation only at inspection
-times (reboots). We use censoring concepts on the proactive
-replacements and units still in operation at the end. This allows us
-to use all of the GPU life time data, including units that did not
-fail. But we ignore the inspection time censoring, treating removal
-times as exact failure times to reduce the complexity of this
-analysis. We expect that because of the volume of data, this would
-make much difference in our conclusions. However, we are making our
-data publicly available and expect that others, especially in the
-survival analysis community, will dive deeper.
+A key concept in survival analysis is censoring, which is about using
+information from study subjects whose exact failure times are not
+available or that have not failed. This applies to our study because
+of proactive GPU replacement before failure, because most units were
+still in operation when the system was shut down, and also because
+life spans were recorded only at inventory times. We use censoring
+concepts on the proactive replacements and units still in operation at
+the end. This allows us to use all of the GPU life time data,
+including units that did not fail. But we ignore the inspection time
+censoring, treating inventory times as exact failure times to reduce
+the complexity of this analysis. We expect that because of the volume
+of data and length of operation time, this would not make much
+difference in our conclusions. However, we are making our data
+publicly available and expect that others, especially in the survival
+analysis community, will dive deeper.
 
-Kaplan-Meier survival analysis (KM) gives the probability of survival
-beyond a given time. It is a nonparametric technique that makes no
-specific failure model assumptions, such as Weibull, Exponential,
-etc. The technique is able to use censored observations and can also
-split the data into subpopulations to compute separate survival
-curves.
+Kaplan-Meier survival analysis (KM) starts with computing the
+probability of survival beyond a given time. It is a nonparametric
+technique that makes no specific failure model assumptions, such as
+Weibull, Exponential, etc. The technique is able to use censored
+observations and can also split the data into subpopulations to
+compute separate survival curves.
 
-If $T$ is taken to be the random variable of a GPU fail time, then its
-probability density function is $f(t)$ and its cumulative distribution
-function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails
-by duration $t$. The survival function is then its complement
+If $T$ is the random variable of a GPU fail time, then its cumulative
+distribution function $F(t) = Pr\{T < t\}$ gives the probability that
+a GPU fails by duration $t$. The survival function is its complement
 \begin{displaymath}
-  S(t) = Pr\{T \geq t\} = 1 - F(t) = \int_t^\infty f(x)dx.
+  S(t) = Pr\{T \geq t\} = 1 - F(t).
 \end{displaymath}
@@
 which is reported in Figure~\ref{fig:km-all-cage-node}. Within each
-batch, separate survival curves are computed for each cage by node
-combination.
+{\em batch}, separate survival curves are computed for each {\em cage}
+by {\em node} combination.
 
@@
 \begin{displaymath}
-  H_k(t) = H_0(t)e^{\sum\limits_{i=1}^n{\beta x}}
+  h(t) = h_0(t)e^{b_1 x_1 + b_2 x_2 \ldots b_k x_k},
 \end{displaymath}
+where $x_i$ are covariates, $h_0(t)$ is the {\em baseline hazard}, and
+the $b_i$ are coefficients that measure the impact of the
+covariates. The quantity $e^{b_i}$ is the hazard ratio for covariate
+$i$. Its interpretation
+
 That is, a base hazard rate multiplied by a function of

### 2020-04-21 dataprep.tex
done for tonight

@@
 and failure event records. Two types of failure events were collected:
-Double Bit Error (DBE) and Off the Bus (OTB). DBE was found to be the
-``signature'' error of eventual GPU failures and a trigger for their
-replacement. \fix{RA: Should we point out that this replacement was only done later on?}
+Double Bit Error (DBE) and Off the Bus (OTB). Both errors were found
+to be the ``signature'' event of GPU failures and a trigger for their
+replacement.
 
 An inventory of GPU serial numbers and their locations is recorded
-each time the system boots. \fix{Don, you mentioned that this was more
-regular later - daily?} This typically occurs once every few days but
-can be as frequent as a few times per day. An inventory at boot time
-typically occurs over a span of \fix{??} minutes. A separate file is
-produced at each boot time.
+each time the system boots. This typically occurs once every few days
+but can be as frequent as a few times per day. An inventory at boot
+time typically occurs over a span of \fix{how many ??} minutes. A
+separate file is produced at each boot time.
 

### 2020-04-21 intro.tex
done for tonight

@@
 The failures were traced to a failing resistor on the circuitboard
-(not the GPU iself). While the 2016 resistor failures were a surprise, there is a known
-phenomenon~\cite{referenceneeded} that describes a two phase process
-that leads to failure. It also explains why there was a very stable
-period preceding the failures. There is specific decay associated
-with the electronic components where, as these products age, they will
-eventually fail thresholds for guard band margin and other things. In
-a lot of cases, they will just fail as well. \fix{Give a reference and
-  expand?}
+(not the GPU chip iself). Throughout this paper, when we refer to
+replacing the GPU, we refer to the entire circuitboard along with its
+GPU chip. While the 2016 resistor failures were a surprise, there is a
+known phenomenon~\cite{referenceneeded} that describes a two phase
+process that leads to failure. It also explains why there was a very
+stable period preceding the failures. There is specific decay
+associated with the electronic components where, as these products
+age, they will eventually fail thresholds for guard band margin and
+other things. In a lot of cases, they will just fail as
+well. \fix{Give a reference and expand?}
 
@@
 cycle and later, essentialy units installed near the beginning of 2014
-and later. \fix{Will we (are we required) make the data publicly
-  available?} The data used in this study, representing over 100,000
-years of operation may be the largest publicly available GPU
-reliability data set.
+and later. We will make the data as well as our analysis codes
+producing the graphics available in a public repository (see
+\fix{reference?}).  The data used in this study, while only a 7 MB
+file, represents over 100,000 years of operation and may be the
+largest publicly available GPU reliability data set.
 

### 2020-04-21 survival.tex
done for tonight

@@
 {\em batch}, separate survival curves are computed for each {\em cage}
-by {\em node} combination.
+by {\em node} combination. Along with the survival curve estimate,
+this analysis provides 95\% confidence region for survival probability
+shaded around the curves.
+
+It appears that transport of cooling air provides a complete
+explanation for relative differences in cage and node survival rates
+in the old batch.  We reach this conclusion with reference to
+Figure~\ref{fig:layout} and relative positions of cages and nodes
+within a cabinet. Both cage and node differences in survival
+probabilities can be explained by an inverse relationship with the
+distance to the bottom of the cabinet, where cooling air is forced
+through the cabinet to the top. The survival curves can be ordered
+({\em cage 0, cage 1, cage 2}) in decreasing order of survival. As the
+blades are placed vertically within the cabinet, pairs of nodes too
+experience the same relationship with distance to the bottom, nodes 2
+and 3 having lower failure rates than nodes 0 and 1.
+
+The survival curves for the new batch are uninteresting as very few
+failures have occurred in the three years they have been in operation.
+
+We don't see a ``bathtub curve'' phenomenon, in fact the opposite is
+apparent in cages 1 and 2 of the old batch. The slope of the survival
+curve is related to the hazard rate. There does not seem to be an
+early ``infant mortality'' period nor a ``wearout'' phenomenon at the
+end.
 
@@
 covariates. The quantity $e^{b_i}$ is the hazard ratio for covariate
-$i$. Its interpretation
+$i$.
 
-That is, a base hazard rate multiplied by a function of
-covariates. The key assumption in the CPH model is that the "hazards"
-are multipliers on the "baseline hazard" but the estimate of the
-baseline hazard is nonparametric, meaning it makes no specific
-distributional assumption and just learns from the data (unlike
-analyses that assume for example a Weibul or an exponential
-model). The multiplicative assumption is that the hazard curves do not
-cross and are multiples of each other. The multiplier is the *hazard
-coefficient*. For example, if the baseline is *node0* and the *node2*
-hazard is 2, then *node2* sees twice as many failures as *node0* on
-average.
-
-For our comparison of hazards, we use as *covariates* all the unpacked
-location variables, indicators whether the GPU has been moved, and our
-notion of old and new GPUs based on first install date. Since many
-GPUs experience time at more than one location, we use the longest
-period location in this analysis.
-
-First, we do a univariate analysis for each covariate separately and
-then we look at a multivariate view of this. Having multiple
-covariates is called "multiple regression" as opposed to "multivariate
-regression" reserved for cases with several response variables. The
-univariate view is comparable to what would one see in a simple time
-between failures analysis for a given population split. It can be
-affected by sampling differences for other covariates. For example, if
-the strongest effect is seen in a variable not used in a split and its
-values are unevenly distributed across the split, its effect can leak
-into the split. This is why we also do a multivariate analysis, where
-the effect of each covariate is adjusted for all the other
-covariates.
-
+CPH is considered a semi-parametric model as there are no assumptions
+about the shape of the baseline hazard function. But its main
+assumption is that the hazards are proportional. There are a number of
+tests for this, including graphical diagnostics in the \pkg{survminer}
+package as well as checking that survival curves for categorical
+covariates do not cross. We ran these diagnostics and concluded that
+the hazards for our location categories are approximately
+proportional. However, the {\em batch} survival functions do cross and
+so we fit the model to the new and the old batches separately. The
+results are presented in Figures~\ref{fig:cox-old}
+and~\ref{fig:cox-new}.
 \begin{figure}
@@
     batch.}
+  \label{fig:cox-old}
 \end{figure}
@@
     batch.}
+  \label{fig:cox-new}
 \end{figure}
 
-Higher numbered columns appear to have a higher (up to about 2x)
-failure rate. There actually seems to be a hint of three groups that
-have similar rates: 0-6 baseline, 7-16 are 1.5x baseline, and 17-24
-are 2x baseline.
-
-Higher numbered rows tend to have a lower (up to about 0.5x) failure rate.
-
-This seems surprisingly strong! Higher numbered cages have failure
-rates 5x and 16x higher than cage = 0. In fact, this seems so strong
-that its effect could be leaking into all other comparisons under
-uneven sampling!
-
-Not much difference between slots. They are all close to the baseline
-slot0.
-
-Interestingly, nodes 2 and 3 have half the failure rate of nodes 0 and
-1. Is this related to the cage variable or is there a real location
-effect?
-
-Now the multivariate Cox regression analysis, where each factor is
-adjusted for all the other factors. This can fix some uneven sampling
-and multivariate dependence.
-
-The col groups persist and are even stronger. The row effect is also
-stronger. The cage effect is out of this world! slot is still without
-any effect. node effect also persists and is a bit stronger.
-
-Let’s separate the “old” and “new” batches of GPUs and run separate
-Cox models.
+\fix{Need to add interpretation}

### 2020-04-21 survival.tex
improved text

@@
 
-The survival curves for the new batch are uninteresting as very few
-failures have occurred in the three years they have been in operation.
+Very few failures have occurred in the new batch and they are clarly
+less prone to failure at the 2.5 year mark. There is a slight change
+at the 3 year mark of cages 1 and 2 but it is also accompanied with a
+rise in uncertainty and survival is still well above levels in the old
+batch.
 
@@
 early ``infant mortality'' period nor a ``wearout'' phenomenon at the
-end.
+end. Rather, we see a steeper slope (higher hazard rate) in the
+middle, associated with the unexpected resistor failures.
 
-Another standard survival data methodology is a Cox proportional
-hazards (CPH) regression analys, which can include covariates and
-estimate relative risk based on the covariates
+To get more comparison power across the locations, we can use a
+technique that in a sense averages over time. Cox proportional hazards
+(CPH) regression analys, can include covariates and estimates relative
+risk averaged ofer time based on the covariates
 \cite{Cox1972,Harrell2015}. The CPH regression function takes the form
@@
 the hazards for our location categories are approximately
-proportional. However, the {\em batch} survival functions do cross and
-so we fit the model to the new and the old batches separately. The
-results are presented in Figures~\ref{fig:cox-old}
+proportional. However, survival functions partitioned on {\em batch}
+do cross and so we fit the model to the new and the old batches
+separately. The results are presented in Figures~\ref{fig:cox-old}
 and~\ref{fig:cox-new}.

### 2020-04-21 survival.tex
text tweaks

@@
 technique that in a sense averages over time. Cox proportional hazards
-(CPH) regression analys, can include covariates and estimates relative
-risk averaged ofer time based on the covariates
+(CPH) regression analysis, can include covariates and estimates relative
+risk averaged over time based on the covariates
 \cite{Cox1972,Harrell2015}. The CPH regression function takes the form
@@
 the $b_i$ are coefficients that measure the impact of the
-covariates. The quantity $e^{b_i}$ is the hazard ratio for covariate
-$i$.
+covariates. The quantity $e^{b_i}$ is the hazard ratio (increased
+average risk over baseline) for covariate $i$.
 

### 2020-04-21 discussion.tex
Improved the discussion section.

@@
 
-\fix{A short summary of the observations in this paper to facilitate the discussion below.}
+\fix{Possibly a short summary of the observations in this paper to facilitate the discussion below.}
 
@@
 
+%%%%%%%%%%%%
+% Mitigation
+%%%%%%%%%%%%
+\subsection{Mitigation}
+\label{section:mitigation}
+
 In Titan, the mitigation component included replacing about 11,000 GPUs and
@@
 reliable nodes played a crucial role in maintaining productivity at reduced
-capability.
+capability. However, jobs running on larger portions of the system, utilizing
+most or all of its resources, were still impacted by the failing GPUs and
+corresponding low system MTBF.
 
@@
 resource management may not be available or cost effective for other
-supercomputing centers dealing with similar unexpected reliability issues in
-their systems. Replacement components may not be readily available and may have
-to be manufactured, which can be impossible for technology that is no longer
-supported by the original manufacturer. There is also a financial aspect, such
-as service contract and/or warranty expiration and the operational budget of a
-supercomputing center not having significant funds for replacement costs. There
-is also the time and financial cost for root-cause analysis and developing a
-hardware and software mitigation strategy in the first place. Reliability-aware
-resource scheduling also has its limitations, as the network architecture needs
-to be taken into account. Titan's 3D torus network creates certainly more
-challenges for efficient job allocations, than the fat tree network of Titan's
-successor, Summit~\cite{olcf:summit}. Nodes associated with the same job need to
-be close to each other on the network for performance. Alternatively, partially
-or completely replacing the aging and failing supercomputer with a new system,
-even if this solution offers less capability, may be more cost effective in the
-end.
+supercomputing centers dealing with similar unexpected reliability issues.
+%
+Replacement components may not be readily available and may have to be
+manufactured, which can be impossible for technology that is no longer supported
+by the original manufacturer. Experienced errors or failures may be caused by
+software, which can be difficult to fix if the software itself or the deployed
+version is no longer supported by the vendor.
+
+There is also a financial aspect, such as service contract and/or warranty
+expiration and the operational budget of a supercomputing center not having
+significant funds for replacement/reengineering costs or a manufacturer or
+vendor not paying for them.
+%
+There is additionally the time and financial cost for root-cause analysis and
+developing a hardware and/or software mitigation strategy in the first place.
+Fixing a problem requires finding and understanding it, which can be difficult 
+in today's complex systems and may require knowledge that only the original 
+manufacturer or vendor possesses.
+
+Reliability-aware resource scheduling also has its limitations, as the network
+architecture needs to be taken into account. Titan's 3D torus network creates
+certainly more challenges for efficient job allocations, than the fat tree
+network of Titan's successor, Summit~\cite{olcf:summit}. Nodes associated with
+the same job need to be close to each other on the network for maximum
+application performance.
+%
+Alternatively, partially or completely replacing the aging and failing
+supercomputer with a new system, even if this solution offers less capability,
+may be more cost effective in the end.
 
@@
 Constantly measuring and reporting current system reliability is not an automated
-real-time task in currently deployed supercomputers. Applications are typically
-not designed to skip a checkpoint when system reliability is good or take more
-frequent checkpoints when system reliability is getting worse.
+real-time task in many currently deployed supercomputers. Applications are
+typically not designed to skip a checkpoint when system reliability is good or
+take more frequent checkpoints when system reliability is getting worse.
+
+%%%%%%%%%%%%%%%%%%%%
+% Long-term Planning
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Long-term Planning}
+\label{section:planning}
 
@@
 
-\fix{\begin{itemize}
-\item Cray and NVIDIA take care of the hardware and some systems and driver
-  software so moving and troubleshooting hardware is their cost.
-\item Checkpoint strategies, allocation strategies, job resource
-  management, are all under OLCF control. What can OLCF do/change to
-  mitigate fallout in terms of delivering computing to customers when
-  facing machine disasters like the two episodes (2013 and 2016) on Titan?
-\end{itemize}}
-
-\fix{RA: Recommendation for system operators if they see similar 
-wide-spread failures in their system well before the expected lifetime 
-of the machine. Should they replace all components or just partially
-upgrade the machine. Tradeoff?}
+Supercomputer center policies regarding reliability monitoring, resource
+allocation and checkpointing strategies need to be powerful and flexible enough
+to facilitate mitigation for such unexpected reliability threats. System
+acquisition contracts may need to include performance requirements for degraded
+operation, such as a 60\% performance capability if the MTBF drops by a
+magnitude.

### 2020-04-21 discussion.tex
Improved the discussion section.

@@
 acquisition contracts may need to include performance requirements for degraded
-operation, such as a 60\% performance capability if the MTBF drops by a
-magnitude.
+operation, such as a certain percentage of performance capability if the MTBF
+drops by a magnitude.

### 2020-04-21 survival.tex
continunig with Cox

@@
 
-\fix{Need to add interpretation}
+We find that the average hazard ratios strongly correlate with
+detailed nuances of the system cooling architecture.  The figures give
+hazard ratios for each of the location variables, giving the ratio for
+each level compared to its first (0) level (the 0 level is always
+1). Due to the exponential nature of the CPH model, the horizontal
+hazard ratio axis is on a log scale. The estimates include
+uncertainty, which too is most reliably computed on a log scale (see,
+for example, \cite{Ostrouchov88}).

### 2020-04-21 survival.tex
interpret cox

@@
 hazard ratios for each of the location variables, giving the ratio for
-each level compared to its first (0) level (the 0 level is always
-1). Due to the exponential nature of the CPH model, the horizontal
-hazard ratio axis is on a log scale. The estimates include
+each level compared to its first (0) level (consequently the 0 level
+is always 1). Due to the exponential nature of the CPH model, the
+horizontal hazard ratio axis is on a log scale. The estimates include
 uncertainty, which too is most reliably computed on a log scale (see,
-for example, \cite{Ostrouchov88}).
+for example, \cite{Ostrouchov88}). In the figures, we also include the
+number of units, N, at risk and the number of events that occurred in
+each category. All the factors ({\em col, row, cage, slot, node}) are
+balanced with respect to the nunber of units at risk and consequently
+nearly orthogonal (each level of a factor contains all levels of the
+other factors) providing an almost ``designed'' experiment nature to
+this analysis.
+
+Our interpretation of correlation with detailed nuances of the cooling
+system comes mostly from the hazard ratios for the old batch. The new
+batch has mot had many failure events and the uncertainty bars of
+nearly all ratios include 1, which is no difference. In the old batch,
+we see that {\em cage} has the strongest effect, putting the highest
+hazard ratio on cage 2, which is consistent with lowest survivals in
+the KM analysis earlier. Its ratio value near 20 has to be interpreted
+with caution because it is a time averaged value on a log scale. This
+aspect is not captured by the uncertainty, which accounts for the
+randomnes of the failures and not for the geometric time averaging
+\cite{coxhazardinterpret}.

### 2020-04-21 survival.tex
spell

@@
 
-Very few failures have occurred in the new batch and they are clarly
+Very few failures have occurred in the new batch and they are clearly
 less prone to failure at the 2.5 year mark. There is a slight change
@@
 each category. All the factors ({\em col, row, cage, slot, node}) are
-balanced with respect to the nunber of units at risk and consequently
+balanced with respect to the number of units at risk and consequently
 nearly orthogonal (each level of a factor contains all levels of the
@@
 aspect is not captured by the uncertainty, which accounts for the
-randomnes of the failures and not for the geometric time averaging
+randomness of the failures and not for the geometric time averaging
 \cite{coxhazardinterpret}.

### 2020-04-21 survival.tex
more cox

@@
 shaded around the curves.
+% Inserting figure here so it appears at top of next page
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{figs/cox_o001.pdf}
+  \caption{GPU hazard ratios from Cox regression model on {\tt old}
+    batch.}
+  \label{fig:cox-old}
+\end{figure}
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
+  \caption{GPU hazard ratios from Cox regression model on {\tt new}
+    batch.}
+  \label{fig:cox-new}
+\end{figure}
 
@@
 CPH is considered a semi-parametric model as there are no assumptions
-about the shape of the baseline hazard function. But its main
+about the shape of the baseline hazard function. Its strongest
 assumption is that the hazards are proportional. There are a number of
@@
 and~\ref{fig:cox-new}.
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\columnwidth]{figs/cox_o001.pdf}
-  \caption{GPU hazard ratios from Cox regression model on {\tt old}
-    batch.}
-  \label{fig:cox-old}
-\end{figure}
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\columnwidth]{figs/cox_n001.pdf}
-  \caption{GPU hazard ratios from Cox regression model on {\tt new}
-    batch.}
-  \label{fig:cox-new}
-\end{figure}
 
@@
 number of units, N, at risk and the number of events that occurred in
-each category. All the factors ({\em col, row, cage, slot, node}) are
-balanced with respect to the number of units at risk and consequently
-nearly orthogonal (each level of a factor contains all levels of the
-other factors) providing an almost ``designed'' experiment nature to
-this analysis.
+each category.
 
-Our interpretation of correlation with detailed nuances of the cooling
-system comes mostly from the hazard ratios for the old batch. The new
-batch has mot had many failure events and the uncertainty bars of
-nearly all ratios include 1, which is no difference. In the old batch,
-we see that {\em cage} has the strongest effect, putting the highest
-hazard ratio on cage 2, which is consistent with lowest survivals in
-the KM analysis earlier. Its ratio value near 20 has to be interpreted
-with caution because it is a time averaged value on a log scale. This
+All the factors ({\em col, row, cage, slot, node}) in the old batch
+are balanced with respect to the number of units at risk and
+consequently nearly orthogonal (each level of a factor contains all
+levels of the other factors) seemingly an almost ``designed''
+experiment nature to this analysis. This is not the case for the new
+batch, where the cage levels have very different numbers of units at
+risk. But this also points out that even for the old batch the balance
+holds only at the outset and as life proceeds, failing units are
+replaced with new units and the balance degrades.
+
+Our interpretation of correlation with details of the cooling system
+comes mostly from the hazard ratios for the old batch. The new batch
+has mot had many failure events and the uncertainty bars of nearly all
+ratios include 1, which is no difference. In the old batch, we see
+that {\em cage} has the strongest effect, putting the highest hazard
+ratio on cage 2, which is consistent with lowest survivals in the KM
+analysis earlier. Its ratio value near 20 has to be interpreted with
+caution because it is a time averaged value on a log scale and suffers
+from the degrading balance mentioned in the previous paragraph. This
 aspect is not captured by the uncertainty, which accounts for the
 randomness of the failures and not for the geometric time averaging
-\cite{coxhazardinterpret}.
+\cite{coxhazardinterpret}. Consequently, we interpret the pattern of
+relative hazards rather than actual ratio magnitudes. More in-depth
+analysis with penalized estimation methods like \cite{bender2019} can
+take such balance issues in the exposure history into account.

### 2020-04-21 tbf.tex
added new result of system-wide MTBF across old and new partitions

@@
 
-\begin{figure}[bt]
-  \begin{center}
-    \includegraphics[width=\columnwidth]{figs/MTBF_CageWise.pdf}
-  \end{center}
-  \caption{MTBFs due to either DBEs or OTBs across the GPUs located in different cages.}
-  \label{fig:CageWise_MTBFs}
-\end{figure}
+%\begin{figure}[bt]
+%  \begin{center}
+%    \includegraphics[width=\columnwidth]{figs/MTBF_CageWise.pdf}
+%  \end{center}
+%  \caption{MTBFs due to either DBEs or OTBs across the GPUs located in different cages.}
+%  \label{fig:CageWise_MTBFs}
+%\end{figure}
 
-\fix{Figure~\ref{fig:CageWise_MTBFs} shows temperature effect clearly. Should we move this 
-result to later in the paper?}
+%\fix{Figure~\ref{fig:CageWise_MTBFs} shows temperature effect clearly. Should we move this 
+%result to later in the paper?}
 
@@
   \caption{Variation of system-wide MTBF over the lifetime of the machine (time is segmented into quarters, 
-i.e., January through March is Q1 and so on).}
+i.e., January through March is Q1, so on).}
   \label{fig:MTBF_sys}
+\end{figure}
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL.pdf}
+  \end{center}
+  \caption{Variation of system-wide MTBF across hypothetical new and old partitions of the machine over time.}
+  \label{fig:MTBF_sys_NewOld}
 \end{figure}
@@
 included in the figure to better visualize various change points). This period marked the start of 
-a rapid decline in system-wide MTBF until 2016-Q4. The rapid rise of the number of failures before this 
+a rapid decline in system-wide MTBF until 2016-Q4. The rapid rise in the number of failures before this 
 quarter can be seen in Figure~\ref{fig:NumFails_sys}. When GPU replacements start to take place in late 
@@
 being unavailable while the reworks were taking place. A single replacement cycle lasted multiple weeks. 
-\fix{need to verify how long the replacement took}
+%\fix{need to verify how long the replacement took}
 There is no definite way to incorporate this unavailability of the machine into MTBF analysis, so it has to
 be said that overall there has been an increase in MTBF of the machine due to the replacements. 
-For example, before 2017-Q1 MTBF as low as 2.7 hours is noted, whereas the lowest MTBF is 5.9 hours after this 
+For example, before 2017-Q1, a MTBF as low as 2.7 hours is noted, whereas the lowest MTBF is 5.9 hours after this 
 period. With such huge variations in system MTBF, it is difficult to reliably run applications even while using 
@@
 
-The mean-time-between DBE and OTB failures are separately tracked as well in this analysis.
+The mean-time-between DBE and OTB failures are also separately tracked in this analysis.
 One observation is the drastic increase in mean-time-between OTB failures after the GPU replacements.
 This is also evident from the reduction of the absolute number of OTB failures in Figure~\ref{fig:NumFails_sys}.
-However, the machine's MTBF is determined by the weakest link and DBE events tend to dictate it.
+However, the system MTBF is determined by the weakest link and the occurrence of DBE events tends to dictate it.
 Even though the replacements helped to increase the MTBF due to DBE events by a factor, the overall system 
-reliability is dictated by the components with the most age in the system. There is an upward trend towards 
-the end in the number of DBE failures in Figure~\ref{fig:NumFails_sys} which almost all of them are due to 
-older GPUs in the system. The old GPUs that occupy a minority portion of the machine after GPU replacements
- and have a disproportionately high number of failures play a major role in determining the state of the 
-machine as noted in the next section.
+reliability is dictated by the components with the most age in the system. There is a re-emergence of upward trend towards 
+the end of life in the number of DBE failures in Figure~\ref{fig:NumFails_sys} which almost all of them are due to 
+older GPUs in the system.
+
+To better understand the difference in reliability of newer and old portions of the machine, 
+Figure~\ref{fig:MTBF_sys_NewOld} shows the drastic difference in system MTBF measured in two hypothetical
+partitions of the machine starting from the time when the majority of GPU replacements have been completed. 
+The new partition of the machine has always 12X better MTBF than the older partition
+of the machine. For example, in 2018-Q4, the MTBF recorded across old partition is about 7.9 hours, whereas it is about
+96 hours (4 days) for the newer partition. The implications of such a huge disparity on applications running on the 
+system are discussed in Section~\ref{section:discussion}. The insignificant difference in MTBF measured across all GPUs and
+old partition of the machine can also be noted in Figure~\ref{fig:MTBF_sys_NewOld}.  
+Therefore, the old GPUs that occupy a minority portion of the machine after GPU replacements, 
+and have a disproportionately high number of failures and contribute towards a low system MTBF play a major role 
+in determining the overall reliability of the machine.
+The next section quantifies the survivability of this machine with its varied MTBF characteristics beyond its operational lifetime.
+\fix{George, can you please check the last sentence, where I make a connection with the next section.}
 
@@
   \caption{The number of DBE and OTB failures observed over the lifetime of the machine.
-A distinction is made for GPU failures happening across old GPUs to signify the number 
+A distinction is made for GPU failures happening across old GPUs to highlight the number 
 of failures across the newer GPUs.}

### 2020-04-22 intro.tex
updated fig

@@
 \label{section:introduction}
-
 The Cray XK7 Titan supercomputer~\cite{olcf:titan} was the \#1 system in the world for a
@@
 
+\begin{figure*}[tb]
+  \centering
+  \includegraphics[width=\textwidth]{figs/chronology001.pdf}
+  \caption{Volume of GPU swaps detected on Titan at inventories. Our
+    analysis begins with units in place by 2014. A break-in period
+    before 2014 involved working out mechanical assembly issues at
+    PCIe connectors.}
+  \label{fig:chronology}
+\end{figure*}
 Figure~\ref{fig:chronology} illustrates the chronology of the rework
@@
 periodic inventories.
-\begin{figure*}[tbp]
-  \centering
-  \includegraphics[width=\textwidth]{figs/chronology001.pdf}
-  \caption{Volume of GPU swaps detected at inventories. The break-in
-    period before 2014 involved working out mechanical assembly issues
-    at PCIe connectors
-    frequent dates.}
-  \label{fig:chronology}
-\end{figure*}
 The first two rework cycles before 2014 can be characterized as a

### 2020-04-22 dataclean.tex
tweaks

@@
 
-As most analysis software uses table-like data, we fill the needed
-repeats of values missing in the raw data (see
-Fig.~\ref{fig:dataraw}). First, we reduce the data to the GPU that
-were removed later than 2013 to focus our analysis epoch on units
-installed in the second rework cycle or later. After this reduction,
-there were stil six older units remaining, which we also removed to
-have a clean third epoch set of units for the analysis. We do some
-further processing to handle time overlaps in a tiny fraction (under
-0.0007 in GPU life and 0.0002 in location life) of recorded life in
-the raw records by simply dropping the overlapping life times.
+As most analysis software relies on rectangular table-like data, we
+fill the needed repeats of values missing in the raw data (see
+Fig.~\ref{fig:dataraw}). To focus our analysis on data after the first
+two rework cycles in the break-in period, we first reduce the data to
+the GPUs that were installed shortly before 2014 started. This was
+done by removing data for any units with a {\em remove} date before
+2014. After this reduction, there were stil six older units remaining,
+which we also removed to have a clean set of units for the
+analysis. We do some further processing to handle time overlaps in a
+tiny fraction (under 0.0007 in GPU life and 0.0002 in location life)
+of recorded life in the raw records by simply dropping the overlapping
+life times.
 

### 2020-04-22 survival.tex
finished cox

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Survival Analysis}
-\label{section:survival}
-\renewcommand{\pkg}[1]{\textsf{#1}}
 \begin{figure*}
@@
 \end{figure*}
+\section{Survival Analysis}
+\label{section:survival}
+\renewcommand{\pkg}[1]{\textsf{#1}}
 In this section we apply survival analysis methods (time to event
@@
 holds only at the outset and as life proceeds, failing units are
-replaced with new units and the balance degrades.
+replaced with new units and the balance degrades because failures are
+location-dependent.
 
@@
 analysis with penalized estimation methods like \cite{bender2019} can
-take such balance issues in the exposure history into account.
+take such balance issues in the exposure history into account and
+provide time-dependent hazard estimates.
+
+In addition to the strong {\em cage} and {\em node} hazard rate
+differences that increase with distance from the bottom of the
+cabines, seen in both KM and CPH results, there is a weaker but
+peculiar pattern in the {\em row} and {\em col} hazard rates. The
+different behavior in {\em col} 0-11 from {\em col} 12-24 is likely
+due to separate supply lines of coolant to the two sets. But the
+interleaving pattern of the first set is still a mystery. We also have
+yet to explain the apparent row effects. A very minor
+airflow effect across slots also seems to be present as it mimics
+faster airflow for the middle slots.

### 2020-04-22 intro.tex
intro tweak

@@
 \label{section:introduction}
-The Cray XK7 Titan supercomputer~\cite{olcf:titan} was the \#1 system in the world for a
-very long time~\cite{top500list}, and has remained a critically important computer
-system through the end of its life in the Summer of 2019. It defied
-scale with 18,688 individual compute nodes and delivered tens of
-billions of computing hours to the U.S. Department of Energy
-mission-critical programs for nearly 7 years.
+The Cray XK7 Titan supercomputer~\cite{olcf:titan} was the \#1 system
+in the world for a very long time~\cite{top500list}, and has remained
+a critically important computer system through the end of its life in
+the Summer of 2019. It defied scale with 18,688 individual compute
+nodes and delivered tens of billions of computing hours to the
+U.S. Department of Energy mission-critical programs for nearly 7
+years.
 
@@
 cycles and stable periods as indicated by the number of GPU changes at
-periodic inventories.
-The first two rework cycles before 2014 can be characterized as a
-break-in period on a new system that is first of its kind and pushes
-many technological boundaries, in this case mainly regarding blade
-mechanical assembly. The resulting field engineering generated high
-GPU swap counts. Late 2013 begins a very long period of stable
-operation with very few issues until about mid 2016, when GPU failures
-begin to rise. This results in the final rework cycle, replacing
-11,000 GPUs throughout much of 2017.
+periodic inventories.  The first two rework cycles before 2014 can be
+characterized as a break-in period on a new system that is first of
+its kind and pushes many technological boundaries, in this case mainly
+regarding blade mechanical assembly. The resulting field engineering
+and data collection changes generated high GPU swap counts before
+2014. Late 2013 begins a very long period of high reliability and
+stable operation with very few issues until about mid 2016, when GPU
+failures begin to rise. This results in the final rework cycle,
+replacing 11,000 GPUs throughout much of 2017. The peak at the end in
+the figure is an artifact of turning off or ``removing'' all the
+units.
 
-The failures were traced to a failing resistor on the circuitboard
-(not the GPU chip iself). Throughout this paper, when we refer to
-replacing the GPU, we refer to the entire circuitboard along with its
-GPU chip. While the 2016 resistor failures were a surprise, there is a
-known phenomenon~\cite{referenceneeded} that describes a two phase
-process that leads to failure. It also explains why there was a very
-stable period preceding the failures. There is specific decay
-associated with the electronic components where, as these products
-age, they will eventually fail thresholds for guard band margin and
-other things. In a lot of cases, they will just fail as
-well. \fix{Give a reference and expand?}
+Finding the root cause for the 2016 failures took a great deal of
+testing and involvement of materials science and microscopy
+researchers. The failures were traced to a resistor on the GPU
+circuit board (not the GPU chip iself). Throughout this paper, when we
+refer to replacing the GPU, we refer to the entire circuitboard along
+with its GPU chip.
 
-Our analysis focuses on GPUs that were installed in the second rework
-cycle and later, essentialy units installed near the beginning of 2014
-and later. We will make the data as well as our analysis codes
-producing the graphics available in a public repository (see
-\fix{reference?}).  The data used in this study, while only a 7 MB
-file, represents over 100,000 years of operation and may be the
-largest publicly available GPU reliability data set.
+Our analysis focuses on GPU boards that were installed in the second
+rework cycle and later, essentially units installed near the beginning
+of 2014 and later. We will make the data as well as our analysis codes
+producing the graphics available in a public repository (see,
+\cite[repository]{repositoryreference}).  While the data is only a 7 MB
+file, due to the size of Titan it represents over 100,000 years of GPU
+operation and may be the largest publicly available GPU reliability
+data set. Our hope is that others will use it for reliability research
+and go beyond the results presented here.
 

### 2020-04-22 related.tex
Fixed related work and discussion section.

@@
 Prior work in investigating Titan's reliability includes a significant amount of work that covers different aspects.
+However, it did not investigate the atypical failure mode discussed in this paper and failure data over the full lifetime of the system.
 
@@
 %
-Tiwari et al. and Nie et al. did not address the atypical failure mode discussed in this paper and only used Titan error and failure data from June 2013 to February 2015.
+Tiwari et al.
+and Nie et al.
+did not address the atypical failure mode discussed in this paper.
+It also only used Titan error and failure data from June 2013 to February 2015, excluding early failures and the failure mode discussed in this paper (which occurred after February 2015).
 

### 2020-04-22 abstract.tex
tweak
a

@@
 operation was forced to execute three very significant rework cycles,
-two on the mechanical assembly and one on the GPUs. We write about the
-last rework cycle and a reliability analysis of the GPUs. Using
-statistical survival analysis techniques, we find that GPU reliability
-is dependent on heat dissipation to an extent that strongly correlates
-with detailed nuances of the system cooling architecture. In addition
-to describing some of the system history, the data collection, data
-cleaning, and our analysis of the data, we provide reliability
-recommendations for designing future state of the art supercomputing
-systems and their operation.
+two on the GPU mechanical assembly and one on the GPU circuitboards. We
+write about the last rework cycle and a reliability analysis of over
+100,000 operation years in the GPU lifetimes, which correspond to
+Titan's 6 year long productive period after an initial break-in
+period. Using time between failures analysis and statistical survival
+analysis techniques, we find that GPU reliability is dependent on heat
+dissipation to an extent that strongly correlates with detailed
+nuances of the system cooling architecture. In addition to describing
+some of the system history, the data collection, data cleaning, and
+our analysis of the data, we provide reliability recommendations for
+designing future state of the art supercomputing systems and their
+operation. We make the data and our analysis codes publicly available.

### 2020-04-22 intro.tex
high-level

@@
 rework cycles, two on the mechanical assembly affecting the PCIe
-connector from the GPU daughtercard to the motherboard, and a
-replacement of about 11,000 GPU assemblies because of a failing
-resistor on a printed circuit board. We write primarily about the epoch
-that includes this last rework cycle. This includes Titan's most
-stable and failure free period and contains the most reliable data on
-GPU operation.
+connector from the GPU daughtercard to the motherboard, and one to
+replace about 11,000 GPU assemblies because of a failing resistor on
+their printed circuit board. We write primarily about the GPU
+operation epoch that includes this last rework cycle. This epoch of
+nearly 6 years includes Titan's most stable and failure free period
+and contains the most reliable data on GPU operation.
 
@@
 cycles and stable periods as indicated by the number of GPU changes at
-periodic inventories.  The first two rework cycles before 2014 can be
-characterized as a break-in period on a new system that is first of
-its kind and pushes many technological boundaries, in this case mainly
-regarding blade mechanical assembly. The resulting field engineering
-and data collection changes generated high GPU swap counts before
-2014. Late 2013 begins a very long period of high reliability and
-stable operation with very few issues until about mid 2016, when GPU
-failures begin to rise. This results in the final rework cycle,
+periodic inventories.  The first two rework cycles before 2014
+involving blade mechanical assembly can be characterized as a break-in
+period on a new system that is first of its kind and pushes many
+technological boundaries. The resulting field engineering as well as
+data collection changes are responsible for the high GPU swap counts
+before 2014. Late 2013 begins a very long period of high reliability
+and stable operation with very few issues until about mid 2016, when
+GPU failures begin to rise. This results in the final rework cycle,
 replacing 11,000 GPUs throughout much of 2017. The peak at the end in
 the figure is an artifact of turning off or ``removing'' all the
-units.
+units as Titan is shut down and disassembled.
 
@@
 rework cycle and later, essentially units installed near the beginning
-of 2014 and later. We will make the data as well as our analysis codes
-producing the graphics available in a public repository (see,
-\cite[repository]{repositoryreference}).  While the data is only a 7 MB
-file, due to the size of Titan it represents over 100,000 years of GPU
-operation and may be the largest publicly available GPU reliability
-data set. Our hope is that others will use it for reliability research
-and go beyond the results presented here.
+of 2014 and later. Figure~\ref{fig:chronology} is the only one
+involving data prior to 2014. All analyses and figures that follow use
+only data from 2014 onward. We make the data as well as our analysis
+codes to reproduce the graphics available in a public repository (see,
+\cite[repository]{repositoryreference}).  While the data is only a 7
+MB file, due to the size of Titan it represents over 100,000 years of
+GPU operation and may be the largest publicly available GPU
+reliability data set. Our hope is that others will use it for
+reliability and survival analysis research and go beyond the results
+presented here.
 
@@
 relative hazard rates in various locations are in
-Sec.~\ref{section:survival}. Finally, we discuss our main conclusions
-and recommendations in Sec.~\ref{section:discussion}.
+Sec.~\ref{section:survival}. A high-level difference between a TBF
+analysis and survival analysis (SA) is that TBF studies largely how
+GPU reliability affects the machine, wheras SA studies largely how the
+machine affects GPU reliability.  Finally, we discuss our main
+conclusions and recommendations in Sec.~\ref{section:discussion}.

### 2020-04-22 survival.tex
chill

@@
 different behavior in {\em col} 0-11 from {\em col} 12-24 is likely
-due to separate supply lines of coolant to the two sets. But the
-interleaving pattern of the first set is still a mystery. We also have
-yet to explain the apparent row effects. A very minor
-airflow effect across slots also seems to be present as it mimics
-faster airflow for the middle slots.
+due to separate supply lines of chilled water to the two sets. While
+the temperature was a consistent 42$^\circ$F, there may have been
+minor flow or pressure differences. But the interleaving pattern of
+the first set is still a mystery. We also have yet to explain the
+apparent row effects. A very minor airflow effect across slots also
+seems to be present as it mimics faster airflow for the middle slots.

### 2020-04-22 dataclean.tex
next

@@
 \label{section:dataclean}
-As we need to recover durations in operation from this data, correct
+As we need to recover durations of GPU operation from this data, correct
 processing involves time adjustments for switching between daylight
@@
 Next, we aggregate into one record per serial number with a total life
-time the first insert time, the last remove time, and a number of
+time, the first insert time, the last remove time, and a number of
 other quantities such as location where the longest time is spent, the
@@
 
-The third rework cycle was used to label GPUs as old batch and new
-batch. The two batches are clearly identifiable in the GPU view. More
+The third rework cycle was used to label GPUs as {\em old} batch and
+{\em new} batch. The two batches are clearly identifiable in the GPU
+view of Fig.~\ref{fig:gpuview} as starting near 2017 timeframe. More
 frequent OTB and DBE events are apparent in the old batch. It is also

### 2020-04-22 dataprep.tex
next

@@
 \label{section:dataprep}
-Data on GPU life times is constructed from two sources: inventory runs
-and failure event records. Two types of failure events were collected:
-Double Bit Error (DBE) and Off the Bus (OTB). Both errors were found
-to be the ``signature'' event of GPU failures and a trigger for their
-replacement.
+Data on Titan GPU life times is constructed from two sources:
+inventory runs and failure event records. Two types of failure events
+were collected: Double Bit Error (DBE) and Off the Bus (OTB). Both
+errors were found to be the ``signature'' event of GPU failures and a
+trigger for their replacement.
 
@@
 each time the system boots. This typically occurs once every few days
-but can be as frequent as a few times per day. An inventory at boot
-time typically occurs over a span of \fix{how many ??} minutes. A
-separate file is produced at each boot time.
-
-The DBE and OTB events are also recorded separately from boot
-time inventories. This was recorded as \fix{ . . . } files.
-
-Inventories are taken at boot time, which can occur anywhere from
-several times per day to more than a month
-apart. Figure~\ref{fig:inventory} shows inter-inventory times that
-appear in the data.
+but can be as frequent as a few times per day. A weekly inventory was
+later added to run independently of system boots when they were fewer
+and farther apart. A single inventory takes about a minute and is
+recorded in a separate file. The DBE and OTB events are also recorded
+separately from the inventories.  Figure~\ref{fig:inventory} shows
+inter-inventory times that appear in the data.
 \begin{figure}[bt]
@@
     unique dates in the data. Note square root scale on the vertical
-    axis.}
+    axis. \fix{George: Add color for another variable? (new-old, date
+      - also doublecheck that these are not DBE/OTB times)??}}
   \label{fig:inventory}
 \end{figure}
-\fix{George: Add color for another variable? (new-old, date)??}
 Notably, there was a period of over a month and many periods of over a
@@
 (SN) and Location of each GPU and creating or updating a separate
-record for each contiguously observed SN-Location combination.
-
-A typical GPU data record is of the form shown in Fig.~\ref{fig:dataraw},
+record for each contiguously observed SN-Location
+combination. Resulting GPU data records are of the form shown in
+Fig.~\ref{fig:dataraw},
 \begin{figure}[tb]
@@
 \end{figure}
-\noindent where records for three GPUs are shown. Each record starts
+where records for three are shown. Each record starts
 with a serial number and locations are coded with {\tt c{\it
-    col-row}c{\it cage}s{\it slot}n{\it node}}. The first GPU record
-shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},
-{\tt c0-1c1s3n3}, with periods off the system, and finally in {\tt
-  c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
-not seen again. The second is intalled in location {\tt c18-4c1s5n1},
-where it stays until the last data collection date on March 3, 2019,
-at 3:19:11 AM. The third GPU is first installed in locations {\tt
-  c5-4c0s7n0}, {\tt c3-3c1s5n0}, {\tt c0-6c1s7n2}, and {\tt
-  c23-6c1s7n2}, where a ``DBE'' is observed on November 2, 2018, and
-it is not seen again.
-
-The location references are cabinet column (0-24), row (0-7), cage
-(0-3), slot (0-7), and node (0-3) with respect to the layout shown in
+    col-row}c{\it cage}s{\it slot}n{\it node}}. The location
+references are cabinet column (0-24), row (0-7), cage (0-3), slot
+(0-7), and node (0-3) with respect to the layout shown in
 Figure~\ref{fig:layout}.
@@
 
+The first GPU record
+shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},
+{\tt c0-1c1s3n3}, with periods off the system, and finally in {\tt
+  c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
+not seen again. The second is intalled in location {\tt c18-4c1s5n1},
+where it stays until the last data collection date on March 3, 2019,
+at 3:19:11 AM. The third GPU is first installed in locations {\tt
+  c5-4c0s7n0}, {\tt c3-3c1s5n0}, {\tt c0-6c1s7n2}, and {\tt
+  c23-6c1s7n2}, where a ``DBE'' is observed on November 2, 2018, and
+it is not seen again.
+
 % \fix{Where does cool air arrive? Is it just dumped out at the top of

### 2020-04-22 survival.tex
done for first submit

@@
 \renewcommand{\pkg}[1]{\textsf{#1}}
-In this section we apply survival analysis methods (time to event
-analysis), which use and combine information across the operational
-lifetimes of all GPUs \cite{survival}. For these analyses, we take
-apart the location string of each GPU into variables {\em col, row,
-  cage, slot, and node} and study the influence of the locations on
-the GPU life times. The construction of a GPU life time is more
-complex than it initially appears because the units are observed only
-at reboot time, because most units were proactively replaced to
-prevent failure, and because some units continue in operation after
-OTB and DBE events (when a second reboot may be successful).
+Survival analysis (SA) methods, sometimes referred as time to event
+methods, focus on the effect of the machine on the GPU units. This is
+useful for determining factors that contribute to failures and
+potentially for root cause analysis. SA methods use and combine
+information across the operational lifetimes of all GPUs
+\cite{survival}. For these analyses, we take apart the location string
+of each GPU into variables {\em col, row, cage, slot, and node} and
+study the influence of the locations on the GPU life times. The
+construction of a GPU life time is more complex than it initially
+appears because the units are observed only at reboot time, because
+most units were proactively replaced to prevent failure, and because
+some units continue in operation after OTB and DBE events (when a
+second reboot may be successful).
 

### 2020-04-22 tbf.tex
finished cox
major updates: elaborate captions, expanded discussion of fig-7, updated discussion of updated fig-8

@@
   \end{center}
-  \caption{Distribution of MTBFs due to DBEs and OTBs across all GPUs.}
+  \caption{Distribution of device-level MTBFs due to DBE and OTB failures across all GPUs during the
+entire lifetime of the machine. DBE and OTB events are separated for each GPU to distinguish between 
+the failure pattern of each event type. Data includes both old and new batch of GPUs with the new ones 
+mostly having MTBFs around or less than 3 years.}
   \label{fig:Device_MTBFs}
@@
 failures. It can be noticed that MTBFs due to DBE and OTB failures of most GPUs are clustered around 
-25,000 hours or 2.8 years. Apart from the center cluster, a significant portion of GPUs has a very 
-low MTBF (less than 1.5 hours) due to both DBE and OTB failures. This shows that a device 
-can see a failure immediately after it is put into service or multiple events are caused due 
-to a single root cause. On the other end of the spectrum, a noticeable portion of GPUs have very high
-MTBFs due to DBE failures. This indicates that most GPUs see a single DBE during their lifetime.
+25,000 hours or 2.8 years. This corresponds to the lifetime of most GPUs in the system after accounting 
+for relocations and replacements done in the machine. The new batch of GPUs almost all of them lie below 
+the average since most of them have a lifetime close to 3 years. It must be pointed out that the new batch did 
+see a significantly less number of failures than the old batch as highlighted in the results presented later on.
+
+Apart from the center cluster, a significant portion of GPUs have a very 
+low MTBF (greater than zero and less than 1.5 hours) due to both DBE and OTB failures. 
+This failure characteristic of the GPUs highlights a period when experiments 
+were being done with putting the GPUs out of service on seeing a failure and moving them to new locations 
+in the machine to possibly mitigate further failures. However, it turned out that the failures re-appeared 
+even with putting the GPU in a new location. In some cases, the failures happened immediately after it is 
+put into service at a different location. The use of insert time as a reference for first failure on a relocated
+GPU to discount the out of service time and in some cases multiple back-to-back failures relatively immediately afterward,
+cause a lot of GPUs to appear in the first bin of the histogram. These experiments continued until machine operators
+figured out the root cause of the GPU failures as discussed earlier.
+
+On the other end of the spectrum, a noticeable portion of GPUs have very high
+MTBFs due to DBE failures. This indicates the population of the GPUs that see a continuous operation in the machine 
+despite its various episodes and see a single DBE event during their lifetime.
 Apart from this behavior, the distribution of MTBF due to DBE events looks very similar to that due 
-to OTB events. Overall, it can be noted from Figure~\ref{fig:Device_MTBFs} that the number of recorded 
-DBE events is significantly higher than the OTB events.
+to OTB events. It can also be noted from Figure~\ref{fig:Device_MTBFs} that the number of recorded 
+DBE events is significantly higher than the OTB events. This highlights the targeted replacement of GPUs 
+with OTB events causing a substantial decrease in OTB events.
 
@@
   \end{center}
-  \caption{Variation of system-wide MTBF over the lifetime of the machine (time is segmented into quarters, 
-i.e., January through March is Q1, so on).}
+  \caption{Variation of system-wide MTBF over the lifetime of the machine. System MTBF is calculated over 
+three month periods to understand the various episodes of the machine. The three month periods are 
+referred to as quarters, i.e., January through March is Q1, and so on. Most GPU replacements started taking 
+place at the end of 2016-Q4 and were completed before 2018-Q1. DBE and OTB events are 
+considered independently as well as together to represent failures occurring on the machine.}
   \label{fig:MTBF_sys}
-\end{figure}
-
-\begin{figure}[bt]
-  \begin{center}
-    \includegraphics[width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL.pdf}
-  \end{center}
-  \caption{Variation of system-wide MTBF across hypothetical new and old partitions of the machine over time.}
-  \label{fig:MTBF_sys_NewOld}
 \end{figure}
@@
 The observation of the MTBF trend shows the different phases which the machine went through. 
-For example, when considering failure due to both DBE and OTB events more than 4X drop in MTBF is 
-noted from 2015-Q3 to 2015-Q4 (note, MTBF of earlier quarters was significantly higher and is not 
-included in the figure to better visualize various change points). This period marked the start of 
-a rapid decline in system-wide MTBF until 2016-Q4. The rapid rise in the number of failures before this 
-quarter can be seen in Figure~\ref{fig:NumFails_sys}. When GPU replacements start to take place in late 
-2016, it triggers an increase in MTBF. However, this change only lasts until 2018-Q1, when we see another 
+Relatively higher MTBFs are noted before 2015-Q4, with some quarters not having any OTB events 
+and overall MTBF being determined solely by DBE events. During this phase, MTBF variation is seen
+from quarter to quarter, but generally, system MTBF remains higher than a day (33 hours) and even reaches as high
+as 10 days in one instance. An alarming drop in system MTBF to less than a day, i.e., 7.7 hours, is observed in 2015-Q4.
+A consistent drop in system-wide MTBF is observed starting from 2015-Q2 until 2016-Q4. 
+The corresponding increase in the number of failures during this phase can be seen in Figure~\ref{fig:NumFails_sys},
+with the peak obtained in 2016-Q4. The usability of the machine comes into question with such low MTBFs, which triggered 
+a phase of installing new GPUs in the machine.
+
+When GPU replacements start to take place in late 
+2016, it triggers a slight increase in MTBF. However, this change only lasts until 2018-Q1, when we see another 
 downward trend of system MTBF. Incidentally, 2018-Q1 also marks the completion of all GPU replacements. 
@@
 be said that overall there has been an increase in MTBF of the machine due to the replacements. 
-For example, before 2017-Q1, a MTBF as low as 2.7 hours is noted, whereas the lowest MTBF is 5.9 hours after this 
-period. With such huge variations in system MTBF, it is difficult to reliably run applications even while using 
-failure recovery approaches such as checkpoint restart, as discussed later on in Section~\ref{section:discussion}.
+For example, the lowest MTBF obtained in 2016-Q4 is 2.7 hours, with the subsequent periods having the 
+lowest MTBF of 5.9 hours. With system MTBF less than a day, even slight variations in MTBF make it 
+difficult to reliably run applications despite using failure recovery approaches such as checkpoint restart, 
+as discussed later on in Section~\ref{section:discussion}.
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/NumFailures_Quarterly_newOld.pdf}
+  \end{center}
+  \caption{The number of DBE and OTB failures observed over the lifetime of the machine.
+A distinction is made for GPU failures happening across old GPUs to highlight the number
+of failures across the newer GPUs. The peak failures are seen in 2016-Q4 (813 failures), 
+which marked the commencement of major replacement of GPUs in the machine.}
+  \label{fig:NumFails_sys}
+\end{figure}
 
@@
 older GPUs in the system.
+
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL.pdf}
+  \end{center}
+  \caption{Variation of system-wide MTBF across hypothetical new and old partitions of the machine over time.
+Results are presented from 2017-Q1 which marked the beginning of when a substantial number of new GPUs are in service.
+DBE and OTB events are both considered as a failure for calculating the MTBF.
+Results highlight the difference in reliability observed by jobs using the new GPUs as compared to old GPUs,
+as well as the minute difference in MTBF measured across the old partition and the whole machine.}
+  \label{fig:MTBF_sys_NewOld}
+\end{figure}
 
@@
 The next section quantifies the survivability of this machine with its varied MTBF characteristics beyond its operational lifetime.
-\fix{George, can you please check the last sentence, where I make a connection with the next section.}
-
-\begin{figure}[bt]
-  \begin{center}
-    \includegraphics[width=\columnwidth]{figs/NumFailures_Quarterly_newOld.pdf}
-  \end{center}
-  \caption{The number of DBE and OTB failures observed over the lifetime of the machine.
-A distinction is made for GPU failures happening across old GPUs to highlight the number 
-of failures across the newer GPUs.}
-  \label{fig:NumFails_sys}
-\end{figure}
+%\fix{George, can you please check the last sentence, where I make a connection with the next section.}

### 2020-04-23 intro.tex
intro fig

@@
 involving data prior to 2014. All analyses and figures that follow use
-only data from 2014 onward. We make the data as well as our analysis
-codes to reproduce the graphics available in a public repository (see,
-\cite[repository]{repositoryreference}).  While the data is only a 7
-MB file, due to the size of Titan it represents over 100,000 years of
-GPU operation and may be the largest publicly available GPU
-reliability data set. Our hope is that others will use it for
-reliability and survival analysis research and go beyond the results
-presented here.
+only data from 2014 onward. We make the data as well as our analysis R
+codes and Python codes to reproduce the graphics in this paper
+publicly available (see, \cite[repository to be added in final
+version]{repositoryreference,rcodes,pycodes}).  While the data is only
+a 7 MB file, due to the size of Titan it represents over 100,000
+collective years of GPU operation and may be the largest publicly
+available GPU reliability data set. Our hope is that others will use
+it for reliability and survival analysis research and go beyond the
+results presented here.
 
@@
 relative hazard rates in various locations are in
-Sec.~\ref{section:survival}. A high-level difference between a TBF
-analysis and survival analysis (SA) is that TBF studies largely how
-GPU reliability affects the machine, wheras SA studies largely how the
+Sec.~\ref{section:survival}. The goals of a TBF analysis and of
+survival analysis (SA) are different: TBF studies largely how GPU
+reliability affects the machine, wheras SA studies largely how the
 machine affects GPU reliability.  Finally, we discuss our main

### 2020-04-23 related.tex
new figs

--- 2020-04-20 related.tex
@@
 
-Prior work in investigating Titan's reliability includes a significant amount of work that covers different aspects.
+Being the largest machine with the most GPUs in the world attracts a lot of attention so particularly during Titan's first few years of operation, many studies were published about different aspects of its reliability.
+However, none of the studies investigate the atypical failure mode discussed in this paper nor do they consider data over the full lifetime of the system through its decomissioning.
 
-Tiwari et al~\cite{Tiwari15Experience,7056044} analyzed the types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
-It also used neutron beam experiments at the Los Alamos Neutron Science Center and at Rutherford Appleton Laboratories to further understand GPU soft errors caused by neutron radiation triggered upsets (bit flips and timing errors).
+Tiwari et al~\cite{Tiwari15Experience,7056044} analyze types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
+This also includes neutron beam experiments at the Los Alamos Neutron Science Center and at Rutherford Appleton Laboratories to further understand GPU soft errors caused by neutron radiation triggered upsets (bit flips and timing errors).
 %
-Tiwari et al~\cite{10.1145/2807591.2807666} further analyzed the types of GPU errors on Titan, including software/firmware related errors and failures, ECC double-bit errors, and GPU ``off the bus'' and ECC page retirement events.
-This work observed a ECC double-bit mean-time between errors (MTBE) of \~160 hours, or \~ one per week, with 86\% in GPU memory and the rest in the GPU register file.
-The experienced GPU ``off the bus'' events were caused by a system integration issue that was fixed and not an inherent GPU or GPU memory architecture flaw.
+Tiwari et al~\cite{10.1145/2807591.2807666} further analyze types of GPU errors on Titan, including software/firmware related errors and failures, ECC double-bit errors, and GPU ``off the bus'' and ECC page retirement events.
+This work observes an ECC double-bit mean-time between errors (MTBE) of about 160 hours, or about one per week, with 86\% occurring in GPU memory and the rest in the GPU register file.
+The GPU ``off the bus'' events were caused by a system integration issue that was fixed and not an inherent GPU or GPU memory architecture flaw.
 %
-Nie et al.~\cite{7446091,nie17characterizing,nie18machine} characterized the soft error behavior of Titan's GPUs in relation to temperature and power consumption to predict their increased occurrence by correlating data in temporal and spatial domains and using machine learning models.
-The work primarily focused on correctable single-bit errors.
+Nie et al.~\cite{7446091,nie17characterizing,nie18machine} characterize the soft error behavior of Titan's GPUs in relation to temperature and power consumption to predict their increased occurrence by correlating data in temporal and spatial domains and using machine learning models.
+The work primarily focuses on correctable single-bit errors.
 %
-Tiwari et al. and Nie et al. did not address the atypical failure mode discussed in this paper and only used Titan error and failure data from June 2013 to February 2015.
+Tiwari et al.
+and Nie et al.
+d0 not address the atypical failure mode discussed in this paper.
+They also only use Titan error and failure data from June 2013 to February 2015, excluding early failures and the failure mode discussed in this paper (which occurred after February 2015).
 
-Zimmer et al.~\cite{8665764} developed a new job scheduling strategy for Titan to counter the GPU failures discussed in this paper and to improve system utilization and productivity.
-The solution used reordering of compute nodes for resource allocation, scheduling larger jobs on more reliable and smaller jobs on less reliable nodes.
+Zimmer et al.~\cite{8665764} develop a new job scheduling strategy for Titan to counter the GPU failures that we discuss and to improve system utilization and productivity.
+The solution uses reordering of the compute nodes for resource allocation, scheduling larger jobs on more reliable and smaller jobs on less reliable nodes.
 
-Ezell~\cite{osti_1086655} created a general understanding of Titan's interconnect failures using an application to stress test Titan's Gemini interconnect.
+Ezell~\cite{osti_1086655} creates a general understanding of Titan's interconnect failures using an application to stress test Titan's Gemini interconnect.
 %
-Work by Kumar et al.~\cite{kumar18understanding} analyzed Titan's interconnect faults, errors and congestion events to improve interconnects resilience and congestion resolution methods.
-The results show that the magnitude of interconnect errors is very high with an uneven distribution across different types of links is uneven.
-They also show that congestion is highly frequent and bursty.
+The work by Kumar et al.~\cite{kumar18understanding} analyzes Titan's interconnect faults, errors and congestion events to improve resilience of the interconnects and congestion resolution methods.
+The results show that the magnitude of interconnect errors is very high with an uneven distribution across different types of links.
+They also show that congestion is very frequent and bursty.
 %
-Gupta et al.~\cite{7266836} investigated the spatial and temporal properties of failures on Titan and their impact on resilience mechanisms and the implications for efficient system operation.
+Gupta et al.~\cite{7266836} investigate the spatial and temporal properties of failures on Titan and their impact on resilience mechanisms and the implications for efficient system operation.
 %
-Gupta et al.~\cite{gupta17failures} also performed a study covering five supercomputers at Oak Ridge National Laboratory, including Titan, that concentrated on developing an understanding of which and how many errors and failure occur in supercomputers over multiple generations and over their years of operation.
+Gupta et al.~\cite{gupta17failures} also perform a study covering five supercomputers at Oak Ridge National Laboratory, including Titan, that concentrates on developing an understanding of which and how many errors and failures occur in supercomputers over multiple generations and over their years of operation.
 This work resulted in a lot of lessons learned, including that the mean-time between failures (MTBF) can change drastically and non-monotonically over time.
 %
-Meneses et al.~\cite{Meneses15Analyzing} analyzed the interplay and workload on Titan using failure and job submission logs. The results indicate that failures depend heavily on workload.
+Meneses et al.~\cite{Meneses15Analyzing} analyze the interplay and workload on Titan using failure and job submission logs. The results indicate that failures depend heavily on workload.
 
-Bautista-Gomez et al.~\cite{bautista-gomez16reducing} used failure data from Titan to adapt the checkpoint frequency to the current reliability of the system using dynamic checkpointing.
+Bautista-Gomez et al.~\cite{bautista-gomez16reducing} use failure data from Titan to adapt the checkpoint frequency to the current reliability of the system using dynamic checkpointing.
 %
-Tiwari et al.~\cite{6903564} also used failure data from Titan and exploited temporal locality in the data to adapt the checkpoint frequency to the current reliability of the system using lazy checkpointing.
+Tiwari et al.~\cite{6903564} also use failure data from Titan and exploit temporal locality in the data to adapt the checkpoint frequency to the current reliability of the system using lazy checkpointing.
 %
@@
 
-Other work in characterizing supercomputer faults, errors and failures focused on the various systems deployed around the world, primarily at US Department of Energy laboratories.
+Other work in characterizing supercomputer faults, errors and failures focuses on the various other systems deployed around the world, primarily at US Department of Energy laboratories.
 %
-Di et al.~\cite{8809553} developed an in-depth understanding of the failure characteristics of the IBM Blue Gene/Q Mira system at Argonne National Laboratory.
-This work showed that 99.4\% of job failures were due to user behavior, such as errors in the code or misconfiguration.
-Martino et al.~\cite{6903615} characterized the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications (NCSA), at the University of Illinois at Urbana-Champaign.
-The results showed that 74.4\% of system-wide outages in Blue Waters were caused by software.
+Di et al.~\cite{8809553} develop an in-depth understanding of failure characteristics of the IBM Blue Gene/Q Mira system at Argonne National Laboratory.
+This work shows that 99.4\% of job failures are due to user behavior, such as errors in the code or misconfiguration.
+Martino et al.~\cite{6903615} characterize the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications (NCSA), at the University of Illinois at Urbana-Champaign.
+The results show that 74.4\% of system-wide outages in Blue Waters were caused by software.
 Blue Waters's XK7 partition had the same architecture as Titan, but did not experience the same GPU failure mode detailed in this paper.
+% Interesting ... do we know why?

### 2020-04-23 dataprep.tex
next

@@
   \caption{Inter-inventory times, taken as time intervals between
-    unique dates in the data. Note square root scale on the vertical
-    axis. \fix{George: Add color for another variable? (new-old, date
-      - also doublecheck that these are not DBE/OTB times)??}}
+    unique dates in the data and removing those less than a day. Note
+    square root scale on the vertical axis.}
   \label{fig:inventory}
@@
 Notably, there was a period of over a month and many periods of over a
-week that Titan operated without a single reboot.
+week that Titan operated without a single reboot, especially in the
+2014-2015 time frame.
 

### 2020-04-23 tbf.tex
next

@@
 
-Using the processed data, we analyze the inter-arrival times between the DBE and OTB events. 
-This analysis is done at the device level and the system level. 
+Using the cleaned data, we analyze the inter-arrival times between the DBE and OTB events. 
+This analysis is done at the device level and at the system level. 
 It provides important insights into the reliability of large-scale machines, where the 
@@
 as opposed to those provided in the device datasheet. The failures are tracked using the
-SN of the GPUs even though a GPU might have been placed at different locations in the 
+{\em sn} of the GPUs even though a GPU might have been placed at different locations in the 
 machine during its lifetime. The time to the first failure on a device is measured by taking
@@
 25,000 hours or 2.8 years. This corresponds to the lifetime of most GPUs in the system after accounting 
-for relocations and replacements done in the machine. The new batch of GPUs almost all of them lie below 
-the average since most of them have a lifetime close to 3 years. It must be pointed out that the new batch did 
-see a significantly less number of failures than the old batch as highlighted in the results presented later on.
+for relocations and replacements done in the machine. Almost all of the new batch of GPUs lie below 
+the average since most have a lifetime close to 3 years. We point out
+that the new batch had a much smaller number of failures than
+the old batch as we highlight in the next section.
 
-Apart from the center cluster, a significant portion of GPUs have a very 
-low MTBF (greater than zero and less than 1.5 hours) due to both DBE and OTB failures. 
-This failure characteristic of the GPUs highlights a period when experiments 
-were being done with putting the GPUs out of service on seeing a failure and moving them to new locations 
-in the machine to possibly mitigate further failures. However, it turned out that the failures re-appeared 
-even with putting the GPU in a new location. In some cases, the failures happened immediately after it is 
-put into service at a different location. The use of insert time as a reference for first failure on a relocated
-GPU to discount the out of service time and in some cases multiple back-to-back failures relatively immediately afterward,
-cause a lot of GPUs to appear in the first bin of the histogram. These experiments continued until machine operators
-figured out the root cause of the GPU failures as discussed earlier.
+Apart from the center cluster, a considerable portion of GPUs also have a very 
+low MTBF. This failure characteristic is mainly an artifact from
+troubleshooting a period of increasing failures in mid to late 2016.
+After a GPU experienced an OTB or a DBE a second reboot was attempted at
+the same or possibly different location. Usually, but not always, a
+second OTB or DBE was generated immediately or possibly after a short
+time. When moved to a new location, this generates an artifact in our
+data with an unusually short MTBF because we use the {\em insert} time as
+a reference for the first failure on a relocated GPU. This drawback is
+balanced by the advantage that this way we can discount out of service times
+which were frequent particularly during early 2017 as can be seen in
+Fig.~\ref{fig:sample_loc}.
 
 On the other end of the spectrum, a noticeable portion of GPUs have very high
-MTBFs due to DBE failures. This indicates the population of the GPUs that see a continuous operation in the machine 
+MTBFs and their failures are mostly DBEs. This is the population of the GPUs that see a continuous operation in the machine 
 despite its various episodes and see a single DBE event during their lifetime.
-Apart from this behavior, the distribution of MTBF due to DBE events looks very similar to that due 
+Apart from this, the distribution of MTBF due to DBE events looks very similar to that due 
 to OTB events. It can also be noted from Figure~\ref{fig:Device_MTBFs} that the number of recorded 
@@
 
-Although individual device MTBFs provide important insights into the varying failure rates
-observed across the machine. It is worthwhile to perform a system-wide MTBF analysis since 
-most high-performance computing applications use a significant number of GPUs in parallel. 
+Now we turn to a system-wide MTBF analysis. Most high-performance
+computing applications, especially on leadership computing
+systems such as Titan, use a large fraction of the entire machine in parallel.  
 In the following analysis, failures occurring across all GPUs are consolidated and time between failures
 is calculated. The old and new GPUs are all considered. At any given time, only a fixed number of GPUs
-are in operation so any variation in time-between-failures is due to individual device reliabilities.
+is in the system so any variation in time-between-failures is due to individual device reliabilities.
 
@@
 With many GPU relocations and replacements across the machine over its lifetime, it is 
-best to consider the variation of MTBF across fixed periods. Herein, we calculate 
-the mean of time-between-failures across three months. Figure~\ref{fig:MTBF_sys}
-clearly shows the considerable change in system-wide MTBF from one period to another. 
-The observation of the MTBF trend shows the different phases which the machine went through. 
-Relatively higher MTBFs are noted before 2015-Q4, with some quarters not having any OTB events 
-and overall MTBF being determined solely by DBE events. During this phase, MTBF variation is seen
-from quarter to quarter, but generally, system MTBF remains higher than a day (33 hours) and even reaches as high
-as 10 days in one instance. An alarming drop in system MTBF to less than a day, i.e., 7.7 hours, is observed in 2015-Q4.
+best to consider the variability of MTBF across fixed periods. Here, we calculate 
+the mean of quarterly (three months) time-between-failures. Figure~\ref{fig:MTBF_sys}
+shows the considerable change in system-wide MTBF from one period to another. 
+We can see that MTBF tells the different phases that occurred on the machine. 
+Relatively high MTBFs are seen before 2015-Q4, with some quarters not having any OTB events 
+and overall MTBF being determined solely by DBE events. During this
+phase, we see some MTBF variation from quarter to quarter, but
+generally, the system MTBF remains higher than one day (33 hours) and reaches as high
+as 10 days in one instance. An alarming drop in the system MTBF to less than a day, (7.7 hours), is observed in 2015-Q4.
 A consistent drop in system-wide MTBF is observed starting from 2015-Q2 until 2016-Q4. 
 The corresponding increase in the number of failures during this phase can be seen in Figure~\ref{fig:NumFails_sys},
-with the peak obtained in 2016-Q4. The usability of the machine comes into question with such low MTBFs, which triggered 
+with a peak in 2016-Q4. The usability of the machine comes into
+question with such low MTBFs and it eventually triggered 
 a phase of installing new GPUs in the machine.
@@
 downward trend of system MTBF. Incidentally, 2018-Q1 also marks the completion of all GPU replacements. 
-So the upward trend noted in the period from 2016-Q4 to 2018-Q1 might be attributed to a portion of the machine 
-being unavailable while the reworks were taking place. A single replacement cycle lasted multiple weeks. 
+So the upward trend noted in the period from 2016-Q4 to 2018-Q1 is
+likely due to phased replacements, each time a portion of the machine 
+being unavailable, thus having a smaller number of GPUs than the full
+machine in operation. 
 %\fix{need to verify how long the replacement took}
-There is no definite way to incorporate this unavailability of the machine into MTBF analysis, so it has to
-be said that overall there has been an increase in MTBF of the machine due to the replacements. 
+There is not definitive way to incorporate this unavailability of the
+machine into MTBF analysis, unless we know how the down times were scheduled.
 For example, the lowest MTBF obtained in 2016-Q4 is 2.7 hours, with the subsequent periods having the 
@@
   \caption{The number of DBE and OTB failures observed over the lifetime of the machine.
-A distinction is made for GPU failures happening across old GPUs to highlight the number
-of failures across the newer GPUs. The peak failures are seen in 2016-Q4 (813 failures), 
-which marked the commencement of major replacement of GPUs in the machine.}
+A distinction is made between failures on old GPUs to highlight the number
+of failures on the newer GPUs. The peak failures are seen in 2016-Q4 (813 failures), 
+which marks the commencement of major replacement of GPUs in the machine.}
   \label{fig:NumFails_sys}
@@
 
-The mean-time-between DBE and OTB failures are also separately tracked in this analysis.
-One observation is the drastic increase in mean-time-between OTB failures after the GPU replacements.
-This is also evident from the reduction of the absolute number of OTB failures in Figure~\ref{fig:NumFails_sys}.
-However, the system MTBF is determined by the weakest link and the occurrence of DBE events tends to dictate it.
-Even though the replacements helped to increase the MTBF due to DBE events by a factor, the overall system 
-reliability is dictated by the components with the most age in the system. There is a re-emergence of upward trend towards 
-the end of life in the number of DBE failures in Figure~\ref{fig:NumFails_sys} which almost all of them are due to 
-older GPUs in the system.
+We also separate the DBE and OTB failures in this analysis. After the
+2017 replacement of many GPUs, there is a drastic increase in
+mean-time-between OTB failures, which is also evident in a reduction of OTB failures in Figure~\ref{fig:NumFails_sys}.
+However, the system MTBF is determined by the weakest link and here the occurrence of DBE events tends to dictate it.
+Even though the replacements helped to increase the MTBF due to DBE events, the overall system 
+reliability is dictated by the components with the most age in the
+system. There is a re-emergence of an upward trend towards 
+the end in DBE failures in Figure~\ref{fig:NumFails_sys}, which
+appears to be due to older GPUs in the system.
 
@@
 partitions of the machine starting from the time when the majority of GPU replacements have been completed. 
-The new partition of the machine has always 12X better MTBF than the older partition
-of the machine. For example, in 2018-Q4, the MTBF recorded across old partition is about 7.9 hours, whereas it is about
-96 hours (4 days) for the newer partition. The implications of such a huge disparity on applications running on the 
-system are discussed in Section~\ref{section:discussion}. The insignificant difference in MTBF measured across all GPUs and
-old partition of the machine can also be noted in Figure~\ref{fig:MTBF_sys_NewOld}.  
-Therefore, the old GPUs that occupy a minority portion of the machine after GPU replacements, 
-and have a disproportionately high number of failures and contribute towards a low system MTBF play a major role 
-in determining the overall reliability of the machine.
-The next section quantifies the survivability of this machine with its varied MTBF characteristics beyond its operational lifetime.
-%\fix{George, can you please check the last sentence, where I make a connection with the next section.}
+The new partition of the machine has a 12X better MTBF than the older
+partition.
+
+For example, in 2018-Q4, the old partition MTBF is about 7.9 hours,
+whereas the newer partitio is 96 hours (4 days). The implications of such a huge disparity on applications running on the 
+system are discussed in
+Section~\ref{section:discussion}. Moreover, while the old partition is
+smaller than the new partition, the old partition drives the overall reliability of
+the machine.

### 2020-04-23 survival.tex
s

@@
 \label{section:survival}
-\renewcommand{\pkg}[1]{\textsf{#1}}
-Survival analysis (SA) methods, sometimes referred as time to event
-methods, focus on the effect of the machine on the GPU units. This is
-useful for determining factors that contribute to failures and
-potentially for root cause analysis. SA methods use and combine
-information across the operational lifetimes of all GPUs
-\cite{survival}. For these analyses, we take apart the location string
-of each GPU into variables {\em col, row, cage, slot, and node} and
-study the influence of the locations on the GPU life times. The
-construction of a GPU life time is more complex than it initially
-appears because the units are observed only at reboot time, because
-most units were proactively replaced to prevent failure, and because
-some units continue in operation after OTB and DBE events (when a
-second reboot may be successful).
+\renewcommand{\pkg}[1]{\textsf{#1}} Survival analysis (SA) methods,
+sometimes referred as time to event methods, focus on the effect of
+the machine on the GPU units. These methods have roots in biomedical
+statistics, where the effect on the patient is most important. So here
+we treat the GPUs as patients to determine factors that contribute to
+failures. SA methods use and combine information across the
+operational lifetimes of all GPUs. For these analyses,
+we take apart the location string of each GPU into variables {\em col,
+  row, cage, slot, and node} and study the influence of the locations
+on the GPU life times. The construction of a GPU life time is more
+complex than it initially appears because the units are observed only
+at reboot time, because most units were proactively replaced to
+prevent failure, and because some units continue in operation after
+OTB and DBE events (when a second reboot may be successful).
 
@@
 life spans were recorded only at inventory times. We use censoring
-concepts on the proactive replacements and units still in operation at
+concepts on the proactive replacements and on units still in operation at
 the end. This allows us to use all of the GPU life time data,
@@
   \includegraphics[width=\columnwidth]{figs/cox_o001.pdf}
-  \caption{GPU hazard ratios from Cox regression model on {\tt old}
+  \caption{GPU hazard ratios from Cox regression model on {\em old}
     batch.}
@@
   \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
-  \caption{GPU hazard ratios from Cox regression model on {\tt new}
+  \caption{GPU hazard ratios from Cox regression model on {\em new}
     batch.}

### 2020-04-23 tbf.tex
last?

@@
   \begin{center}
-    \includegraphics[width=\columnwidth]{figs/MTBF_GPUwise.pdf}
+    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_GPUwise.pdf}
   \end{center}
@@
 
-Apart from the center cluster, a considerable portion of GPUs also have a very 
+Apart from the center cluster, many GPUs also have a very 
 low MTBF. This failure characteristic is mainly an artifact from
@@
 which were frequent particularly during early 2017 as can be seen in
-Fig.~\ref{fig:sample_loc}.
+Fig.~\ref{fig:locview}.
 
 On the other end of the spectrum, a noticeable portion of GPUs have very high
-MTBFs and their failures are mostly DBEs. This is the population of the GPUs that see a continuous operation in the machine 
+MTBFs and their failures are mostly DBEs. These are the GPUs that see continuous operation in the machine 
 despite its various episodes and see a single DBE event during their lifetime.
 Apart from this, the distribution of MTBF due to DBE events looks very similar to that due 
-to OTB events. It can also be noted from Figure~\ref{fig:Device_MTBFs} that the number of recorded 
-DBE events is significantly higher than the OTB events. This highlights the targeted replacement of GPUs 
+to OTB events. We also note in Figure~\ref{fig:Device_MTBFs} that the number of recorded 
+DBE events is much higher than the OTB events. This highlights the targeted replacement of GPUs 
 with OTB events causing a substantial decrease in OTB events.
@@
   \begin{center}
-    \includegraphics[width=\columnwidth]{figs/MTBF_quaterly_sys.pdf}
+    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_quaterly_sys.pdf}
   \end{center}
@@
 machine into MTBF analysis, unless we know how the down times were scheduled.
-For example, the lowest MTBF obtained in 2016-Q4 is 2.7 hours, with the subsequent periods having the 
+The lowest MTBF obtained in 2016-Q4 is 2.7 hours, with subsequent periods having the 
 lowest MTBF of 5.9 hours. With system MTBF less than a day, even slight variations in MTBF make it 
@@
   \begin{center}
-    \includegraphics[width=\columnwidth]{figs/NumFailures_Quarterly_newOld.pdf}
+    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/NumFailures_Quarterly_newOld.pdf}
   \end{center}
@@
 
+\begin{figure}[bt]
+  \begin{center}
+    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL.pdf}
+  \end{center}
+  \caption{Variation of system-wide MTBF across hypothetical new and
+    old partitions of the machine after a substantial number of new GPUs
+were put in service.
+DBE and OTB events both determine the MTBF.
+Results highlight the difference in reliability for jobs using the new GPUs as compared to old GPUs,
+as well as now MTBF is driven by the weaker old partition.}
+  \label{fig:MTBF_sys_NewOld}
+\end{figure}
 We also separate the DBE and OTB failures in this analysis. After the
@@
 appears to be due to older GPUs in the system.
-
-\begin{figure}[bt]
-  \begin{center}
-    \includegraphics[width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL.pdf}
-  \end{center}
-  \caption{Variation of system-wide MTBF across hypothetical new and old partitions of the machine over time.
-Results are presented from 2017-Q1 which marked the beginning of when a substantial number of new GPUs are in service.
-DBE and OTB events are both considered as a failure for calculating the MTBF.
-Results highlight the difference in reliability observed by jobs using the new GPUs as compared to old GPUs,
-as well as the minute difference in MTBF measured across the old partition and the whole machine.}
-  \label{fig:MTBF_sys_NewOld}
-\end{figure}
 

### 2020-04-23 dataclean.tex
final tweaks

@@
   \label{fig:gpuview}
+\end{figure*}
+\begin{figure*}[tb]
+  \includegraphics[width=6.5in]{figs/sample_loc.pdf}
+  \caption{Location view of GPU life and failures.  Legend: black dots
+    are installs, black lines are lifetimes at installed location,
+    blue squares are OTB events, red triangles are DBE events, and
+    black ] are ``last seen'' events.}
+  \label{fig:locview}
 \end{figure*}
@@
 To get some intuition for the GPU lifetimes on Titan, we give two
-views of 80 randomly selected GPUs (Figure~\ref{fig:gpuview}) and 80
+views of 60 randomly selected GPUs (Figure~\ref{fig:gpuview}) and 80
 randomly selected locations (Figure~\ref{fig:locview}).
-\begin{figure*}[tb]
-  \includegraphics[width=6.5in]{figs/sample_loc.pdf}
-  \caption{Location view of GPU life and failures.  Legend: black dots
-    are installs, black lines are lifetimes at installed location,
-    blue squares are OTB events, red triangles are DBE events, and
-    black ] are ``last seen'' events.}
-  \label{fig:locview}
-\end{figure*}
 The GPU view

### 2020-04-23 discussion.tex
Fixed related work and discussion section.
final tweaks

--- 2020-04-21 discussion.tex
@@
 \label{section:discussion}
-
-\fix{Possibly a short summary of the observations in this paper to facilitate the discussion below.}
 
@@
 The options of replacing failed components and employing reliability-aware
-resource management may not be available or cost effective for other
+resource management may not be available or be cost effective for other
 supercomputing centers dealing with similar unexpected reliability issues.
@@
 Replacement components may not be readily available and may have to be
-manufactured, which can be impossible for technology that is no longer supported
+manufactured, which can be impossible for a technology that is no longer supported
 by the original manufacturer. Experienced errors or failures may be caused by
@@
 
-There is also a financial aspect, such as service contract and/or warranty
+There is also a financial aspect, such as a service contract and/or warranty
 expiration and the operational budget of a supercomputing center not having
@@
 There is additionally the time and financial cost for root-cause analysis and
-developing a hardware and/or software mitigation strategy in the first place.
+for developing a hardware and/or software mitigation strategy in the first place.
 Fixing a problem requires finding and understanding it, which can be difficult 
@@
 the same job need to be close to each other on the network for maximum
-application performance.
+application performance. Also, node outages or less reliable nodes within a
+tight group of reliable nodes create resource allocation
+holes. Saparating two lower nodes in each cage 0 on Titan would not
+create a contiguous partition, yet this group would be the most
+reliable in 2016 as was shown in the preceeding section.
 %
@@
 acquisition contracts may need to include performance requirements for degraded
-operation, such as a 60\% performance capability if the MTBF drops by a
-magnitude.
+operation, such as a certain percentage of performance capability if the MTBF
+drops by an order of magnitude.

### 2020-05-24 discussion.tex
mostly typos and two graphs

@@
 tight group of reliable nodes create resource allocation
-holes. Saparating two lower nodes in each cage 0 on Titan would not
+holes. Separating two lower nodes in each cage 0 on Titan would not
 create a contiguous partition, yet this group would be the most

### 2020-05-24 intro.tex
mostly typos and two graphs

@@
 We begin by describing related work in Sec.~\ref{section:related}. Our
-data collection is desctibed in Sec.~\ref{section:dataprep} and the
+data collection is described in Sec.~\ref{section:dataprep} and the
 data cleaning, checking, and lifetime summarization process in
@@
 Sec.~\ref{section:survival}. The goals of a TBF analysis and of
-survival analysis (SA) are different: TBF studies largely how GPU
-reliability affects the machine, wheras SA studies largely how the
+survival analysis (SA) are different: TBF studies largely show GPU
+reliability affects the machine, whereas SA studies largely how the
 machine affects GPU reliability.  Finally, we discuss our main

### 2020-05-24 related.tex
mostly typos and two graphs

@@
 and Nie et al.
-d0 not address the atypical failure mode discussed in this paper.
+do not address the atypical failure mode discussed in this paper.
 They also only use Titan error and failure data from June 2013 to February 2015, excluding early failures and the failure mode discussed in this paper (which occurred after February 2015).
@@
 This work shows that 99.4\% of job failures are due to user behavior, such as errors in the code or misconfiguration.
-Martino et al.~\cite{6903615} characterize the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications (NCSA), at the University of Illinois at Urbana-Champaign.
+Martino et al.~\cite{6903615} characterize the errors and failures of the Blue Waters Cray XE6/XK7 system at the National Center for Supercomputing Applications, at the University of Illinois at Urbana-Champaign.
 The results show that 74.4\% of system-wide outages in Blue Waters were caused by software.

### 2020-05-25 intro.tex
added revews and Makefile entry

@@
   \includegraphics[width=\textwidth]{figs/chronology001.pdf}
-  \caption{Volume of GPU swaps detected on Titan at inventories. Our
-    analysis begins with units in place by 2014. A break-in period
-    before 2014 involved working out mechanical assembly issues at
-    PCIe connectors.}
+  \caption{Volume of GPU swaps detected on Titan at individual
+    inventories (narrow blue) and yealy sum totals for 2014 and later
+    (wide gray). High swap volumes prior to 2014 reflect two major
+    rework cycles on mechanical assembly issues at PCIe connectors
+    that can be characterized as a break-in period. Our analysis in
+    this paper excludes the break-in period and is focused on units in
+    place by 2014 or swapped in during the years that follow.} 
   \label{fig:chronology}

### 2020-05-26 intro.tex
mostly refs

@@
   \includegraphics[width=\textwidth]{figs/chronology001.pdf}
-  \caption{Volume of GPU swaps detected on Titan at individual
+  \caption{A seven-year tale of GPU life on a leadership class
+    supercomputer: Volume of GPU swaps detected on Titan at individual
     inventories (narrow blue) and yealy sum totals for 2014 and later
-    (wide gray). High swap volumes prior to 2014 reflect two major
-    rework cycles on mechanical assembly issues at PCIe connectors
-    that can be characterized as a break-in period. Our analysis in
-    this paper excludes the break-in period and is focused on units in
-    place by 2014 or swapped in during the years that follow.} 
+    (wide gray) over its entire lifetime. High swap volumes prior to
+    2014 reflect two major rework cycles on mechanical assembly issues
+    at PCIe connectors that can be characterized as a break-in
+    period. Our analysis in this paper excludes the break-in period
+    and is focused on units in place by the end of 2013 or swapped in
+    during the years that follow.}
   \label{fig:chronology}

### 2020-05-26 survival.tex
mostly typos and two graphs
mostly refs

@@
 \label{section:survival}
-\renewcommand{\pkg}[1]{\textsf{#1}} Survival analysis (SA) methods,
+\renewcommand{\pkg}[1]{\textsf{#1}} Survival analysis (SA) methods
+\cite{survival-book}, 
 sometimes referred as time to event methods, focus on the effect of
@@
 
-Kaplan-Meier survival analysis (KM) starts with computing the
+Kaplan-Meier survival analysis (KM)
+\cite{kaplan-meier1958,survival-book} starts with computing the
 probability of survival beyond a given time. It is a nonparametric
@@
 It is the probability of being operational at duration $t$.  We use
-the R packages \pkg{survival} and \pkg{survminer} for the KM analysis,
-which is reported in Figure~\ref{fig:km-all-cage-node}. Within each
-{\em batch}, separate survival curves are computed for each {\em cage}
-by {\em node} combination. Along with the survival curve estimate,
-this analysis provides 95\% confidence region for survival probability
+the R packages \pkg{survival} \cite{survival-package} and
+\pkg{survminer} \cite{survminer-package} for the KM analysis, which is
+reported in Figure~\ref{fig:km-all-cage-node}. Within each {\em
+  batch}, separate survival curves are computed for each {\em cage} by
+{\em node} combination. Along with the survival curve estimate, this
+analysis provides 95\% confidence region for survival probability
 shaded around the curves.
@@
 comes mostly from the hazard ratios for the old batch. The new batch
-has mot had many failure events and the uncertainty bars of nearly all
+has not had many failure events and the uncertainty bars of nearly all
 ratios include 1, which is no difference. In the old batch, we see

### 2020-05-26 dataprep.tex
mostly typos and two graphs
cabinet fig

@@
     unique dates in the data and removing those less than a day. Note
-    square root scale on the vertical axis.}
+    square root scale on the vertical axis and missing days for 15,
+    17, 19, and 22-55.}
   \label{fig:inventory}
@@
     col-row}c{\it cage}s{\it slot}n{\it node}}. The location
-references are cabinet column (0-24), row (0-7), cage (0-3), slot
+references are cabinet column (0-24), row (0-7), cage (0-2), slot
 (0-7), and node (0-3) with respect to the layout shown in
@@
   \centering
+  \includegraphics[width=\columnwidth]{figs/TitanCabinet.png}
+  \begin{minipage}{0.35\columnwidth}
+    \includegraphics[width=\columnwidth]{figs/TitanBlade.png}
+  \end{minipage}
+  \hfil
   \begin{minipage}{0.49\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/TitanCabinet.png}
-    \includegraphics[width=0.9\columnwidth]{figs/TitanCabinetSide.png}
-  \end{minipage}
-  \begin{minipage}{0.49\columnwidth}
-    \includegraphics[width=\columnwidth]{figs/TitanBlade.png}
-
-    \vspace{0.8\columnwidth}
     \includegraphics[width=\columnwidth]{figs/TitanLayout.png}
   \end{minipage}
-  \caption{Titan cabinet (front view and side view with air flow) on
-    the left, blade layout and floor layout on the right. Cooling
-    airflow is bottom to top within a cabinet, taking ambient air in
-    at the foot and relasing it back out at the top. Cooling fluid is
-    passed through two heat exchange evaporators, one below Cage 0 and
-    one above Cage 2.}
+  \caption{Titan cabinet (front view and side view with air flow) and
+    single blade layout (bottom left) and floor layout (bottom
+    right). Cooling airflow is bottom to top within a cabinet, taking
+    ambient air in mostly at the foot and relasing it back out at the
+    top. Cooling fluid is passed through two heat exchange
+    evaporators, one below Cage 0 and one above Cage 2.}
   \label{fig:layout}
@@
   c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
-not seen again. The second is intalled in location {\tt c18-4c1s5n1},
+not seen again. The second is installed in location {\tt c18-4c1s5n1},
 where it stays until the last data collection date on March 3, 2019,

### 2020-05-26 dataprep.tex
cabinet fig
reviwer 1 commented

@@
     unique dates in the data and removing those less than a day. Note
-    square root scale on the vertical axis and missing days for 15,
-    17, 19, and 22-55.}
+    square root scale on the vertical axis.}
   \label{fig:inventory}
 \end{figure}
-Notably, there was a period of over a month and many periods of over a
-week that Titan operated without a single reboot, especially in the
-2014-2015 time frame.
+Notably, there was a period of nearly two months and many periods of over a
+week that Titan operated without a single reboot in the
+2014-2015 time frame. By the time the weekly inventories were added,
+failure rates started to increase and lead to more frequent reboots.
 
@@
     ambient air in mostly at the foot and relasing it back out at the
-    top. Cooling fluid is passed through two heat exchange
-    evaporators, one below Cage 0 and one above Cage 2.}
+    top. Cooling fluid is passed through heat exchange
+    evaporators, below Cage 0 and above Cage 2.}
   \label{fig:layout}

### 2020-05-27 intro.tex
done for today

@@
   \includegraphics[width=\textwidth]{figs/chronology001.pdf}
-  \caption{A seven-year tale of GPU life on a leadership class
-    supercomputer: Volume of GPU swaps detected on Titan at individual
+  \caption{Volume of GPU swaps detected on Titan at individual
     inventories (narrow blue) and yealy sum totals for 2014 and later

### 2020-05-27 dataprep.tex
done for today
fig 4 and \rev{} color command

@@
   \caption{Inter-inventory times, taken as time intervals between
-    unique dates in the data and removing those less than a day. Note
-    square root scale on the vertical axis.}
+    unique dates in the data and removing those less than a day.}
   \label{fig:inventory}
@@
 \begin{figure}[tb]
-{\tiny
+{\notsotiny
 \begin{verbatim}
@@
 \end{figure}
-where records for three are shown. Each record starts
-with a serial number and locations are coded with {\tt c{\it
-    col-row}c{\it cage}s{\it slot}n{\it node}}. The location
-references are cabinet column (0-24), row (0-7), cage (0-2), slot
-(0-7), and node (0-3) with respect to the layout shown in
-Figure~\ref{fig:layout}.
+where records for three are shown. Each record starts with a serial
+number and locations are coded with {\tt c{\it col-row}c{\it
+    cage}s{\it slot}n{\it node}}. The location references are cabinet
+column (0-24), row (0-7), cage (0-2), slot (0-7), and node (0-3) with
+respect to the layout shown in Figure~\ref{fig:layout}. \rev{For
+  example, we mark location {\tt c17-4c1s3n1} with a yellow dot in
+  each view of the Figure.}
 \begin{figure}

### 2020-05-27 acknowledgments.tex
Cleaning up.
finished reviewer 3

@@
+% !TEX root = main.tex
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Acknowledgments
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Acknowledgments}
-\label{sec:ack}
-\fix{Standard and all}
+\label{section:acknowledgments}
+
+We are extremely grateful to four anonymous reviewers who provided detailed and
+insightful comments on our initial submission, which resulted in a
+greatly improved presentation and deeper conclusions.
+
+This work was supported by \doubleblind{the U.S. Department of Energy, Office of Science,
+Office of Advanced Scientific Computing Research, program managers Robinson Pino
+and Lucy Nowell, under contract number DE-AC05-00OR22725.}{doubleblind}

### 2020-05-27 dataclean.tex
mostly typos and two graphs
finished reviewer 3

@@
 done by removing data for any units with a {\em remove} date before
-2014. After this reduction, there were stil six older units remaining,
+2014. After this reduction, there were still six older units remaining,
 which we also removed to have a clean set of units for the
@@
 To get some intuition for the GPU lifetimes on Titan, we give two
-views of 60 randomly selected GPUs (Figure~\ref{fig:gpuview}) and 80
-randomly selected locations (Figure~\ref{fig:locview}).
+views of 60 randomly selected GPUs (\rev{Fig.}~\ref{fig:gpuview}) and 80
+randomly selected locations (\rev{Fig.}~\ref{fig:locview}).
 The GPU view

### 2020-05-27 dataprep.tex
finished reviewer 3

@@
 Initial processing updates an inventory by checking the Serial Number
-(SN) and Location of each GPU and creating or updating a separate
-record for each contiguously observed SN-Location
+(\rev{{\em SN}}) and Location of each GPU and creating or updating a separate
+record for each contiguously observed \rev{{\em{SN-Location}}}
 combination. Resulting GPU data records are of the form shown in
@@
 column (0-24), row (0-7), cage (0-2), slot (0-7), and node (0-3) with
-respect to the layout shown in Figure~\ref{fig:layout}. \rev{For
+respect to the layout shown in \rev{Fig.}~\ref{fig:layout}. \rev{For
   example, we mark location {\tt c17-4c1s3n1} with a yellow dot in
-  each view of the Figure.}
+  each view of the figure.}
 \begin{figure}

### 2020-05-27 intro.tex
finished reviewer 3

@@
 Sec.~\ref{section:dataclean}. A time between failures (TBF) analysis
-is given in Sec.~\ref{section:tbf}. Survival analysis based on
+is given in Sec.~\ref{section:tbf}. Survival analysis \rev{(SA)} based on
 Kaplan-Meier modeling (KM) and Cox regression modeling (CR) of
@@
 Sec.~\ref{section:survival}. The goals of a TBF analysis and of
-survival analysis (SA) are different: TBF studies largely show GPU
+SA are different: TBF studies largely show GPU
 reliability affects the machine, whereas SA studies largely how the

### 2020-05-27 survival.tex
finished reviewer 3

@@
 \pkg{survminer} \cite{survminer-package} for the KM analysis, which is
-reported in Figure~\ref{fig:km-all-cage-node}. Within each {\em
+reported in \rev{Fig.}~\ref{fig:km-all-cage-node}. Within each {\em
   batch}, separate survival curves are computed for each {\em cage} by
@@
 in the old batch.  We reach this conclusion with reference to
-Figure~\ref{fig:layout} and relative positions of cages and nodes
+\rev{Fig.}~\ref{fig:layout} and relative positions of cages and nodes
 within a cabinet. Both cage and node differences in survival
@@
 do cross and so we fit the model to the new and the old batches
-separately. The results are presented in Figures~\ref{fig:cox-old}
+separately. The results are presented in \rev{Fig.}~\ref{fig:cox-old}
 and~\ref{fig:cox-new}.
@@
 horizontal hazard ratio axis is on a log scale. The estimates include
-uncertainty, which too is most reliably computed on a log scale (see,
-for example, \cite{Ostrouchov88}). In the figures, we also include the
-number of units, N, at risk and the number of events that occurred in
-each category.
+\rev{95\% confidence intervals}, which too \rev{are} most reliably
+computed on a log scale (see, for example, \cite{Ostrouchov88}). In
+the figures, we also include the number of units, N, at risk and the
+number of events that occurred in each category.
 
@@
 randomness of the failures and not for the geometric time averaging
-\cite{coxhazardinterpret}. Consequently, we interpret the pattern of
+\rev{\cite{Hernn2010TheHO}}. Consequently, we interpret the pattern of
 relative hazards rather than actual ratio magnitudes. More in-depth

### 2020-05-27 tbf.tex
mostly typos and two graphs
finished reviewer 3

@@
 A histogram of MTBFs measured across GPUs which had at least one failure event is shown 
-in Figure~\ref{fig:Device_MTBFs}. This is a practical assessment of device reliabilities 
+in \rev{Fig.}~\ref{fig:Device_MTBFs}. This is a practical assessment of device reliabilities 
 as opposed to those provided in the device datasheet. The failures are tracked using the
-{\em sn} of the GPUs even though a GPU might have been placed at different locations in the 
+\rev{{\em SN}} of the GPUs even though a GPU might have been placed at different locations in the 
 machine during its lifetime. The time to the first failure on a device is measured by taking
@@
 Apart from this, the distribution of MTBF due to DBE events looks very similar to that due 
-to OTB events. We also note in Figure~\ref{fig:Device_MTBFs} that the number of recorded 
+to OTB events. We also note in \rev{Fig.}~\ref{fig:Device_MTBFs} that the number of recorded 
 DBE events is much higher than the OTB events. This highlights the targeted replacement of GPUs 
@@
 A consistent drop in system-wide MTBF is observed starting from 2015-Q2 until 2016-Q4. 
-The corresponding increase in the number of failures during this phase can be seen in Figure~\ref{fig:NumFails_sys},
+The corresponding increase in the number of failures during this phase can be seen in \rev{Fig.}~\ref{fig:NumFails_sys},
 with a peak in 2016-Q4. The usability of the machine comes into
@@
 2017 replacement of many GPUs, there is a drastic increase in
-mean-time-between OTB failures, which is also evident in a reduction of OTB failures in Figure~\ref{fig:NumFails_sys}.
+mean-time-between OTB failures, which is also evident in a reduction of OTB failures in \rev{Fig.}~\ref{fig:NumFails_sys}.
 However, the system MTBF is determined by the weakest link and here the occurrence of DBE events tends to dictate it.
@@
 system. There is a re-emergence of an upward trend towards 
-the end in DBE failures in Figure~\ref{fig:NumFails_sys}, which
+the end in DBE failures in \rev{Fig.}~\ref{fig:NumFails_sys}, which
 appears to be due to older GPUs in the system.
@@
 To better understand the difference in reliability of newer and old portions of the machine, 
-Figure~\ref{fig:MTBF_sys_NewOld} shows the drastic difference in system MTBF measured in two hypothetical
+\rev{Fig.}~\ref{fig:MTBF_sys_NewOld} shows the drastic difference in system MTBF measured in two hypothetical
 partitions of the machine starting from the time when the majority of GPU replacements have been completed. 
@@
 For example, in 2018-Q4, the old partition MTBF is about 7.9 hours,
-whereas the newer partitio is 96 hours (4 days). The implications of such a huge disparity on applications running on the 
+whereas the newer partition is 96 hours (4 days). The implications of such a huge disparity on applications running on the 
 system are discussed in

### 2020-05-28 dataprep.tex
done for today

@@
 \end{figure}
-Notably, there was a period of nearly two months and many periods of over a
-week that Titan operated without a single reboot in the
-2014-2015 time frame. By the time the weekly inventories were added,
-failure rates started to increase and lead to more frequent reboots.
+Notably, there was a period of nearly two months and many periods of
+over a week that Titan operated without a single reboot in the
+2014-2015 time frame. \rev{However, by the time the weekly inventories
+  were added, failure rates started to increase and lead to more
+  frequent reboots.}
 
 Initial processing updates an inventory by checking the Serial Number
-(\rev{{\em SN}}) and Location of each GPU and creating or updating a separate
-record for each contiguously observed \rev{{\em{SN-Location}}}
-combination. Resulting GPU data records are of the form shown in
-Fig.~\ref{fig:dataraw},
+(\rev{{\em SN}}) and Location of each GPU and creating or updating a
+separate record for each contiguously observed
+\rev{{\em{SN-Location}}} combination. Resulting GPU data records are
+of the form shown in Fig.~\ref{fig:dataraw},
 \begin{figure}[tb]
@@
 }
-\caption{A few records of raw data produced from inventories that is
-  processed further in our analysis.}
+\caption{A few records of raw data produced from inventories \rev{and
+    log files} that is processed further in our analysis.}
 \label{fig:dataraw}
@@
   \end{minipage}
-  \caption{Titan cabinet (front view and side view with air flow) and
-    single blade layout (bottom left) and floor layout (bottom
-    right). Cooling airflow is bottom to top within a cabinet, taking
-    ambient air in mostly at the foot and relasing it back out at the
-    top. Cooling fluid is passed through heat exchange
-    evaporators, below Cage 0 and above Cage 2.}
+  \caption{\rev{Titan cabinet (front view and side view with air flow)
+      and single blade layout (bottom left) and floor layout (bottom
+      right). Cooling airflow is bottom to top within a cabinet,
+      taking ambient air in mostly at the foot and relasing it back
+      out at the top. Cooling fluid is passed through heat exchange
+      evaporators, below Cage 0 and above Cage 2. Yellow dot marks
+      location {\tt c17-4c1s3n1}}}
   \label{fig:layout}
@@
 
-The first GPU record
+The first GPU record in \rev{Fig.~\ref{fig:dataraw}}
 shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},

### 2020-05-28 intro.tex
done for today

@@
   \includegraphics[width=\textwidth]{figs/chronology001.pdf}
-  \caption{Volume of GPU swaps detected on Titan at individual
+  \caption{\rev{Volume of GPU swaps detected on Titan at individual
     inventories (narrow blue) and yealy sum totals for 2014 and later
@@
     and is focused on units in place by the end of 2013 or swapped in
-    during the years that follow.}
+    during the years that follow}.}
   \label{fig:chronology}

### 2020-05-30 code/TitanGPUmodel.Rmd
added R code

---
title: "Titan GPU model"
author: "GO"
output:
  html_document:
    css: style.css
    theme: cerulean
---
```{r setup, include=FALSE}
out_dir = "../figs/" # paper related figures will go here
in_dir = "../data/" # input files live here
history_file = paste0(in_dir, "titan.gpu.history.txt")
service_nodes = paste0(in_dir, "titan.service.txt")
source("TitanGPUsetup.R")
Rmd.time = proc.time()
```
`r today()`
Read the data. Last inventory was run 01/20/2020 but Titan was turned off on 08/01/2019
```{r}
g_raw = fread(history_file, col.names = c("SN", "location", "insert", "remove"))
g_raw$remove[stringr::str_starts(g_raw$remove, "01/20/2020")] =
  "08/01/2019 20:07:33"
cat("Total of", (unique_SN = length(unique(g_raw$SN))), "unique SN\n")
cat("Total of", (unique_loc = length(unique(g_raw$location))),
    "unique locations\n")
```

**Enable DST corrected dates** for *insert* and *remove* with Eastern time zone.
**Enable missing values**, and **create** some new variables.
Variable *event* is "life" when both *insert* and *remove* are present, "life0" when *insert* = *remove*, otherwise it is whatever string is in *insert*, which is "DBE" or "Off The BUS" (renamed to "OTB"). 
**Fill in** (repeat) serial numbers and locations for all records.
```{r}
time_zone = "America/New_York"
g_ev = g_raw %>% 
  mutate(event = if_else(is.na(lubridate::mdy_hms(insert, quiet = TRUE)),
                         insert,  # if not a date, event is the insert string
                         "life"), # if a date, it's a life span
         event = if_else(event == "Off The BUS", "OTB", event), # recode "Off The BUS" to "OTB"
         location = na_if(location, ""),  # blank to NA for zoo later
         insert = lubridate::mdy_hms(insert, tz = time_zone, quiet = TRUE),
         remove = lubridate::mdy_hms(remove, tz = time_zone), # !quiet incase NA
         duration = lubridate::as.duration(remove - insert), # in seconds
         event = as.factor(if_else(event == "life" & as.numeric(duration) == 0,
                                   "life0", event)),
         SN = ifelse(SN == "", as.character(NA), SN), # blank to NA for zoo
         SN = zoo::na.locf(SN), # fill in missing SN (forward)
         SN_ok = grepl("[0-9]{13}", SN), # verify that SN is 13 numbers
         location = zoo::na.locf(location), # fill in missing location (forward)
         location_ok = grepl("c[0-9][0-9]?-[0-7]c[0-2]s[0-7]n[0-3]", location))
```

**List** bad SN and bad location raw data. **Remove** bad SN and bad location data. **Read and remove** service node data.
```{r fig.width = 10}
good_date_remove = lubridate::mdy_hms("01/01/2014 00:00:00", tz = time_zone)
good_date_insert = lubridate::mdy_hms("11/01/2013-00:00:00", tz = time_zone)
good_date_last = lubridate::mdy_hms("12/01/2016 00:00:00", tz = time_zone)

g_raw[!g_ev$SN_ok, ]  # list bad SN records
g_raw[!g_ev$location_ok, ] # list bad location records
service = fread(service_nodes, col.names = c("node", "loc", "location", "type", "up", "run"))
cat("Total of", (unique_sloc = length(unique(service$location))),
    "unique service locations\n")
service_notseen =
  service$location[!(service$location %in% unique(g_ev$location))]
cat("Service locations not referenced in data:", service_notseen, "\n")

gc_ev = g_ev %>%
  filter(SN_ok, location_ok, !(location %in% service$location)) %>%
  select(-SN_ok, -location_ok) # remove service node data (must be after zoo!)
cat("Total of", (unique_SN = length(unique(gc_ev$SN))), "unique clean SN\n")
cat("Total of", (unique_loc = length(unique(gc_ev$location))),
    "unique clean locations\n")
top_remove_dates = sort(table(g_ev$remove), decreasing = TRUE)[1:25]
trdat = top_remove_dates[order(names(top_remove_dates))]
top_insert_dates = sort(table(g_ev$remove), decreasing = TRUE)[1:25]
top_insert_dates[order(names(top_insert_dates))]
trd_all = data.frame(ht = ifelse(as.numeric(trdat) < 2000, 2000, 
                                 as.numeric(trdat)*0.95),
                     date = as.POSIXct(names(trdat)))
trd = trd_all[c(1:7, 20:25), ]

ggp(ggplot(filter(g_ev %>% filter(event == "life"), remove < max(remove)), aes(remove)) + 
      geom_histogram(
        data = filter(g_ev, remove < max(remove)),
#        data = filter(g_ev, remove >= good_date_remove, remove < max(remove)),
        mapping = aes(remove), binwidth = 60*60*24*182.5, alpha = 0.2,
        boundary = good_date_remove) +
      geom_histogram(binwidth = 60*60*24, color = "blue") + theme_bw() +
      theme(axis.title.x = element_blank()) + scale_x_datetime(breaks = "1 year"), #+
#      scale_y_continuous(limits = c(0, 9250), breaks = seq(0, 9000, 1000)),
    file = "chronology", width = 10, height = 3)

removes_life = table((g_ev$remove[g_ev$event == "life"]))
removes_life0 = table((g_ev$remove[g_ev$event == "life0"]))
removes_OTB = table((g_ev$remove[g_ev$event == "OTB"]))
removes_DBE = table((g_ev$remove[g_ev$event == "DBE"]))
chron = data.frame(remove = lubridate::ymd_hms(names(removes_life)), count = as.numeric(removes_life))
ggplot(chron, aes(count, remove)) + geom_point() + facet_grid(year(remove) ~ ., scales = "free")
```
The next analysis looks at *life0* events, plots their frequency, and lists a sample of 10 such events in the raw input data.
```{r}
zeros = g_ev[g_ev$event == "life0", ]
ggplot(zeros, aes(remove)) + 
      geom_histogram(data = zeros,
        mapping = aes(remove), binwidth = 60*60*24*182.5, alpha = 0.2,
        boundary = good_date_remove) +
      geom_histogram(binwidth = 60*60*24, color = "blue") + theme_bw() +
      theme(axis.title.x = element_blank()) + scale_x_datetime(breaks = "1 year")
sn_zeros = zeros$SN[sample(length(zeros$SN), 10)]
for(sn in sn_zeros) print(g_raw[(i = which(g_raw$SN == sn)):(i+10), ])
```

Begin data selection by removing all GPU life times that have *remove* date before *good_date* and all that have zero life times. The premise is that record keeping was different and possibly inconsistent before *r good_date*. **gc_ev_nzg** is used in many analyses going forward. Five original SNs remain after this and are removed manually to have a clean "post 2nd rework cycle" data set. 
```{r}
gc_ev_nzg = gc_ev %>%
  filter(remove > good_date_remove, # drop GPUs removed before good_date!
         !(SN %in% c("0323712022786", "0323712007923", "0323712007994",
                   "0323712022956", "0323712042970")), # rm insert dates < 2013-06
         event != "life0") %>% mutate(event = fct_drop(event))  # remove life0's
last_inventory = max(gc_ev_nzg$remove, na.rm = TRUE) # last right-censoring date
cat("gc_ev dim", dim(gc_ev), "processed into gc_ev_nzg dim", dim(gc_ev_nzg), "\n")
cat("before good_date", sum(gc_ev$remove <= good_date_remove, na.rm = TRUE), ": life0",
    sum(gc_ev$event == "life0", na.rm = TRUE), "remove is NA", sum(is.na(gc_ev$remove)), "\n")
```
**Mark time overlaps within SN, then remove SN records that overlap and plot before-after**. 
SN time overlaps are possible when an inventory is incomplete. A unit is not detected as removed, yet is detected as inserted in a different location - thus the SN appears at both locations until the next inventory.
```{r fig.height = 10}
gc_life = mark.overlaps(gc_ev_nzg %>% filter(event == "life") %>%
                          select(-event), SN)
gc_ev_fail = gc_ev_nzg %>% select(SN, remove, location, event) %>%
  filter(event != "life") %>% arrange(SN, remove, location) %>%
  mutate(event = fct_drop(event))
plot.life_ev(gc_life, gc_ev_nzg, SN, overlaps = TRUE, outs = FALSE, file = "overlap_SN")
gc_life = gc_life %>% filter(!overlap_rec)
plot.life_ev(gc_life, gc_ev_nzg, SN, overlaps = TRUE, outs = FALSE, file = "overlap_SN_after") 
```
For locations, **remove only the overlapping life** but keep the rest of the SN. The overlaps seem very short so this makes sense. Why do location overlaps exist? In some cases, they seem to be the full blade of 4 GPUs. Blades were sometimes hot-swapped by CRAY and if this was during an inventory run, the results could be unpredictable. **Plot overlap at locations again after removing the offending life spans.*
```{r fig.height = 10}
gc_life = mark.overlaps(gc_life, location)
plot.life_ev(gc_life, gc_ev_nzg, location, overlaps = TRUE, outs = FALSE, file = "overlap_Loc")

for(i in which(g_raw$location == "c7-7c0s7n3")) print(g_raw[(i-5):i, ])
for(i in which(g_raw$location == "c22-0c2s0n0")) print(g_raw[(i-5):i, ])
gc_life = gc_life %>% filter(!overlap_rec)
plot.life_ev(gc_life, gc_ev_nzg, location, overlaps = TRUE, outs = FALSE, file = "overlap_Loc_after")
```
After removing overlaps, determine *out* = "last seen" record that may indicate censoring
```{r}
gc_life = gc_life %>% group_by(SN) %>%
  mutate(out = if_else(remove == max(remove) & remove < last_inventory,
                                     TRUE, FALSE)) %>% ungroup()
```
**Inventory interval analysis**
```{r fig.width=10, fig.height=4}
life_insert = gc_life$insert # collect insert dates
life_remove = gc_life$remove # collect remove dates
dates_dbe = sort(unique(filter(gc_ev_fail, event == "DBE")$remove)) # unique DBE dates
dates_otb = sort(unique(filter(gc_ev_fail, event == "OTB")$remove)) # unique OTB dates
dates_life = sort(unique(c(life_insert, life_remove))) # unique insert,remove

life_intervals =
  data.frame(days = as.numeric(diff(dates_life))/(60*60*24),
             year = factor(as.character(year(dates_life[-1])))) %>% filter(days >= 1) %>% 
  mutate(days = factor(as.integer(floor(days))))
ggp(ggplot(life_intervals, aes(days, fill = year)) + stat_count() +
      theme_bw(), file = "attention_intervals", height = 3)
```

```{r}
nsample = 88 # the max number for a full page pair of figures
set.seed(76513)  # for reproducibility!
fig_height = as.integer(nsample*0.1 + 0.5)
```
**Plot a sample** of `r nsample` GPUs to check that things look right ...
* starts: black dots (gray for zero lifetime)
* lifetimes: black lines
* DBEs: red triangles
* OTB: blue triangles
* Last seen: black ]
```{r}
SN_sample = sample(unique(gc_ev_nzg$SN), nsample)
ev_sample = gc_ev_nzg %>% filter(SN %in% SN_sample)
life_sample = gc_life %>% filter(SN %in% SN_sample)
plot.life_ev(life_sample, ev_sample, SN, "sample_SN", overlaps = FALSE)
```

**Plot a sample** of `r nsample` locations ...
```{r}
loc_sample = sample(unique(gc_ev_nzg$location), nsample)
ev_sample = gc_ev_nzg %>% filter(location %in% loc_sample)
life_sample = gc_life %>% filter(location %in% loc_sample) %>% 
  separate(location, c(NA, "col", "row", "cage", "slot", "node"),
           sep = "[-csn]", convert = TRUE, remove = FALSE) %>%
  mutate(location = reorder(location, col)) # orders in increasing col
plot.life_ev(life_sample, ev_sample, location, "sample_loc", overlaps = FALSE)
```

Add a % uptime in a location histogram. Might be a good diagnostic

Reduce to only "life" records and mark them with ending DBE, OTB, "out", or none events. The result is just *insert* and *remove* with *event* marks to indicate DBE or OTB or "out" (last seen). *bad* indicates that a DBE or OTB occurred yet the GPU was not taken out. For a first cut, keep the *bad* ones in. Also keep *duration* to later aggregate into full life times.

```{r}
gc_life = gc_life %>% select(SN, location, insert, remove, duration, out) %>%
  arrange(SN, remove, location)
gc_fulljoin = gc_life %>% full_join(gc_ev_fail, by = c("SN", "remove", "location"))
gc_full = gc_fulljoin %>% group_by(SN, remove, location) %>%
  mutate(insert = if_else(row_number() > 1, as.POSIXct(NA), insert),
         duration = if_else(row_number() > 1, as.numeric(NA), as.numeric(duration)),
         out = if_else(row_number() > 1, as.logical(NA), out)) %>% ungroup()
```
**Output** gc_full data frame to a CSV file for TBF analysis with Python code
```{r}
gc_write = as.data.frame(lapply(gc_full, as.character), stringsAsFactors = FALSE)
write_csv(gc_write, paste0(out_dir, "gc_full.csv"), na = "", col_names = TRUE)
```
**Message to reader of file:** All fields are character strings. Dates are with timezone. Missing values as "<NA>" strings.

Next, aggregate into one record per *SN* with total life time and first *insert* time. Indicate if event occurred or still in service: out = fail, dbe = fail, otb = fail, NA = right censored (still in service). Add some other variables, such as GPU in single location or moved one or more times, mark *new_batch* group, etc. to use in modeling later.

**This is where life times are determined**, connected with a location and with an event (OTB, DBE, out). Note that location is imprecise if more than one is used (we use longest). Get proportion of time at reference location *time_max_loc*.
```{r}
gc_summary = gc_full %>% arrange(SN, remove) %>% group_by(SN) %>%
  summarize(
    time = sum(duration, na.rm = TRUE), # total life time
    nlife = sum(!is.na(duration)), # number of life records
    nloc = length(unique(location)), # Number of locations where lived
    last = max(remove), # last time stamp
    max_loc = ifelse(!all(is.na(insert)), location[which.max(duration)], ""),
    max_loc_events = ifelse(all(is.na(event)), 0,
                            length(event[location == max_loc])),
    time_max_loc = sum(duration[location == max_loc], na.rm = TRUE)/time,
    dbe = sum(event == "DBE", na.rm = TRUE),
    dbe_loc = ifelse(all(is.na(duration)) | dbe == 0, NA,
                     max_loc %in% location[event == "DBE"]),
    otb = sum(event == "OTB", na.rm = TRUE),
    otb_loc = ifelse(all(is.na(duration)) | otb == 0, NA,
                     max_loc %in% location[event == "OTB"]),
    out = any(out, na.rm = TRUE), # removed
    batch = if_else(min(insert, na.rm = TRUE) >
                      lubridate::mdy_hms("01/01/2016 00:00:00", tz = time_zone), "new", "old")
  )
gc_summary
```

**Separate max_loc** into components (col, row, cage, slot, node). Note that this is the place this *SN* was located the longest. This muddies the hazard analysis somewhat as several locations can be the "treatment" for a given *SN*. But the proportions of time at max_loc tend to be very high. Time-dependent covariates, see: https://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
```{r}
gc_summary_loc = gc_summary %>% separate(max_loc, c(NA, "col", "row", "cage", "slot", "node"),
           sep = "[-csn]", convert = TRUE, remove = TRUE) %>% # c{col}-{row}c{cage}s{slot}n{node}
  mutate( col = as.factor(col), row = as.factor(row), cage = as.factor(cage), 
          slot = as.factor(slot), node = as.factor(node),
          days = time/(60*60*24),
          years = days/365,
          dead = out & dbe + otb > 0,
          dead_otb = out & otb > 0,
          dead_dbe = out & dbe > 0) # define fail & censored

gc_summary_loc_o = gc_summary_loc %>% filter(batch == "old")
gc_summary_loc_n = gc_summary_loc %>% filter(batch == "new")
gc_summary_loc_censor1 = gc_summary_loc %>% 
  mutate(dead = if_else(batch == "old" & years > 1, FALSE, dead), 
         days = if_else(batch == "old" & years > 1, 1*365, days),
         years = if_else(batch =="old" & years > 1, 1, years))  # censor at 1 year
gc_summary_loc_censor2 = gc_summary_loc %>% 
  mutate(dead = if_else(batch == "old" & years > 2, FALSE, dead), 
         days = if_else(batch == "old" & years > 2, 2*365, days),
         years = if_else(batch =="old" & years > 2, 2, years))  # censor at 2 years
gc_summary_loc_censor3 = gc_summary_loc %>% 
  mutate(dead = if_else(batch == "old" & years > 3, FALSE, dead), 
         days = if_else(batch == "old" & years > 3, 3*365, days),
         years = if_else(batch =="old" & years > 3, 3, years))  # censor at 3 years
gc_summary_loc_censor4 = gc_summary_loc %>% 
  mutate(dead = if_else(batch == "old" & years > 4, FALSE, dead), 
         days = if_else(batch == "old" & years > 4, 4*365, days),
         years = if_else(batch =="old" & years > 4, 4, years))  # censor at 4 years
```
**Output** gc_full to a CSV file
```{r}
gc_write = as.data.frame(lapply(gc_summary_loc, as.character), stringsAsFactors = FALSE)
write_csv(gc_write, paste0(out_dir, "gc_summary_loc.csv"), na = "", col_names = TRUE)
```
Get some stats for sanity check.
```{r}
cat("Total", nrow(gc_summary_loc), "unique lifetimes\n")
cat("  Censored:", sum(!gc_summary_loc$dead), "  dead:", sum(gc_summary_loc$dead),
    "  Known alive:", sum(!gc_summary_loc$out), "\n")
cat("  DBE or OTB during life:", sum(gc_summary_loc$dbe > 0 | gc_summary_loc$otb >0), "\n")
cat("           More than one:", sum(gc_summary_loc$dbe + gc_summary_loc$otb > 1, na.rm = TRUE), "\n")
ggplot(gc_summary_loc, aes(x = time)) + geom_histogram(bins = 60) + ggtitle("GPU Life Times")
ggplot(gc_summary_loc, aes(x = time)) + geom_histogram(bins = 60) + 
  facet_wrap(~ batch) + ggtitle(paste("GPU Life Times - new and old"))
ggplot(gc_summary_loc, aes(x = time_max_loc)) + geom_histogram(bins = 200) +
  ggtitle("Life Proportion at Longest Location") + theme_bw()
ggplot((gc_summary_loc %>% filter(dead)), aes(x = time_max_loc)) +
      geom_histogram(bins = 200) + theme_bw() +
      ggtitle("Life Proportion at Longest Location - Dead Only")
```

Kaplan-Meier survival analysis gives the probability of survival beyond a given time. It is a nonparametric technique that makes no specific failure model assumptions, such as Weibull, Exponential, etc. It can also split the data into subpopulations and compute separate survival curves.

If $T$ is taken to be the random variable of a GPU fail time, then its probability density function is $f(t)$ and its cumulative distribution function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails by duration $t$. The survival function is then
$$S(t) = Pr\{T \geq t\} = 1 - F(t) = \int_t^\infty f(x)dx,$$ 
which is the probability of being operational just before duration $t$.

A sister concept to the survival function is the hazard function:
$$H(t) = \frac{f(t)}{S(t)},$$
which is what is sometimes called the "bathtub" curve in reliability.

Units removed without a DBE are right-censored! They are not failures!

Next, we consider splitting the GPU into groups - installed before 2016 (old) or later (new).

**SN categories**
I tried disassembling the *SN* into digits and did not find any compelling results.

Now for Kaplan-Meier analysis ...
```{r fig.height = 8}
library(survival)
library(survminer)
ggp(ggsurvplot(survfit(
  Surv(years, dead, type = 'right') ~ node, data = gc_summary_loc),
  facet.by = c("batch", "cage"), conf.int = TRUE, risk.table = TRUE,
  ggtheme = theme_bw(), xlab = "Years", legend.title = "Node", censor.size = 2,
  legend = c(0.08, 0.73), legend.labs = 0:3, ylim = c(0.08, 1), size = 0.4, 
  break.y.by = 0.1, censor = FALSE),
  width = 10.5, height = 6, file = "km_cage-node_a")

fit = coxph(Surv(years, dead, type = 'right') ~ row + col + cage + slot + node,
            data = gc_summary_loc_o)
## below seems to have an issue with RStudio: 
## https://community.rstudio.com/t/ggcoxdiagnostics-warning-message/51867
## ggcoxdiagnostics(fit, type = "schoenfeld", ox.scale = "time")
test.ph = cox.zph(fit)
ggcoxzph(test.ph)
```

Try PAMM models via pammtools
```{r}
library(dplyr)
library(tidyr)
library(pammtools)
mdat = gc_summary_loc_o %>% select(SN, years, dead, col, row, cage, slot, node)
ped = as_ped(Surv(years, dead, type = "right") ~ ., data = mdat, id = "SN",
             cut = seq(0, 6, 1/12)) #c(0, 1.5, 2, 2.5, 3, 4, 6))
library(survival)
library(mgcv)
base_df = basehaz(coxph(Surv(years, dead) ~ 1, data = mdat)) %>% rename(nelson_aalen = hazard)
ggplot(base_df, aes(x = time, y = nelson_aalen)) + geom_stephazard() +
  ylab(expression(hat(Lambda)(t))) + xlab("Years") +
  ggtitle("Nelson-Aalen estimate of the cumulative hazard")
pam = gam(ped_status ~ s(tend), data = ped, offset = offset, family = poisson())
plot(pam)
```
The Cox proportional hazards (CPH) regression analys takes the hazard function above and a set of covariates $x$ and models $k$s GPU hazard as
$$H_k(t) = H_0(t)e^{\sum\limits_{i=1}^n{\beta x}}$$
That is, a base hazard rate multiplied by a function of covariates. The key assumption in the CPH model is that the "hazards" are multipliers on the "baseline hazard" but the estimate of the baseline hazard is nonparametric, meaning it makes no specific distributional assumption and just learns from the data (unlike analyses that assume for example a Weibul or an exponential model). The multiplicative assumption is that the hazard curves do not cross and are multiples of each other. The multiplier is the *hazard coefficient*. For example, if the baseline is *node0* and the *node2* hazard is 2, then *node2* sees twice as many failures as *node0* on average.

For our comparison of hazards, we use as *covariates* all the unpacked location variables, indicators whether the GPU has been moved, and our notion of old and new GPUs based on first install date. Some GPUs experience time at more than one location but these are few. We use the longest period location in this analysis. We do separate analysis for old and new batches.
```{r fig.width = 5, fig.height = 10}
response = "Surv(time, event = dead) ~"
gc_summary_loc_o = gc_summary_loc %>% filter(batch == "old")
gc_summary_loc_n = gc_summary_loc %>% filter(batch == "new")

library(forestmodel)
panels = list(
  forest_panel(width = 0.03),
  forest_panel(width = 0.10, display = variable, fontface = "bold",
               heading = "Variable"),
  forest_panel(width = 0.05, display = level, fontface = "bold"),
  forest_panel(width = 0.05, display = n, heading = "N"),
  forest_panel(width = 0.05, display = n_events, heading = "Events"),
  forest_panel(width = 0.03, item = "vline", hjust = 0.5),
  forest_panel(width = 0.91, item = "forest", line_x = 0, hjust = 0.5,
               heading = "Hazard ratio", linetype = "dashed"),
  forest_panel(width = 0.03))

covariates = c("col", "row", "cage", "slot", "node")
col2torus = order(c(0, seq(1, 23, 2), seq(24, 2, -2)))
levels(gc_summary_loc_o$col) = 
  paste0(formatC(levels(gc_summary_loc_o$col), width = 2),
         " (X-", formatC(col2torus, width = 2), ")")
mv_model = coxph(as.formula(paste(response, paste(covariates, collapse = " + "))),
                 data = gc_summary_loc_o)
ggp(forest_model(mv_model, panels, factor_separate_line = TRUE), width = 5,
    height = 7, file = "cox_o")

tcol = col2torus[gc_summary_loc_o$col]
gc_summary_loc_o = gc_summary_loc_o %>% mutate(col = reorder(col, tcol))
mv_model = coxph(as.formula(paste(response, paste(covariates, collapse = " + "))),
                 data = gc_summary_loc_o)
ggp(forest_model(mv_model, panels, factor_separate_line = TRUE), width = 5,
    height = 7, file = "cox_o_t")

levels(gc_summary_loc_n$col) = 
  paste0(formatC(levels(gc_summary_loc_n$col), width = 2),
         " (X-", formatC(col2torus, width = 2), ")")
mv_model = coxph(as.formula(paste(response, paste(covariates, collapse = " + "))),
                 data = gc_summary_loc_n)
ggp(forest_model(mv_model, panels, factor_separate_line = TRUE), width = 5,
    height = 7, file = "cox_n")

tcol = col2torus[gc_summary_loc_n$col]
gc_summary_loc_n = gc_summary_loc_n %>% mutate(col = reorder(col, tcol))
mv_model = coxph(as.formula(paste(response, paste(covariates, collapse = " + "))),
                 data = gc_summary_loc_n)
ggp(forest_model(mv_model, panels, factor_separate_line = TRUE), width = 5,
    height = 7, file = "cox_n_t")
```

```{r}
cat("Total time:\n"); print(proc.time() - Rmd.time)
```

### 2020-05-30 code/TitanGPUsetup.R

@@ -1,0 +1,119 @@
+## setup and functions
+
+knitr::opts_chunk$set(echo = TRUE)
+options(width = 150)
+
+suppressMessages(library(data.table))
+suppressMessages(library(tidyverse))
+suppressMessages(library(lubridate))
+
+## Function to mark overlapping life spans in unit (sn or location). Adds two
+## variables: overlap_unit and overlap_rec, marking all records in a unit and
+## specific overlapping record in a unit, respectively.
+mark.overlaps = function(glife, unit) {
+  unit = enquo(unit) # to remove need for quoting and use !! in eval
+  glife %>% 
+    arrange(!! unit, insert) %>% group_by(!! unit) %>% 
+    mutate(order = base::order(insert), # increasing order of inserts
+           overlap_unit = 
+             if_else(n() == 1,
+                     FALSE,
+                     any(insert[order][-1] < lag(remove[order])[-1])),
+           overlap_rec = 
+             if_else(rep(n() == 1, n()),
+                     rep(FALSE, n()), 
+                     c(FALSE, insert[order][-1] < lag(remove[order])[-1]))) %>%
+    ungroup() %>% select(-order) %>% arrange(!! unit, remove)
+}
+
+## Function for plotting life (unit is sn or location) with events 
+## and optionally overlaps.
+plot.life_ev = function(glife, gevents, unit, file = FALSE, overlaps = TRUE,
+                        outs = TRUE) {
+  unit = enquo(unit) # enquoting treatment for quoteless operation
+  td = as.duration(sum(glife$duration))
+  if (overlaps) {
+    glife = glife %>% filter(overlap_unit)
+    units = unique(unlist(glife %>% select(!! unit)))
+    gevents = gevents %>% filter(!! unit %in% units)
+    nu = length(units)
+    cat("There are", nu, "units with life overlap:\n")
+    od = as.duration(sum(glife$duration[glife$overlap_rec == TRUE]))
+    print(paste("Overlap life time", od, "of total life time", td,
+                "(", od/td, "proportion )"))
+    p = ggplot(glife, aes(x = insert, xend = remove, y = !! unit,
+                          yend = !! unit, color = overlap_rec)) + 
+      scale_color_manual(values=c("black", "red"))
+  } else {
+    p = ggplot(glife, aes(x = insert, xend = remove,
[…]

### 2020-05-30 survival.tex
done for today
updated some figs

@@
 \label{section:survival}
-\renewcommand{\pkg}[1]{\textsf{#1}} Survival analysis (SA) methods
+Survival analysis (SA) methods
 \cite{survival-book}, 
-sometimes referred as time to event methods, focus on the effect of
+sometimes referred to as time to event methods, focus on the effect of
 the machine on the GPU units. These methods have roots in biomedical
@@
   \label{fig:cox-old}
+\end{figure}
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{figs/cox_o_t001.pdf}
+  \caption{\rev{GPU hazard ratios from Cox regression model on {\em
+        old} batch. Putting columns in logical torus order removes the
+      peculiar interleaving pattern and presents a more physically
+      reasonable response. This suggests that scheduling job fill from
+      X-0 had measurable temperature impact on the cooler side of
+      Titan (evaporators get engaged only at a temperature
+      threshold).}}
+  \label{fig:cox-torus}
 \end{figure}

### 2020-06-03 code/TitanGPUmodel.Rmd
talked around pre 2014 data

@@
       geom_histogram(
-        data = filter(g_ev, remove < max(remove)),
-#        data = filter(g_ev, remove >= good_date_remove, remove < max(remove)),
-        mapping = aes(remove), binwidth = 60*60*24*182.5, alpha = 0.2,
+#        data = filter(g_ev, remove < max(remove)),
+        data = filter(g_ev, remove >= good_date_remove, remove < max(remove)),
+        mapping = aes(remove), binwidth = 60*60*24*365, alpha = 0.2,
         boundary = good_date_remove) +
@@
     file = "chronology", width = 10, height = 3)
-
-removes_life = table((g_ev$remove[g_ev$event == "life"]))
-removes_life0 = table((g_ev$remove[g_ev$event == "life0"]))
-removes_OTB = table((g_ev$remove[g_ev$event == "OTB"]))
-removes_DBE = table((g_ev$remove[g_ev$event == "DBE"]))
-chron = data.frame(remove = lubridate::ymd_hms(names(removes_life)), count = as.numeric(removes_life))
-ggplot(chron, aes(count, remove)) + geom_point() + facet_grid(year(remove) ~ ., scales = "free")
 ```
@@
 ```{r}
-zeros = g_ev[g_ev$event == "life0", ]
+d2013a = as.POSIXlt.character("2014-05-30 00:00:00 EST")
+d2013b = as.POSIXlt.character("2014-06-23 00:00:00 EST")
+zeros = g_ev %>% filter(event == "life0" | event == "life", remove > d2013a, remove < d2013b)
 ggplot(zeros, aes(remove)) + 
-      geom_histogram(data = zeros,
-        mapping = aes(remove), binwidth = 60*60*24*182.5, alpha = 0.2,
-        boundary = good_date_remove) +
-      geom_histogram(binwidth = 60*60*24, color = "blue") + theme_bw() +
-      theme(axis.title.x = element_blank()) + scale_x_datetime(breaks = "1 year")
+#      geom_histogram(data = zeros,
+#        mapping = aes(remove), binwidth = 60*60*24*182.5, alpha = 0.2,
+#        boundary = good_date_remove) +
+      geom_histogram(binwidth = 60*60, color = "blue") + theme_bw() +
+      theme(axis.title.x = element_blank()) + scale_x_datetime(breaks = "1 day") + ylim(c(0, 25))
 sn_zeros = zeros$SN[sample(length(zeros$SN), 10)]
@@
   width = 10.5, height = 6, file = "km_cage-node_a")
-
+```
+Consider diagnostics for proportionalty.
+```{r}
 fit = coxph(Surv(years, dead, type = 'right') ~ row + col + cage + slot + node,
@@
 ggcoxzph(test.ph)
-```
-
-Try PAMM models via pammtools
-```{r}
-library(dplyr)
-library(tidyr)
-library(pammtools)
-mdat = gc_summary_loc_o %>% select(SN, years, dead, col, row, cage, slot, node)
-ped = as_ped(Surv(years, dead, type = "right") ~ ., data = mdat, id = "SN",
-             cut = seq(0, 6, 1/12)) #c(0, 1.5, 2, 2.5, 3, 4, 6))
-library(survival)
-library(mgcv)
-base_df = basehaz(coxph(Surv(years, dead) ~ 1, data = mdat)) %>% rename(nelson_aalen = hazard)
-ggplot(base_df, aes(x = time, y = nelson_aalen)) + geom_stephazard() +
-  ylab(expression(hat(Lambda)(t))) + xlab("Years") +
-  ggtitle("Nelson-Aalen estimate of the cumulative hazard")
-pam = gam(ped_status ~ s(tend), data = ped, offset = offset, family = poisson())
-plot(pam)
 ```
@@
 
+col2torus = order(c(0, seq(1, 23, 2), seq(24, 2, -2)))
 library(forestmodel)
@@
 covariates = c("col", "row", "cage", "slot", "node")
-col2torus = order(c(0, seq(1, 23, 2), seq(24, 2, -2)))
 levels(gc_summary_loc_o$col) = 

### 2020-06-03 dataprep.tex
talked around pre 2014 data

@@
 Data on Titan GPU life times is constructed from two sources:
-inventory runs and failure event records. Two types of failure events
-were collected: Double Bit Error (DBE) and Off the Bus (OTB). Both
-errors were found to be the ``signature'' event of GPU failures and a
-trigger for their replacement.
+inventory runs and failure event \rev{log} records. Two types of
+failure events were collected \rev{for this study}: Double Bit Error
+(DBE) and Off the Bus (OTB). \rev{While other types of errors were
+  recorded in log files, as is reported in \cite{7266836}, these two
+  kinds of errors became dominant in 2016 and were found to be the
+  ``signature'' event of the GPU board failing resistor and a trigger
+  for GPU replacement.}
 
@@
 and farther apart. A single inventory takes about a minute and is
-recorded in a separate file. The DBE and OTB events are also recorded
-separately from the inventories.  Figure~\ref{fig:inventory} shows
-inter-inventory times that appear in the data.
+recorded in a separate file. \rev{An inventory can be incomplete if a
+  node response times out but such occurrences were relatively rare}.
+Figure~\ref{fig:inventory} shows inter-inventory times that appear in
+the data.
 \begin{figure}[bt]
@@
   were added, failure rates started to increase and lead to more
-  frequent reboots.}
+  frequent reboots and thus inventories.}
 
@@
 column (0-24), row (0-7), cage (0-2), slot (0-7), and node (0-3) with
-respect to the layout shown in \rev{Fig.}~\ref{fig:layout}. \rev{For
+respect to the layout shown in \rev{Fig.}~\ref{fig:layout}. \rev{Note
+  that there is an extra empty cabinet between columns 10 and 11 to
+  accommodate a few ceiling support columns. For
   example, we mark location {\tt c17-4c1s3n1} with a yellow dot in
@@
       out at the top. Cooling fluid is passed through heat exchange
-      evaporators, below Cage 0 and above Cage 2. Yellow dot marks
-      location {\tt c17-4c1s3n1}}}
+      evaporators, below Cage 0 and above Cage 2. Ambient air in the
+      room is conditioned at nominal 75$^\circ$~F. This system has a
+      dozen intakes near the ceiling and dumps cool air under the
+      floor in several locations. A number of floor tiles are
+      perforated.  Yellow dot marks GPU location {\tt c17-4c1s3n1}}}
   \label{fig:layout}

### 2020-06-03 intro.tex
talked around pre 2014 data

@@
 periodic inventories. The first two rework cycles before 2014
-involving blade mechanical assembly can be characterized as a break-in
-period on a new system that is first of its kind and pushes many
-technological boundaries. The resulting field engineering as well as
-data collection changes are responsible for the high GPU swap counts
-before 2014. Late 2013 begins a very long period of high reliability
-and stable operation with very few issues until about mid 2016, when
-GPU failures begin to rise. This results in the final rework cycle,
-replacing 11,000 GPUs throughout much of 2017. The peak at the end in
-the figure is an artifact of turning off or ``removing'' all the
-units as Titan is shut down and disassembled.
+involving blade mechanical assemblies can be characterized as a
+break-in period on a new system that is first of its kind and pushes
+many technological boundaries. \rev{This early period included
+  extensive field engineering and experimentation to determine cause,
+  down times, and large temporal and completeness gaps in inventory
+  runs. We include the early swap data in this figure only for
+  completeness and to underscore the amount of work required to field
+  a world's largest supercomputer. However, we exclude this period
+  from further analysis because we have a gap in explanation for one
+  of the swap spikes and because the separation is made possible by
+  the second, late 2013, GPU rework cycle.} Late 2013 begins a very
+long period of high reliability and stable operation with very few
+issues until about mid 2016, when GPU failures begin to rise. This
+results in the final rework cycle, replacing 11,000 GPUs throughout
+much of 2017.
 
@@
 involving data prior to 2014. All analyses and figures that follow use
-only data from 2014 onward. We make the data as well as our analysis R
+data from 2014 onward. We make the data as well as our analysis R
 codes and Python codes to reproduce the graphics in this paper
@@
 Sec.~\ref{section:survival}. The goals of a TBF analysis and of
-SA are different: TBF studies largely show GPU
+SA are different: TBF studies largely \rev{show} GPU
 reliability affects the machine, whereas SA studies largely how the

### 2020-06-03 survival.tex
talked around pre 2014 data

@@
   \caption{GPU hazard ratios from Cox regression model on {\em old}
-    batch.}
+    batch. \rev{All variables, {\em col}, {\em row}, {\em cage}, {\em
+      slot}, and {\em node} are with respect to spatial locations
+    shown in Fig.~\ref{fig:layout}}}
   \label{fig:cox-old}
@@
   \caption{\rev{GPU hazard ratios from Cox regression model on {\em
-        old} batch. Putting columns in logical torus order removes the
-      peculiar interleaving pattern and presents a more physically
-      reasonable response. This suggests that scheduling job fill from
-      X-0 had measurable temperature impact on the cooler side of
-      Titan (evaporators get engaged only at a temperature
-      threshold).}}
+        old} batch, with the {\em col} variable in logical
+      X-coordinate torus order. The torus order removes the peculiar
+      interleaving pattern in Fig.~\ref{fig:cox-old} and presents a
+      response that can be explained by a combination of system
+      cooling and job scheduling.}}
   \label{fig:cox-torus}
-\end{figure}
-\begin{figure}
-  \centering
-  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
-  \caption{GPU hazard ratios from Cox regression model on {\em new}
-    batch.}
-  \label{fig:cox-new}
 \end{figure}
@@
 and~\ref{fig:cox-new}.
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
+  \caption{GPU hazard ratios from Cox regression model on {\em new}
+    batch.}
+  \label{fig:cox-new}
+\end{figure}
 

### 2020-06-03 code/TitanGPUmodel.Rmd
Reviewer 1 done

@@
              year = factor(as.character(year(dates_life[-1])))) %>% filter(days >= 1) %>% 
-  mutate(days = factor(as.integer(floor(days))))
+  mutate(days = factor(as.integer(floor(days)), levels = 1:56))
 ggp(ggplot(life_intervals, aes(days, fill = year)) + stat_count() +
-      theme_bw(), file = "attention_intervals", height = 3)
+      scale_x_discrete(breaks = seq(1, 56, 2), drop = FALSE) + theme_bw(),
+    file = "attention_intervals", height = 3)
 ```

### 2020-06-03 dataprep.tex
Reviewer 1 done

@@
 failure events were collected \rev{for this study}: Double Bit Error
-(DBE) and Off the Bus (OTB). \rev{While other types of errors were
-  recorded in log files, as is reported in \cite{7266836}, these two
-  kinds of errors became dominant in 2016 and were found to be the
-  ``signature'' event of the GPU board failing resistor and a trigger
-  for GPU replacement.}
+(DBE) and Off the Bus (OTB). \rev{DBE is a memory error detected by
+  parity bit checking, which can correct a single bit flip and detect
+  but not correct a double bit flip. OTB is the loss of data swim lane
+  connections to the GPU. While other types of errors were recorded in
+  log files, as is reported in \cite{7266836}, OTB and DBE became
+  dominant in 2016 and were found to be the ``signature'' event of the
+  GPU board failing resistor and a trigger for GPU replacement.}
 
-An inventory of GPU serial numbers and their locations is recorded
-each time the system boots. This typically occurs once every few days
-but can be as frequent as a few times per day. A weekly inventory was
-later added to run independently of system boots when they were fewer
-and farther apart. A single inventory takes about a minute and is
-recorded in a separate file. \rev{An inventory can be incomplete if a
-  node response times out but such occurrences were relatively rare}.
-Figure~\ref{fig:inventory} shows inter-inventory times that appear in
-the data.
+An inventory of GPU serial numbers and their locations was recorded
+each time the system boots, \rev{typically for software updates and
+  patches but sometimes for hardware swaps, although warm swaps of
+  blades were usually possible.} Boots typically occured once every
+few days but sometimes even a few times per day. A single inventory
+takes about a minute and is recorded in a separate file. \rev{It can
+  be incomplete if a node response times out but such occurrences were
+  relatively rare.
 \begin{figure}[bt]
@@
 \end{figure}
-Notably, there was a period of nearly two months and many periods of
-over a week that Titan operated without a single reboot in the
-2014-2015 time frame. \rev{However, by the time the weekly inventories
-  were added, failure rates started to increase and lead to more
-  frequent reboots and thus inventories.}
+Figure~\ref{fig:inventory} shows that during 2014 and 2015,
+inventories were far apart, even 56 days once in 2014. The purpose is
+to illustrate that exact times for GPU removal from service are
+``censored'' and we only get the inventory time information. We discuss
+censoring and its implications in Sec.~\ref{section:survival}.}
 

### 2020-06-03 survival.tex
Reviewer 1 done

@@
 Survival analysis (SA) methods
-\cite{survival-book}, 
+\rev{\cite{survival-book}}, 
 sometimes referred to as time to event methods, focus on the effect of
@@
 Kaplan-Meier survival analysis (KM)
-\cite{kaplan-meier1958,survival-book} starts with computing the
+\rev{\cite{kaplan-meier1958,survival-book}} starts with computing the
 probability of survival beyond a given time. It is a nonparametric
@@
       slot}, and {\em node} are with respect to spatial locations
-    shown in Fig.~\ref{fig:layout}}}
+    shown in Fig.~\ref{fig:layout}. Notably, the {\em col} are
+    physical columns of cabinets.}}
   \label{fig:cox-old}
@@
       interleaving pattern in Fig.~\ref{fig:cox-old} and presents a
-      response that can be explained by a combination of system
+      response that can be explained by an interaction of system
       cooling and job scheduling.}}
@@
 risk averaged over time based on the covariates
-\cite{Cox1972,Harrell2015}. The CPH regression function takes the form
+\rev{\cite{Cox1972,Harrell2015}}. The CPH regression function takes the form
 \begin{displaymath}

### 2020-06-03 abstract.tex
added missing pdfs

@@
 operation was forced to execute three very significant rework cycles,
-two on the GPU mechanical assembly and one on the GPU circuitboards. We
-write about the last rework cycle and a reliability analysis of over
-100,000 operation years in the GPU lifetimes, which correspond to
-Titan's 6 year long productive period after an initial break-in
-period. Using time between failures analysis and statistical survival
-analysis techniques, we find that GPU reliability is dependent on heat
-dissipation to an extent that strongly correlates with detailed
-nuances of the system cooling architecture. In addition to describing
-some of the system history, the data collection, data cleaning, and
-our analysis of the data, we provide reliability recommendations for
-designing future state of the art supercomputing systems and their
-operation. We make the data and our analysis codes publicly available.
+two on the GPU mechanical assembly and one on the GPU
+circuitboards. We write about the last rework cycle and a reliability
+analysis of over 100,000 operation years in the GPU lifetimes, which
+correspond to Titan's 6 year long productive period after an initial
+break-in period. Using time between failures analysis and statistical
+survival analysis techniques, we find that GPU reliability is
+dependent on heat dissipation to an extent that strongly correlates
+with detailed nuances of the system cooling architecture \rev{and job
+scheduling}. In addition to describing some of the system history, the
+data collection, data cleaning, and our analysis of the data, we
+provide reliability recommendations for designing future state of the
+art supercomputing systems and their operation. We make the data and
+our analysis codes publicly available.

### 2020-06-03 discussion.tex
Added text about monitoring improvements.

@@
 drops by an order of magnitude.
+
+\rev{The experienced reliability issues with Titan had a direct impact on the
+development and deployment of Titan's successor, Summit, and even on Summit's
+successor, Frontier. As a result, more and better monitoring data is already
+collected on Summit. NVIDIA's GPU management and monitoring software has been
+significantly improved. Temperature monitoring has been significantly improved
+as well.}

### 2020-06-03 code/TitanGPUmodel.Rmd
updated Rmd and added silver corrosion ref

@@
 ```{r setup, include=FALSE}
+options(repos = c(CRAN = "http://cran.rstudio.com")) # set CRAN mirror
+install.packages("remotes")
+remotes::install_github("NikNakk/forestmodel") # only package from GitHub
+## all other packages will prompt to install from CRAN
+
 out_dir = "../figs/" # paper related figures will go here

### 2020-06-03 dataprep.tex
updated Rmd and added silver corrosion ref

@@
   \caption{\rev{Titan cabinet (front view and side view with air flow)
-      and single blade layout (bottom left) and floor layout (bottom
-      right). Cooling airflow is bottom to top within a cabinet,
+      and single blade layout (bottom left with {\em node} numbers)
+      and floor layout (bottom right with {\em col} and {\em row}
+      numbers). Cooling airflow is bottom to top within a cabinet,
       taking ambient air in mostly at the foot and relasing it back
@@
       evaporators, below Cage 0 and above Cage 2. Ambient air in the
-      room is conditioned at nominal 75$^\circ$~F. This system has a
-      dozen intakes near the ceiling and dumps cool air under the
-      floor in several locations. A number of floor tiles are
-      perforated.  Yellow dot marks GPU location {\tt c17-4c1s3n1}}}
+      room is conditioned at nominal 75$^\circ$~F. The room air system
+      has a dozen intakes near the ceiling and dumps cool air under
+      the floor in several locations. A number of floor tiles are
+      perforated.  Yellow dot marks GPU location {\tt c17-4c1s3n1}.}}
   \label{fig:layout}
@@
 
-The first GPU record in \rev{Fig.~\ref{fig:dataraw}}
-shows installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0},
-{\tt c0-1c1s3n3}, with periods off the system, and finally in {\tt
+The first GPU record in \rev{Fig.~\ref{fig:dataraw}} shows
+installation in locations {\tt c17-4c1s3n1}, {\tt c21-1c2s5n0}, {\tt
+  c0-1c1s3n3}, with periods off the system, and finally in {\tt
   c13-1c1s3n3}, where it stays until August 1, 2017, after which it is
@@
   c23-6c1s7n2}, where a ``DBE'' is observed on November 2, 2018, and
-it is not seen again.
+it is not seen again. \rev{This is the data set we make publicly
+  available.}
 
-% \fix{Where does cool air arrive? Is it just dumped out at the top of
-%   the heat exchanger and circulate down? I noticed airholes in some
-%   floor tiles (more on the right than on the left). Is there cool air
-%   under the floor?}
-
-% \fix{How does the coolant circulate? I notice three sets of pipes on
-%   each side of each row: (2 x 3) x 8 = 48 sets.}
+% I noticed more perforated floor tiles on the right than on the left
+% in the Titan decomissioning disassembly video. Anecdotally, the
+% right side of the machine was hotter than the left so some
+% perforated floor tiles were moved to the right.

### 2020-06-04 reviews/response.txt
done today

We thank all the reviewers for their exceptionally helpful and
thorough comments. Thorough reviews take time and interest! We have
incorporated suggested changes in our revised version and marked the
revised text in blue. Here, we respond to reviewers' "Comments for
Revision" and due to limited space explain mostly only those detailed
comments that are not marked or incorporated in our revised paper
version.

Overall, there were two substantive additions in addition to numerous
smaller cosmetic and presentation improvements. The first is a
corroboration of our airflow implications on temperature with data
from another publication. The second is finding a contributing factor
to temperature in job scheduling

### 2020-06-04 dataclean.tex
done for today
done today

@@
 \begin{figure*}[bt]
-  \includegraphics[width=6.5in]{figs/sample_sn.pdf}
+  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf}
   \caption{Serial number view of GPU life and failures. Legend: black
@@
 \begin{figure*}[tb]
-  \includegraphics[width=6.5in]{figs/sample_loc.pdf}
+  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf}
   \caption{Location view of GPU life and failures.  Legend: black dots
@@
 done by removing data for any units with a {\em remove} date before
-2014. After this reduction, there were still six older units remaining,
-which we also removed to have a clean set of units for the
+2014. After this reduction, there were still six older units
+remaining, which we also removed to have a clean set of units for the
 analysis. We do some further processing to handle time overlaps in a
@@
 of recorded life in the raw records by simply dropping the overlapping
-life times.
+life times. \rev{A detailed followup of a few overlaps found that they
+  are caused by incomplete inventories, where a node query times out,
+  creating the appearance that a unit is in two locations.}
 

### 2020-06-04 dataprep.tex
done today

@@
   parity bit checking, which can correct a single bit flip and detect
-  but not correct a double bit flip. OTB is the loss of data swim lane
-  connections to the GPU. While other types of errors were recorded in
-  log files, as is reported in \cite{7266836}, OTB and DBE became
-  dominant in 2016 and were found to be the ``signature'' event of the
-  GPU board failing resistor and a trigger for GPU replacement.}
+  but not correct a double bit flip. OTB is the loss of host CPU
+  connection to the GPU. While other types of errors were recorded in
+  log files, as is reported in \cite{7266836,tiwari15Understand}, OTB
+  and DBE became dominant in 2016 and were found to be the
+  ``signature'' event of the GPU board failing resistor and a trigger
+  for GPU replacement.}
 

### 2020-06-04 related.tex
done today

@@
 
-Tiwari et al~\cite{Tiwari15Experience,7056044} analyze types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
+Tiwari et al.~\cite{Tiwari15Experience,tiwari15Understand} analyze types of GPU errors on Titan and on a GPU cluster at Los Alamos National Laboratory.
 This also includes neutron beam experiments at the Los Alamos Neutron Science Center and at Rutherford Appleton Laboratories to further understand GPU soft errors caused by neutron radiation triggered upsets (bit flips and timing errors).
 %
-Tiwari et al~\cite{10.1145/2807591.2807666} further analyze types of GPU errors on Titan, including software/firmware related errors and failures, ECC double-bit errors, and GPU ``off the bus'' and ECC page retirement events.
+Tiwari et al.~\cite{10.1145/2807591.2807666} further analyze types of GPU errors on Titan, including software/firmware related errors and failures, ECC double-bit errors, and GPU ``off the bus'' and ECC page retirement events.
 This work observes an ECC double-bit mean-time between errors (MTBE) of about 160 hours, or about one per week, with 86\% occurring in GPU memory and the rest in the GPU register file.

### 2020-06-04 survival.tex
updated Rmd and added silver corrosion ref
done today

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}
-  \centering
-  \includegraphics[width=7in]{figs/km_cage-node_a001.pdf}
-  \caption{Comparison of the old and new batches, including survival
-    differences based on {\tt cage} and {\tt node} GPU locations.}
-  \label{fig:km-all-cage-node}
-\end{figure*}
 \section{Survival Analysis}
@@
 analysis community, will dive deeper.
+\begin{figure*}
+  \centering
+  \includegraphics[width=7in]{figs/km_cage-node_a001.pdf}
+  \caption{Comparison of the old and new batches, including survival
+    differences based on {\tt cage} and {\tt node} GPU locations.}
+  \label{fig:km-all-cage-node}
+\end{figure*}
 
@@
 % Inserting figure here so it appears at top of next page
-\begin{figure}
+\begin{figure}[bt]
   \centering
@@
   \centering
+  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
+  \caption{GPU hazard ratios from Cox regression model on {\em new}
+    batch.}
+  \label{fig:cox-new}
+\end{figure}
+\begin{figure}[bt]
+  \centering
   \includegraphics[width=\columnwidth]{figs/cox_o_t001.pdf}
   \caption{\rev{GPU hazard ratios from Cox regression model on {\em
-        old} batch, with the {\em col} variable in logical
-      X-coordinate torus order. The torus order removes the peculiar
+        old} batch, with the {\em col} variable in interconnect torus
+      X-coordinate order. The torus order removes the peculiar
       interleaving pattern in Fig.~\ref{fig:cox-old} and presents a
@@
 experience the same relationship with distance to the bottom, nodes 2
-and 3 having lower failure rates than nodes 0 and 1.
+and 3 having lower failure rates than nodes 0 and 1. \rev{An
+  increasing temperature gradient averaging about $4^\circ$C per
+  cage is reported in \cite[p.7]{tiwari15Understand}, independently
+  confirming our airflow temperature interpretation.}
 
@@
 technique that in a sense averages over time. Cox proportional hazards
-(CPH) regression analysis, can include covariates and estimates relative
-risk averaged over time based on the covariates
-\rev{\cite{Cox1972,Harrell2015}}. The CPH regression function takes the form
+(CPH) regression analysis, can include covariates and estimates
+relative risk averaged over time based on the covariates
+\rev{\cite{Cox1972,Harrell2015}}. The CPH regression function takes
+the form
 \begin{displaymath}
@@
 do cross and so we fit the model to the new and the old batches
-separately. The results are presented in \rev{Fig.}~\ref{fig:cox-old}
+separately. The results are presented in \rev{Fig.}~\ref{fig:cox-old},
 and~\ref{fig:cox-new}.
-\begin{figure}
-  \centering
-  \includegraphics[width=\columnwidth]{figs/cox_n001.pdf}
-  \caption{GPU hazard ratios from Cox regression model on {\em new}
-    batch.}
-  \label{fig:cox-new}
-\end{figure}
 
 We find that the average hazard ratios strongly correlate with
-detailed nuances of the system cooling architecture.  The figures give
+detailed nuances of the system cooling architecture \rev{and are also
+  impacted by job scheduling}.  The figures give
 hazard ratios for each of the location variables, giving the ratio for
@@
 aspect is not captured by the uncertainty, which accounts for the
-randomness of the failures and not for the geometric time averaging
+randomness of the failures but not for the geometric time averaging
 \rev{\cite{Hernn2010TheHO}}. Consequently, we interpret the pattern of
@@
 different behavior in {\em col} 0-11 from {\em col} 12-24 is likely
-due to separate supply lines of chilled water to the two sets. While
-the temperature was a consistent 42$^\circ$F, there may have been
-minor flow or pressure differences. But the interleaving pattern of
-the first set is still a mystery. We also have yet to explain the
-apparent row effects. A very minor airflow effect across slots also
-seems to be present as it mimics faster airflow for the middle slots.
+\rev{due to the presence of the file system to the right of {\em col}
+  24, as can be seen in the floor layout of Fig.~\ref{fig:layout}. An
+  explanation due to separate supply lines of chilled water to the two
+  sets is not likely as the evaporators (see Fig.~\ref{fig:layout})
+  are thermostat controlled.}
+
+\rev{The peculiar interleaving pattern of the first set of columns can
+  be explained by Titan's folded torus interconnect. If we order the
+  columns by their torus X coordinate \cite{ezell2013}, seen in
+  Fig.~\ref{fig:cox-torus}, the column pattern retains the proximity
+  to the file system explanation and loses the interleaving pattern
+  peculiarity. Moreover, Titan filled temporal and spatial gaps in
+  nodes starting with low torus {\em X} coordinate. This potentially
+  explains that the low {\em col} and low {\em X} coordinate columns
+  have higher hazard due to higher workload than the low {\em col} and
+  high {\em X} coordinate columns.  It is possible that the torus Y
+  coordinate can help explain the apparent row effects, but we have
+  difficulty interpreting the row description in \cite{ezell2013}.} A
+very minor airflow effect across slots also seems to be present as it
+mimics faster airflow for the middle slots.

### 2020-06-04 code/TitanGPUmodel.Rmd
testing reproducibility

@@
 options(repos = c(CRAN = "http://cran.rstudio.com")) # set CRAN mirror
-install.packages("remotes")
+installed = installed.packages()
+if(! "remotes" %in% installed) install.packages("remotes")
+if(! ("forestmodel" %in% installed && packageVersion("forestmodel") >= "0.6.1"))
   remotes::install_github("NikNakk/forestmodel") # only package from GitHub

### 2020-06-04 code/TitanGPUmodel.Rmd
reproducibility

@@
 ## all other packages will prompt to install from CRAN
+library(data.table)
+library(magrittr)
+library(dplyr)
+library(ggplot2)
+library(forcats)
+library(lubridate)
+library(survival)
+library(survminer)
+library(forestmodel)
 
@@
 ```
-`r today()`
+`r`
 Read the data. Last inventory was run 01/20/2020 but Titan was turned off on 08/01/2019
@@
 life_sample = gc_life %>% filter(location %in% loc_sample) %>% 
-  separate(location, c(NA, "col", "row", "cage", "slot", "node"),
+  tidyr::separate(location, c(NA, "col", "row", "cage", "slot", "node"),
            sep = "[-csn]", convert = TRUE, remove = FALSE) %>%
@@
 gc_write = as.data.frame(lapply(gc_full, as.character), stringsAsFactors = FALSE)
-write_csv(gc_write, paste0(out_dir, "gc_full.csv"), na = "", col_names = TRUE)
+readr::write_csv(gc_write, paste0(out_dir, "gc_full.csv"), na = "", col_names = TRUE)
 ```
@@
 ```{r}
-gc_summary_loc = gc_summary %>% separate(max_loc, c(NA, "col", "row", "cage", "slot", "node"),
+gc_summary_loc = gc_summary %>% tidyr::separate(max_loc, c(NA, "col", "row", "cage", "slot", "node"),
            sep = "[-csn]", convert = TRUE, remove = TRUE) %>% # c{col}-{row}c{cage}s{slot}n{node}
@@
 gc_write = as.data.frame(lapply(gc_summary_loc, as.character), stringsAsFactors = FALSE)
-write_csv(gc_write, paste0(out_dir, "gc_summary_loc.csv"), na = "", col_names = TRUE)
+readr::write_csv(gc_write, paste0(out_dir, "gc_summary_loc.csv"), na = "", col_names = TRUE)
 ```

### 2020-06-04 code/TitanGPUsetup.R

@@ -3,9 +3,9 @@
 knitr::opts_chunk$set(echo = TRUE)
 options(width = 150)
 
-suppressMessages(library(data.table))
-suppressMessages(library(tidyverse))
-suppressMessages(library(lubridate))
+#suppressMessages(library(data.table))
+#suppressMessages(library(tidyverse))
+#suppressMessages(library(lubridate))
 
 ## Function to mark overlapping life spans in unit (sn or location). Adds two
 ## variables: overlap_unit and overlap_rec, marking all records in a unit and

### 2020-06-04 code/TitanGPUmodel.Rmd
reproducibility fix and added data

@@
 
-out_dir = "../figs/" # paper related figures will go here
-in_dir = "../data/" # input files live here
-history_file = paste0(in_dir, "titan.gpu.history.txt")
-service_nodes = paste0(in_dir, "titan.service.txt")
+out_figs = "../figs/" # paper related figures will go here
+data_dir = "../data/" # input files live here
+history_file = paste0(data_dir, "titan.gpu.history.txt")
+service_nodes = paste0(data_dir, "titan.service.txt")
 source("TitanGPUsetup.R")
@@
 gc_write = as.data.frame(lapply(gc_full, as.character), stringsAsFactors = FALSE)
-readr::write_csv(gc_write, paste0(out_dir, "gc_full.csv"), na = "", col_names = TRUE)
+readr::write_csv(gc_write, paste0(data_dir, "gc_full.csv"), na = "", col_names = TRUE)
 ```
@@
 gc_write = as.data.frame(lapply(gc_summary_loc, as.character), stringsAsFactors = FALSE)
-readr::write_csv(gc_write, paste0(out_dir, "gc_summary_loc.csv"), na = "", col_names = TRUE)
+readr::write_csv(gc_write, paste0(data_dir, "gc_summary_loc.csv"), na = "", col_names = TRUE)
 ```
@@
 ```{r fig.height = 8}
-library(survival)
-library(survminer)
 ggp(ggsurvplot(survfit(
@@
 col2torus = order(c(0, seq(1, 23, 2), seq(24, 2, -2)))
-library(forestmodel)
 panels = list(

### 2020-06-04 code/TitanGPUsetup.R

@@ -3,10 +3,6 @@
 knitr::opts_chunk$set(echo = TRUE)
 options(width = 150)
 
-#suppressMessages(library(data.table))
-#suppressMessages(library(tidyverse))
-#suppressMessages(library(lubridate))
-
 ## Function to mark overlapping life spans in unit (sn or location). Adds two
 ## variables: overlap_unit and overlap_rec, marking all records in a unit and
 ## specific overlapping record in a unit, respectively.
@@ -69,7 +65,7 @@ plot.life_ev = function(glife, gevents, unit, file = FALSE, overlaps = TRUE,
                        axis.title.x = element_blank())
 
   if(is.character(file)) {
-    pdf(paste0(out_dir, file, ".pdf"),
+    pdf(paste0(out_figs, file, ".pdf"),
       height = as.integer(nu/18 + 0.1, width = 6))
     print(p)
     dev.off()
@@ -82,7 +78,7 @@ plot.life_ev = function(glife, gevents, unit, file = FALSE, overlaps = TRUE,
 ## blanks replaced by underscores and .pdf appended. Also echoes to console 
 ## print by default - optional.
 ggp = function(p, width = 7, height = 7, file = "gpp", echo = TRUE) {
-  pdf(paste0(out_dir, file, "%03d.pdf"), width, height,
+  pdf(paste0(out_figs, file, "%03d.pdf"), width, height,
       onefile = FALSE)
   print(p)
   dev.off()

### 2020-06-04 related.tex
fixed reference

@@
 
-Ezell~\cite{osti_1086655} creates a general understanding of Titan's interconnect failures using an application to stress test Titan's Gemini interconnect.
+Ezell~\cite{ezell2013} creates a general understanding of Titan's interconnect failures using an application to stress test Titan's Gemini interconnect.
 %

### 2020-06-04 dataclean.tex
combine view figs

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}[bt]
-  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf}
-  \caption{Serial number view of GPU life and failures. Legend: black
-    dots are installs, black lines are lifetimes at installed
-    location, blue squares are OTB events, red triangles are DBE
-    events, and black ] are ``last seen'' events.}
-  \label{fig:gpuview}
-\end{figure*}
-\begin{figure*}[tb]
-  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf}
-  \caption{Location view of GPU life and failures.  Legend: black dots
-    are installs, black lines are lifetimes at installed location,
-    blue squares are OTB events, red triangles are DBE events, and
-    black ] are ``last seen'' events.}
-  \label{fig:locview}
-\end{figure*}
 \section{Viewing and Cleaning the Data}
@@
 To get some intuition for the GPU lifetimes on Titan, we give two
-views of 60 randomly selected GPUs (\rev{Fig.}~\ref{fig:gpuview}) and 80
-randomly selected locations (\rev{Fig.}~\ref{fig:locview}).
+views of \rev{80} randomly selected GPUs and \rev{80} randomly
+selected locations (\rev{Fig.}~\ref{fig:gpu-view}).
+\begin{figure*}[p]
+  \begin{center}
+  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf} \\
+  (a) \\[1em]
+  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf} \\
+  (b)
+  \end{center}
+  \caption{\rev{Serial number view (a) and location view (b) of GPU
+      life and failures. Both SN and locations are randomly
+      selected. Legend: black dots are installs, black lines are
+      lifetimes at installed location, blue squares are OTB events,
+      red triangles are DBE events, and black ] are ``last seen''
+      events.}}
+  \label{fig:gpu-view}
+\end{figure*}
 The GPU view
@@
 The third rework cycle was used to label GPUs as {\em old} batch and
-{\em new} batch. The two batches are clearly identifiable in the GPU
-view of Fig.~\ref{fig:gpuview} as starting near 2017 timeframe. More
-frequent OTB and DBE events are apparent in the old batch. It is also
-clear from this view that practically all new GPUs stayed at their
-initial install location whereas the old GPUs were occasionally
-reinstalled at new locations. Nevertheless, a separate analysis
-determined that the vast majority of time of the vast majority of
-units is spent at one location by both the new and the old units.
+{\em new} batch. The two batches are clearly identifiable in the
+\rev{SN} view of Fig.~\ref{fig:gpu-view} as those starting near 2017
+timeframe. More frequent OTB and DBE events are apparent in the old
+batch. It is also clear from this view that practically all new GPUs
+stayed at their initial install location whereas the old GPUs were
+occasionally reinstalled at new locations. Nevertheless, a separate
+analysis determined that the vast majority of time of the vast
+majority of units is spent at one location by both the new and the old
+units.
 

### 2020-06-04 intro.tex
updated Rmd and added silver corrosion ref
combine view figs

@@
 many technological boundaries. \rev{This early period included
-  extensive field engineering and experimentation to determine cause,
-  down times, and large temporal and completeness gaps in inventory
-  runs. We include the early swap data in this figure only for
-  completeness and to underscore the amount of work required to field
-  a world's largest supercomputer. However, we exclude this period
-  from further analysis because we have a gap in explanation for one
-  of the swap spikes and because the separation is made possible by
-  the second, late 2013, GPU rework cycle.} Late 2013 begins a very
-long period of high reliability and stable operation with very few
-issues until about mid 2016, when GPU failures begin to rise. This
-results in the final rework cycle, replacing 11,000 GPUs throughout
-much of 2017.
+  extensive field engineering and experimentation, down times, as well
+  as temporal and completeness gaps in inventory runs due to down
+  times. We include the early swap data in this figure only for
+  completeness and to underscore the massive amount of hardware work
+  required to field a world's largest supercomputer. However, as we
+  focus on GPU operation after the second rework cycle we exclude this
+  early period from further analysis.} Late 2013 begins a very long
+period of high reliability and stable operation with very few issues
+until about mid 2016, when GPU failures begin to rise. This results in
+the final rework cycle, replacing 11,000 GPUs throughout much of 2017.
 
 Finding the root cause for the 2016 failures took a great deal of
-testing and involvement of materials science and microscopy
-researchers. The failures were traced to a resistor on the GPU
-circuit board (not the GPU chip iself). Throughout this paper, when we
-refer to replacing the GPU, we refer to the entire circuitboard along
-with its GPU chip.
+testing and the involvement of materials science and microscopy
+researchers. The failures were traced to a resistor on the GPU circuit
+board (not the GPU chip iself) \fix{Was anything published about this?
+  If not, let's use \cite{Vargas2009TheCO}}. Throughout this paper,
+  when we refer to replacing the GPU, we refer to the entire
+  circuitboard along with its GPU chip.
 

### 2020-06-04 tbf.tex
combine view figs

@@
 which were frequent particularly during early 2017 as can be seen in
-Fig.~\ref{fig:locview}.
+the location view of Fig.~\ref{fig:gpu-view}.
 

### 2020-06-04 dataclean.tex
figure placement

--- 2020-06-04 dataclean.tex
@@
 To get some intuition for the GPU lifetimes on Titan, we give two
-views of 60 randomly selected GPUs (\rev{Fig.}~\ref{fig:gpuview}) and 80
-randomly selected locations (\rev{Fig.}~\ref{fig:locview}).
+views of \rev{80} randomly selected GPUs and \rev{80} randomly
+selected locations \rev{in Fig.~\ref{fig:gpu-view}}.
 The GPU view

### 2020-06-04 tbf.tex
figure placement

@@
 \label{section:tbf}
+
+\begin{figure*}[p]
+  \begin{center}
+  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf} \\
+  (a) \\[1em]
+  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf} \\
+  (b)
+  \end{center}
+  \caption{\rev{Serial number view (a) and location view (b) of GPU
+      life and failures. Both SN and locations are randomly
+      selected. Legend: black dots are installs, black lines are
+      lifetimes at installed location, blue squares are OTB events,
+      red triangles are DBE events, and black ] are ``last seen''
+      events.}}
+  \label{fig:gpu-view}
+\end{figure*}
+Using the cleaned data, we analyze the inter-arrival times between the DBE and OTB events. 
+This analysis is done at the device level and at the system level. 
+It provides important insights into the reliability of large-scale machines, where the 
+failure rate of an individual device is significantly different from the overall reliability
+of the machine.
 
@@
 \end{figure}
-
-Using the cleaned data, we analyze the inter-arrival times between the DBE and OTB events. 
-This analysis is done at the device level and at the system level. 
-It provides important insights into the reliability of large-scale machines, where the 
-failure rate of an individual device is significantly different from the overall reliability
-of the machine.
-
 A histogram of MTBFs measured across GPUs which had at least one failure event is shown 

### 2020-06-04 dataclean.tex
update views fig

@@
 To get some intuition for the GPU lifetimes on Titan, we give two
-views of \rev{80} randomly selected GPUs and \rev{80} randomly
-selected locations \rev{in Fig.~\ref{fig:gpu-view}}.
-The GPU view
-visually documents the life of each GPU unit: when it was installed and
-removed at various locations, its DBE and OTB events, and the last
+views of \rev{90} randomly selected GPUs and \rev{90} randomly
+selected locations \rev{in Fig.~\ref{fig:gpu-view}}.  The GPU view
+visually documents the life of each GPU unit: when it was installed
+and removed at various locations, its DBE and OTB events, and the last
 time it was seen. The location view documents the life of a location:

### 2020-06-04 dataprep.tex
update views fig

@@
 respect to the layout shown in \rev{Fig.}~\ref{fig:layout}. \rev{Note
-  that there is an extra empty cabinet between columns 10 and 11 to
-  accommodate a few ceiling support columns. For
-  example, we mark location {\tt c17-4c1s3n1} with a yellow dot in
-  each view of the figure.}
+  that there is an extra empty column of cabinets between columns 10
+  and 11 to accommodate a few ceiling support columns. To aid
+  orientation in the figure, we mark an example location {\tt
+    c17-4c1s3n1} with a yellow dot in each view of the figure.}
 \begin{figure}

### 2020-06-04 tbf.tex
update views fig

@@
 
+% figure put here for better placement in compiled document
 \begin{figure*}[p]
   \begin{center}
-  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf} \\
-  (a) \\[1em]
-  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf} \\
-  (b)
+    \raisebox{-.5\height}{
+      \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf}
+     }(a) \\[1em]
+     \raisebox{-.5\height}{
+       \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf}
+     }(b)
   \end{center}
@@
       life and failures. Both SN and locations are randomly
-      selected. Legend: black dots are installs, black lines are
-      lifetimes at installed location, blue squares are OTB events,
-      red triangles are DBE events, and black ] are ``last seen''
-      events.}}
+      selected. Black dots are installs, black lines are lifetimes at
+      installed location, blue squares are OTB events, red triangles
+      are DBE events, and black ] are ``last seen'' events. Such views
+      were critical to understanding the data and to verifying various
+      data processing decisions.}}
   \label{fig:gpu-view}

### 2020-06-05 reviews/response.txt
corrosion sentence in intro

@@
-We thank all the reviewers for their exceptionally helpful and
-thorough comments. Thorough reviews take time and interest! We have
-incorporated suggested changes in our revised version and marked the
-revised text in blue. Here, we respond to reviewers' "Comments for
-Revision" and due to limited space explain mostly only those detailed
-comments that are not marked or incorporated in our revised paper
-version.
+We thank the reviewers for their exceptionally helpful and thorough
+comments. Thorough reviews take time, interest, and expertise! Here,
+we respond to reviewers' "Comments for Revision", marked by (C), and
+due to limited space explain only a selection of detailed comments,
+marked by (D). However, we have addressed all of the reviewers
+comments in our revised paper either by making the requested change or
+by providing additional explanation.
 
-Overall, there were two substantive additions in addition to numerous
-smaller cosmetic and presentation improvements. The first is a
-corroboration of our airflow implications on temperature with data
-from another publication. The second is finding a contributing factor
-to temperature in job scheduling
+Our revision includes two substantive additions and numerous
+improvements to explanations, graphics, and presentation. The first
+substantive addition is a corroboration of our airflow implications on
+temperature with temperature data from another publication. The second
+substantive addition is a new result that job scheduling and folded
+torus interconnect were contributing factors to temperature patterns,
+and explain a peculiar interleaving column hazard variability.
+
+Reviewer 1:
+
+(C) Figure 2 is improved and corrected. The inventory dates are the same
+for old and new batches, except that new units begin to appear in 2016 and
+this is evident from the year colors.
+
+Figure 7 is now split between old and new batches.
+
+We now have location-specific information for average temperature
+across cages and it agrees with our airflow effect on temperature
+interpretation (V.-6th paragraph).
+
+(D)
+
+Reviewer 2:
+
+(C) We have strengthened our explanations

### 2020-06-05 intro.tex
corrosion sentence in intro

@@
 researchers. The failures were traced to a resistor on the GPU circuit
-board (not the GPU chip iself) \fix{Was anything published about this?
-  If not, let's use \cite{Vargas2009TheCO}}. Throughout this paper,
-  when we refer to replacing the GPU, we refer to the entire
-  circuitboard along with its GPU chip.
+board (not the GPU chip iself) due to silver sulphide
+corrosion. \fix{Was anything published about this?  If not, let's use:
+  Growth of such corrosion products in ambient air on microelectroncs
+  parts is described in \cite{Vargas2009TheCO} and is cosistent with
+  failures starting only after a critical amount of corrosion builds
+  up, a situation that matches the experience on Titan}.
 
-Our analysis focuses on GPU boards that were installed in the second
-rework cycle and later, essentially units installed near the beginning
-of 2014 and later. Figure~\ref{fig:chronology} is the only one
-involving data prior to 2014. All analyses and figures that follow use
-data from 2014 onward. We make the data as well as our analysis R
-codes and Python codes to reproduce the graphics in this paper
-publicly available (see, \cite[repository to be added in final
+Throughout this paper, when we refer to replacing the GPU, we refer to
+the entire circuitboard along with its GPU chip. Our analysis focuses
+on GPU boards that were installed in the second rework cycle and
+later, essentially units installed near the beginning of 2014 and
+later. Figure~\ref{fig:chronology} is the only one involving data
+prior to 2014. All analyses and figures that follow use data from 2014
+onward. We make the data as well as our analysis R codes and Python
+codes to reproduce the graphics in this paper publicly available (see,
+\cite[repository to be added in final
 version]{repositoryreference,rcodes,pycodes}).  While the data is only

### 2020-06-05 survival.tex
corrosion sentence in intro

@@
 experience the same relationship with distance to the bottom, nodes 2
-and 3 having lower failure rates than nodes 0 and 1. \rev{An
-  increasing temperature gradient averaging about $4^\circ$C per
-  cage is reported in \cite[p.7]{tiwari15Understand}, independently
-  confirming our airflow temperature interpretation.}
+and 3 having lower failure rates than nodes 0 and 1.
+
+\rev{An increasing average temperature gradient of about $4^\circ$C
+  per cage is reported in \cite[Fig.10]{tiwari15Understand},
+  independently confirming our airflow temperature
+  interpretation. They also suggest that DBEs may be sensitive to
+  temperature, although consider their evidence preliminary due to
+  relatively low number of DBEs in 2014-1015 and high variability of
+  $10^\circ$C to $15^\circ$C in their measured cage replicates.}
 

### 2020-06-05 code/tbf-analyses/tbf_analyses.py
python code to perform time-between-failure analyses on Titan log data

@@ -1,0 +1,932 @@
+#### This code was used for time-between-failure analysis in the following paper (pending acceptance):
+## Authors list (TBD, hidden for review), ``GPU Lifetimes on Titan Supercomputer: Survival Analysis and Reliability,'' 
+## The International Conference for High Performance Computing, Networking, Storage, and Analysis (SC20),
+## Atlanta, GA, Nov. 15-20, 2020.
+## ...
+#### If using this code or this work, please cite the above paper in your work.
+
+
+#### This code reads in a csv file with GPU failure data and does the following:
+## 1. GPU-wise mean-time-between-failure analyses (see Fig-6 in paper).
+## 2. System-wide mean-time-between-failure analyses over lifetime (see Fig-7 through 9 in paper).
+####
+
+#### last modified: 2020-06-04 ##################################################################
+
+#### package imports ############################################################################
+## for data parsing 
+import csv
+import time
+from datetime import datetime
+import re
+
+## for plots and fitting
+import numpy as np
+import matplotlib.pyplot as plt
+
+#### helper functions for parsing temporal data #################################################
+## takes time string input and convert to epoch
+def epoch(timestring):
+    return int(time.mktime(time.strptime(timestring, "%Y-%m-%d %H:%M:%S")))
+
+def convertToTime(timestring):  
+    return time.strptime(timestring, "%Y-%m-%d %H:%M:%S")
+
+#### Parse failure data, while recording failure times ##########################################
+
+# for time-wise breakdown of TBF
+ALL_RAW_DBE_DATETIMES = [] # includes both new and old
+ALL_RAW_OTB_DATETIMES = []
+
+ALL_DBE_EPOCHS = []
+ALL_OTB_EPOCHS = []
+
+ALL_RAW_DBE_dict = {} # these are later used to form old/new sets
+ALL_RAW_OTB_dict = {}
+
+# GPUwise TBF 
+DBE_TBF_GPUwise = [] # includes both new and old
+OTB_TBF_GPUwise = []
+
+# records start-times (multiple in some cases) and event-times
+# start-times are indicated with a record of -1
+DBE_dict_GPUwise = {}
+OTB_dict_GPUwise = {}
+clean_dict_GPUwise = {}  # nor DBE or OTB
+
+# records start-times GPU-wise for old-new analysis later on
+oldNew_dict_GPUwise = {}
[…]

### 2020-06-05 reviews/response.txt
some rephrasing

@@
 
+We have added more explantion to most figures, making them more
+self-contained and relevant to our conclusions.
+
 Reviewer 1:
 
-(C) Figure 2 is improved and corrected. The inventory dates are the same
+(C)
+Figure 2 is improved and corrected. The inventory dates are the same
 for old and new batches, except that new units begin to appear in 2016 and
@@
 
-(C) We have strengthened our explanations
+(C)
+We have strengthened our explanations in several sections and
+particularly around our airflow explanation in Fig.4. Both substantive
+improvements described above relate to this point, on one hand
+strengthening it with temperature data, and on the other hand also finding
+a more nuanced impact of scheduling.
+
+(D)
+Section 3: see changes in first paragraph.
+
+Reviewer 3:
+
+(C)
+Generalization: This analysis is already generating interest from
+Summit operations. A lesson learned is that Titan was not well
+instrumented for data collection - thus our difficulty in obtaining
+temperature data. Frontier will likely have even more
+instrumentation.
+
+Novelty in graphics: Figures 1 and 5 were specifically designed for
+this data using R's amazingly high-level and flexible ggplot2
+concepts. Figures 10-13 are standard in R's survival analysis
+packages.
+
+(D)
+
+Reviewer 4:
+
+(C)
+
+(D)

### 2020-06-05 dataclean.tex
some rephrasing

@@
 of recorded life in the raw records by simply dropping the overlapping
-life times. \rev{A detailed followup of a few overlaps found that they
-  are caused by incomplete inventories, where a node query times out,
-  creating the appearance that a unit is in two locations.}
+life times. \rev{A detailed followup of a few of these overlaps found
+  that they are caused by incomplete inventories, where for a moved
+  GPU a node query of times out, creating the appearance that a unit
+  is in two locations.}
 

### 2020-06-05 discussion.tex
Renamed the Discussion Section to Conclusions.
some rephrasing

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Discussion
+% Conclusions
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{\rev{Conclusions}}
 \label{section:discussion}
 
-The experienced unexpected reliability issues with the Titan supercomputer raise
-several questions and discussion points. First, there is the mitigation component
-that keeps the system operating at an acceptable level and possibly even restores
-some, if not all, of the lost capability. Second, there is the long-term planning
-component that addresses such scenarios in future-generation supercomputers.
+\rev{The failure rates of the old batch are not matched by the new
+  batch nor by experience at other facilities with the same type of
+  GPU components, making this an unexpected event that had
+  considerable impact on operations and on availability of the
+  system. On a positive but ironic note, the corrosion process in the
+  failing resistors made the GPUs act as a sensitive instrument for
+  obtaining long-term integrals of component heat dynamics and
+  provided lessons learned. Here we dicsuss the conclusions we can
+  draw from the experience on mitigation, which keeps the system
+  operating at an acceptable level and possibly even restores some of
+  the lost capability, and on long-term planning, which addresses such
+  scenarios in future-generation supercomputers.}
 

### 2020-06-05 tbf.tex

@@
   \begin{center}
-    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_GPUwise.pdf}
+    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_GPUwise_yrs_OldNew.pdf}
   \end{center}
-  \caption{Distribution of device-level MTBFs due to DBE and OTB failures across all GPUs during the
-entire lifetime of the machine. DBE and OTB events are separated for each GPU to distinguish between 
-the failure pattern of each event type. Data includes both old and new batch of GPUs with the new ones 
-mostly having MTBFs around or less than 3 years.}
+  \caption{\rev{Distribution of device-level MTBFs due to failures across all GPUs during the
+lifetime of the machine. The DBE and OTB events are separately plotted to distinguish between 
+the failure pattern of each event type. Data includes both old and new batch of GPUs with the new GPUs
+having MTBFs scattered between 0 and 3 years as highlighted by yellow highlights in the plot. 
+Each bin in the distribution represents almost 2 weeks.}}
   \label{fig:Device_MTBFs}
@@
 the insert time as the reference point, whereas, a simple difference is taken for subsequent
-failures. It can be noticed that MTBFs due to DBE and OTB failures of most GPUs are clustered around 
-25,000 hours or 2.8 years. This corresponds to the lifetime of most GPUs in the system after accounting 
+failures. \rev{It can be noticed that MTBFs due to DBE and OTB failures of old GPUs are clustered around 
+2.8 years.} This corresponds to the lifetime of most GPUs in the system after accounting 
 for relocations and replacements done in the machine. Almost all of the new batch of GPUs lie below 
-the average since most have a lifetime close to 3 years. We point out
-that the new batch had a much smaller number of failures than
-the old batch as we highlight in the next section.
+the average since most have a lifetime close to 3 years. \rev{We point out
+that the new batch had a much smaller number of failures (a total of 127 DBEs and OTBs) than
+the old batch (a total of 5320 DBEs and OTBs). With this high disproportionality, it is difficult
+to compare device-level MTBFs across the new and old batches. On the other hand, it indicates better 
+reliability of the new GPUs as highlighted in the next section.}
 
@@
 with OTB events causing a substantial decrease in OTB events.
-
-%\begin{figure}[bt]
-%  \begin{center}
-%    \includegraphics[width=\columnwidth]{figs/MTBF_CageWise.pdf}
-%  \end{center}
-%  \caption{MTBFs due to either DBEs or OTBs across the GPUs located in different cages.}
-%  \label{fig:CageWise_MTBFs}
-%\end{figure}
-
-%\fix{Figure~\ref{fig:CageWise_MTBFs} shows temperature effect clearly. Should we move this 
-%result to later in the paper?}
 
@@
 machine in operation. 
-%\fix{need to verify how long the replacement took}
-There is not definitive way to incorporate this unavailability of the
+There is no definitive way to incorporate this unavailability of the
 machine into MTBF analysis, unless we know how the down times were scheduled.
@@
   \begin{center}
-    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL.pdf}
+    \includegraphics[trim={0 1em 0 1em},clip,width=\columnwidth]{figs/MTBF_quaterly_sys_NewOldALL_newPart.pdf}
   \end{center}
@@
     old partitions of the machine after a substantial number of new GPUs
-were put in service.
+were put in service. \rev{The dashed line plotted with secondary y-axis shows 
+the size of new partition over time. 
 DBE and OTB events both determine the MTBF.
 Results highlight the difference in reliability for jobs using the new GPUs as compared to old GPUs,
-as well as now MTBF is driven by the weaker old partition.}
+as well as now MTBF is driven by the less reliable old partition despite occupying only a minority
+portion of the machine.}}
   \label{fig:MTBF_sys_NewOld}
@@
 \rev{Fig.}~\ref{fig:MTBF_sys_NewOld} shows the drastic difference in system MTBF measured in two hypothetical
-partitions of the machine starting from the time when the majority of GPU replacements have been completed. 
-The new partition of the machine has a 12X better MTBF than the older
-partition.
+partitions of the machine starting from the time when a major number of GPU replacements had been completed.
+\rev{The Fig.~\ref{fig:MTBF_sys_NewOld} also shows the proportion of the machine with new GPUs over time. 
+The majority of the machine had new GPUs installed starting in 2017-Q3.   
+This new partition of the machine has a 12X better MTBF than the older
+partition.}
 

### 2020-06-05 discussion.tex
tweak

@@
   system. On a positive but ironic note, the corrosion process in the
-  failing resistors made the GPUs act as a sensitive instrument for
-  obtaining long-term integrals of component heat dynamics and
-  provided lessons learned. Here we dicsuss the conclusions we can
-  draw from the experience on mitigation, which keeps the system
-  operating at an acceptable level and possibly even restores some of
-  the lost capability, and on long-term planning, which addresses such
-  scenarios in future-generation supercomputers.}
+  failing resistors made the specific GPU batch act as a sensitive
+  instrument for obtaining long-term integrals of component heat
+  dynamics and provided lessons learned. Here we dicsuss the
+  conclusions we can draw from the experience on mitigation, which
+  keeps the system operating at an acceptable level and possibly even
+  restores some of the lost capability, and on long-term planning,
+  which addresses such scenarios in future-generation supercomputers.}
 

### 2020-06-05 survival.tex
lustre heat?

@@
   sets is not likely as the evaporators (see Fig.~\ref{fig:layout})
-  are thermostat controlled.}
+  are thermostat controlled.} \fix{The file system, however. did not
+  have evaporator heat exchangers and relied on ambient room air for
+  heat transport, affecting room temperature on the right side of
+  Titan. !!Do we have room temp measurement data?!!}
 

### 2020-06-05 reviews/response.txt
lustre heat?
novelty notes

@@
 We have added more explantion to most figures, making them more
-self-contained and relevant to our conclusions.
+self-contained and relevant to our conclusions. We also made some
+small but useful tweaks in our narrative that is not highlited in blue
+so as to not distract from the more substantive changes responding to
+reviewer comments.
 
@@
 temperature data. Frontier will likely have even more
-instrumentation.
+instrumentation. We now note this in our conclusions.
 
-Novelty in graphics: Figures 1 and 5 were specifically designed for
-this data using R's amazingly high-level and flexible ggplot2
-concepts. Figures 10-13 are standard in R's survival analysis
-packages.
+Novelty: We have added notes about novelty in the visualizations and
+our analysis in the introduction.
 

### 2020-06-05 intro.tex
novel viz
novelty notes

@@
 data cleaning, checking, and lifetime summarization process in
-Sec.~\ref{section:dataclean}. A time between failures (TBF) analysis
-is given in Sec.~\ref{section:tbf}. Survival analysis \rev{(SA)} based on
+Sec.~\ref{section:dataclean}. \rev{Here we develop novel
+  visualizations of GPU life times and errors that were needed to
+  understand the data and our data processing decisions.} A time
+between failures (TBF) analysis is given in
+Sec.~\ref{section:tbf}. Survival analysis \rev{(SA)} based on
 Kaplan-Meier modeling (KM) and Cox regression modeling (CR) of
 relative hazard rates in various locations are in
-Sec.~\ref{section:survival}. The goals of a TBF analysis and of
-SA are different: TBF studies largely \rev{show} GPU
-reliability affects the machine, whereas SA studies largely how the
-machine affects GPU reliability.  Finally, we discuss our main
-conclusions and recommendations in Sec.~\ref{section:discussion}.
+Sec.~\ref{section:survival}. The goals of a TBF analysis and of SA are
+different: TBF studies largely \rev{show} GPU reliability affects the
+machine, whereas SA studies largely how the machine affects GPU
+reliability. \rev{The application of SA methods is novel in the HPC
+  reliability context. It has roots in biostatistics and epidemiology,
+  where it is used to discover causes of disease.} Finally, we discuss
+our main conclusions and recommendations in
+Sec.~\ref{section:discussion}.

### 2020-06-05 discussion.tex
tweak

--- 2020-06-05 discussion.tex
@@
   system. On a positive but ironic note, the corrosion process in the
-  failing resistors made the GPUs act as a sensitive instrument for
-  obtaining long-term integrals of component heat dynamics and
-  provided lessons learned. Here we dicsuss the conclusions we can
-  draw from the experience on mitigation, which keeps the system
-  operating at an acceptable level and possibly even restores some of
-  the lost capability, and on long-term planning, which addresses such
-  scenarios in future-generation supercomputers.}
+  failing resistors made the specific GPU batch act as sensitive
+  instruments for obtaining cummulative trend information from
+  component heat dynamics and provided lessons learned. Here we
+  dicsuss the conclusions we can draw from the experience on
+  mitigation, which keeps the system operating at an acceptable level
+  and possibly even restores some of the lost capability, and on
+  long-term planning, which addresses such scenarios in
+  future-generation supercomputers.}
 

### 2020-06-05 dataprep.tex
Corrected DBE description.
preparing for OLCF import

@@
 failure events were collected \rev{for this study}: Double Bit Error
-(DBE) and Off the Bus (OTB). \rev{DBE is a memory error detected by
-  parity bit checking, which can correct a single bit flip and detect
+(DBE) and Off the Bus (OTB). \rev{DBE is an error correcting code (ECC)
+  detection in GPU memory, which can correct a single bit flip and detect
   but not correct a double bit flip. OTB is the loss of host CPU
@@
   \end{minipage}
-  \caption{\rev{Titan cabinet (front view and side view with air flow)
-      and single blade layout (bottom left with {\em node} numbers)
-      and floor layout (bottom right with {\em col} and {\em row}
-      numbers). Cooling airflow is bottom to top within a cabinet,
-      taking ambient air in mostly at the foot and relasing it back
-      out at the top. Cooling fluid is passed through heat exchange
-      evaporators, below Cage 0 and above Cage 2. Ambient air in the
-      room is conditioned at nominal 75$^\circ$~F. The room air system
-      has a dozen intakes near the ceiling and dumps cool air under
-      the floor in several locations. A number of floor tiles are
-      perforated.  Yellow dot marks GPU location {\tt c17-4c1s3n1}.}}
+  \caption{\rev{Titan cabinet (front view and left side view with air
+      flow) and single blade layout (bottom left with {\em node}
+      numbers) and floor layout (bottom right with {\em col} and {\em
+        row} numbers). Cooling airflow is bottom to top within a
+      cabinet, taking ambient air in mostly at the foot and relasing
+      it back out at the top. Cooling fluid is passed through heat
+      exchange evaporators, below Cage 0 and above Cage 2. Ambient air
+      in the room is conditioned at nominal 75$^\circ$~F. The room air
+      system has a dozen intakes near the ceiling and dumps cool air
+      under the floor in several locations. A number of floor tiles
+      are perforated.  Yellow dot marks GPU location {\tt
+        c17-4c1s3n1}.}}
   \label{fig:layout}

### 2020-06-05 discussion.tex
preparing for OLCF import

--- 2020-06-05 discussion.tex
@@
   system. On a positive but ironic note, the corrosion process in the
-  failing resistors made the GPUs act as a sensitive instrument for
-  obtaining long-term integrals of component heat dynamics and
-  provided lessons learned. Here we dicsuss the conclusions we can
-  draw from the experience on mitigation, which keeps the system
-  operating at an acceptable level and possibly even restores some of
-  the lost capability, and on long-term planning, which addresses such
-  scenarios in future-generation supercomputers.}
+  failing resistors made the specific GPU batch act as sensitive
+  instruments for obtaining cummulative trend information from
+  component heat dynamics, giving the current analysis the strong
+  signals observed, and provided lessons learned. Here we dicsuss the
+  conclusions we can draw from the experience on mitigation, which
+  keeps the system operating at an acceptable level and possibly even
+  restores some of the lost capability, and on long-term planning,
+  which addresses such scenarios in future-generation supercomputers.}
 

### 2020-06-05 survival.tex
preparing for OLCF import

@@
   temperature, although consider their evidence preliminary due to
-  relatively low number of DBEs in 2014-1015 and high variability of
+  relatively low number of DBEs in 2014-2015 and high variability of
   $10^\circ$C to $15^\circ$C in their measured cage replicates.}

### 2020-08-25 code/TitanGPUmodel.Rmd
updated Rmd

@@
 ---
-title: "Titan GPU model"
-author: "GO"
+title: "Titan GPU models"
+author: "George Ostrouchov"
 output:
@@
 ---
-```{r setup, include=FALSE}
+
+R scripts to accompany SC20 paper "GPU Lifetimes on Titan Supercomputer: Survival Analysis and Reliability" by George Ostrouchov, Don Maxwell, Rizwan Ashraf, Christian Engelmann, Mallikarjun Shankar, and James Rogers.
+
+```{r setup}
 options(repos = c(CRAN = "http://cran.rstudio.com")) # set CRAN mirror
@@
 library(forestmodel)
+sessionInfo()
+```
 
+The above documents software versions and platform that was used to run the analysis.
+```{r}
 out_figs = "../figs/" # paper related figures will go here
@@
 service_nodes = paste0(data_dir, "titan.service.txt")
-source("TitanGPUsetup.R")
+
+source("TitanGPUsetup.R") ## define some additional functions
 Rmd.time = proc.time()
 ```
-`r`
-Read the data. Last inventory was run 01/20/2020 but Titan was turned off on 08/01/2019
+
+**Read the data**.
+
+Last inventory files processing was done on 01/20/2020 but Titan was turned off on 08/01/2019 so change these *remove* dates.
 ```{r}
 g_raw = fread(history_file, col.names = c("SN", "location", "insert", "remove"))
-g_raw$remove[stringr::str_starts(g_raw$remove, "01/20/2020")] =
-  "08/01/2019 20:07:33"
+lastrun_dates = stringr::str_starts(g_raw$remove, "01/20/2020")
+g_raw$remove[lastrun_dates] = "08/01/2019 20:07:33"
+cat("Total of", sum(lastrun_dates), "remove dates changed to lights-out date")
 cat("Total of", (unique_SN = length(unique(g_raw$SN))), "unique SN\n")
@@
 **Enable missing values**, and **create** some new variables.
-Variable *event* is "life" when both *insert* and *remove* are present, "life0" when *insert* = *remove*, otherwise it is whatever string is in *insert*, which is "DBE" or "Off The BUS" (renamed to "OTB").
+Variable *event* is "life" when both *insert* and *remove* are present, "life0" when *insert* = *remove*, otherwise it is whatever string is in *insert*, which is "DBE" or "Off The BUS" (re-coded to "OTB").
 **Fill in** (repeat) serial numbers and locations for all records.
@@
          SN_ok = grepl("[0-9]{13}", SN), # verify that SN is 13 numbers
-         location = zoo::na.locf(location), # fill in missing location (forward)
+         location = zoo::na.locf(location), # fill in missing location forward
          location_ok = grepl("c[0-9][0-9]?-[0-7]c[0-2]s[0-7]n[0-3]", location))
@@
 
-**List** bad SN and bad location raw data. **Remove** bad SN and bad location data. **Read and remove** service node data.
-```{r fig.width = 10}
+**List** bad SN and bad location raw data. **Remove** bad SN and bad location data. Cause is unknown but there are very few.
+```{r}
 good_date_remove = lubridate::mdy_hms("01/01/2014 00:00:00", tz = time_zone)
@@
 g_raw[!g_ev$location_ok, ] # list bad location records
+```
+
+**Read** service nodes and remove associated data.
+```{r}
 service = fread(service_nodes, col.names = c("node", "loc", "location", "type", "up", "run"))
@@
     "unique clean locations\n")
-top_remove_dates = sort(table(g_ev$remove), decreasing = TRUE)[1:25]
-trdat = top_remove_dates[order(names(top_remove_dates))]
-top_insert_dates = sort(table(g_ev$remove), decreasing = TRUE)[1:25]
-top_insert_dates[order(names(top_insert_dates))]
-trd_all = data.frame(ht = ifelse(as.numeric(trdat) < 2000, 2000, 
-                                 as.numeric(trdat)*0.95),
-                     date = as.POSIXct(names(trdat)))
-trd = trd_all[c(1:7, 20:25), ]
+```
 
+Make a chronology plot of Titan GPU life.
+Overlay yearly histogram over daily histogram.
+Yearly excludes data before *good_date_remove*.
+Both plots are restricted to "life" records only.
+(Note to reviewers: an earlier manuscript version included "life0" "DBE" "OTB" events in this chronology, artificially inflating GPU swap numbers, mostly in the first year.)
+```{r  fig.width = 10}
 ggp(ggplot(filter(g_ev %>% filter(event == "life"), remove < max(remove)), aes(remove)) + 
       geom_histogram(
-#        data = filter(g_ev, remove < max(remove)),
         data = filter(g_ev, remove >= good_date_remove, remove < max(remove)),
@@
     file = "chronology", width = 10, height = 3)
-```
-The next analysis looks at *life0* events, plots their frequency, and lists a sample of 10 such events in the raw input data.
-```{r}
-d2013a = as.POSIXlt.character("2014-05-30 00:00:00 EST")
-d2013b = as.POSIXlt.character("2014-06-23 00:00:00 EST")
-zeros = g_ev %>% filter(event == "life0" | event == "life", remove > d2013a, remove < d2013b)
-ggplot(zeros, aes(remove)) + 
-#      geom_histogram(data = zeros,
-#        mapping = aes(remove), binwidth = 60*60*24*182.5, alpha = 0.2,
-#        boundary = good_date_remove) +
-      geom_histogram(binwidth = 60*60, color = "blue") + theme_bw() +
-      theme(axis.title.x = element_blank()) + scale_x_datetime(breaks = "1 day") + ylim(c(0, 25))
-sn_zeros = zeros$SN[sample(length(zeros$SN), 10)]
-for(sn in sn_zeros) print(g_raw[(i = which(g_raw$SN == sn)):(i+10), ])
 ```
@@
 ```
-**Mark time overlaps within SN, then remove SN records that overlap and plot before-after**. 
+
+**Mark time overlaps within SN, then remove the overlap record and plot before-after**. 
 SN time overlaps are possible when an inventory is incomplete.
@@
 plot.life_ev(gc_life, gc_ev_nzg, SN, overlaps = TRUE, outs = FALSE, file = "overlap_SN")
+```
+
+Now the same SN after removing overlap records:
+```{r fig.height=10}
 gc_life = gc_life %>% filter(!overlap_rec)
@@
 ```
+
 For locations, **remove only the overlapping life** but keep the rest of the SN.
@@
 Blades were sometimes hot-swapped by CRAY and if this was during an inventory run, the results could be unpredictable.
-**Plot overlap at locations again after removing the offending life spans.*
 ```{r fig.height = 10}
 gc_life = mark.overlaps(gc_life, location)
-plot.life_ev(gc_life, gc_ev_nzg, location, overlaps = TRUE, outs = FALSE, file = "overlap_Loc")
+plot.life_ev(gc_life, gc_ev_nzg, location, overlaps = TRUE, outs = FALSE, 
+             file = "overlap_Loc")
+```
 
-for(i in which(g_raw$location == "c7-7c0s7n3")) print(g_raw[(i-5):i, ])
-for(i in which(g_raw$location == "c22-0c2s0n0")) print(g_raw[(i-5):i, ])
+Print raw data for a few overlap locations to understand reasons:
+```{r}
+for(i in which(g_raw$location == "c7-7c0s7n3")) print(g_raw[(i-5):(i+5), ])
+for(i in which(g_raw$location == "c22-0c2s0n0")) print(g_raw[(i-5):(i+5), ])
+```
+
+**Plot overlap at locations again after removing the offending life spans.**
+```{r fig.height = 10}
 gc_life = gc_life %>% filter(!overlap_rec)
-plot.life_ev(gc_life, gc_ev_nzg, location, overlaps = TRUE, outs = FALSE, file = "overlap_Loc_after")
+plot.life_ev(gc_life, gc_ev_nzg, location, overlaps = TRUE, outs = FALSE, 
+             file = "overlap_Loc_after")
 ```
@@
 ```
-**Inventory interval analysis**
+
+**Inventory interval analysis**. Use differences between sorted unique *insert* and *remove* dates of "life" records to determine inventory intervals.
 ```{r fig.width=10, fig.height=4}
@@
 life_remove = gc_life$remove # collect remove dates
-dates_dbe = sort(unique(filter(gc_ev_fail, event == "DBE")$remove)) # unique DBE dates
-dates_otb = sort(unique(filter(gc_ev_fail, event == "OTB")$remove)) # unique OTB dates
 dates_life = sort(unique(c(life_insert, life_remove))) # unique insert,remove
@@
   data.frame(days = as.numeric(diff(dates_life))/(60*60*24),
-             year = factor(as.character(year(dates_life[-1])))) %>% filter(days >= 1) %>% 
+             year = factor(as.character(year(dates_life[-1])))) %>% 
+  filter(days >= 1) %>% 
   mutate(days = factor(as.integer(floor(days)), levels = 1:56))
@@
 
+Seems longest interval without an inventory is 56 days!
+
+Next, sample a few SN and locations to display a sense of GPU life for the paper.
 ```{r}
-nsample = 88 # the max number for a full page pair of figures
+## set sampling for SN and Locations display
+nsample = 90 # the max number for a full page pair of figures
 set.seed(76513)  # for reproducibility!
@@
 ```
+
 **Plot a sample** of `r nsample` GPUs to check that things look right ...
@@
 
-Add a % uptime in a location histogram. Might be a good diagnostic
-
 Reduce to only "life" records and mark them with ending DBE, OTB, "out", or none events.
@@
 ```
+
 **Output** gc_full data frame to a CSV file for TBF analysis with Python code
@@
 ```
+
 **Message to reader of file:** All fields are character strings. Dates are with timezone. Missing values as "<NA>" strings.
+
+**Print raw records** for Figure 3:
+```{r}
+for(i in which(g_raw$SN == "0323812007945")) print(g_raw[(i-5):(i+5), ])
+for(i in which(g_raw$SN == "0325216047736")) print(g_raw[(i-5):(i+5), ])
+for(i in which(g_raw$SN == "0323812008856")) print(g_raw[(i-5):(i+5), ])
+```
 
@@
 
-**Separate max_loc** into components (col, row, cage, slot, node).
-Note that this is the place this *SN* was located the longest.
-This muddies the hazard analysis somewhat as several locations can be the "treatment" for a given *SN*.
-But the proportions of time at max_loc tend to be very high.
-Time-dependent covariates, see: https://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
+**Separate max_loc** into components (col, row, cage, slot, node) to be used as covariates in survival analysis.
+Note that this location is the place this *SN* was located the longest.
+This muddies the survival analysis somewhat as several locations can be the "treatment" for a given *SN*.
+But the proportions of time at max_loc tend to be very high (see later analysis below).
 ```{r}
@@
 ```
+
+This is the second data set we make available.
 **Output** gc_full to a CSV file
@@
 ```
+
 Get some stats for sanity check.
@@
 cat("           More than one:", sum(gc_summary_loc$dbe + gc_summary_loc$otb > 1, na.rm = TRUE), "\n")
-ggplot(gc_summary_loc, aes(x = time)) + geom_histogram(bins = 60) + ggtitle("GPU Life Times")
-ggplot(gc_summary_loc, aes(x = time)) + geom_histogram(bins = 60) + 
-  facet_wrap(~ batch) + ggtitle(paste("GPU Life Times - new and old"))
+```
+
+Nothing really remarkable in the stats.
+Censored + dead = total unique.
+Most with a DBE or OTB are marked dead.
+A few, 101 units (= 5190 - 5089) with DBEs or OTBs are still alive (that's just under 2%).
+551 units (about 10%) had more than one DBE or OTB.
+
+Look at life proportion at longest location:
+```{r}
 ggplot(gc_summary_loc, aes(x = time_max_loc)) + geom_histogram(bins = 200) +
@@
 
-Kaplan-Meier survival analysis gives the probability of survival beyond a given time.
-It is a nonparametric technique that makes no specific failure model assumptions, such as Weibull, Exponential, etc.
-It can also split the data into subpopulations and compute separate survival curves.
-
-If $T$ is taken to be the random variable of a GPU fail time, then its probability density function is $f(t)$ and its cumulative distribution function $F(t) = Pr\{T < t\}$ gives the probability that a GPU fails by duration $t$.
-The survival function is then
-$$S(t) = Pr\{T \geq t\} = 1 - F(t) = \int_t^\infty f(x)dx,$$ 
-which is the probability of being operational just before duration $t$.
-
-A sister concept to the survival function is the hazard function:
-$$H(t) = \frac{f(t)}{S(t)},$$
-which is what is sometimes called the "bathtub" curve in reliability.
-
-Units removed without a DBE are right-censored! They are not failures!
-
-Next, we consider splitting the GPU into groups - installed before 2016 (old) or later (new).
+The overwhelming majority of time is spent in the "longest" location so it is reasonable to assume the the longest location is the "treatment" of a GPU.
+But a time-dependent analysis is possible from the current data with some additional re-coding and following this time-dependent covariates technique: https://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf.
+We hope that others will undertake this more detailed analysis from our data.
 
 **SN categories**
-I tried disassembling the *SN* into digits and did not find any compelling results.
+We tried disassembling the *SN* into digits but did not find compelling relationship with life times except the known new-batch old-batch differences.
 
-Now for Kaplan-Meier analysis ...
+**Kaplan-Meier survival analysis** gives the probability of survival beyond a given time.
+It is a nonparametric technique that makes no specific failure model assumptions, such as Weibull, Exponential, etc.
+It can also split the data into subpopulations based on the covariates and compute separate survival curves.
 ```{r fig.height = 8}
@@
 ```
-Consider diagnostics for proportionalty.
-```{r}
-fit = coxph(Surv(years, dead, type = 'right') ~ row + col + cage + slot + node,
-            data = gc_summary_loc_o)
-## below seems to have an issue with RStudio: 
-## https://community.rstudio.com/t/ggcoxdiagnostics-warning-message/51867
-## ggcoxdiagnostics(fit, type = "schoenfeld", ox.scale = "time")
-test.ph = cox.zph(fit)
-ggcoxzph(test.ph)
-```
+
 The Cox proportional hazards (CPH) regression analys takes the hazard function above and a set of covariates $x$ and models $k$s GPU hazard as
@@
 
-For our comparison of hazards, we use as *covariates* all the unpacked location variables, indicators whether the GPU has been moved, and our notion of old and new GPUs based on first install date.
-Some GPUs experience time at more than one location but these are few.
-We use the longest period location in this analysis.
-We do separate analysis for old and new batches.
+First, consider some diagnostics for proportionalty:
+```{r fig.width = 10, fig.height = 11}
+fit = coxph(Surv(years, dead, type = 'right') ~ row + col + cage + slot + node,
+            data = gc_summary_loc_o)
+test.ph = cox.zph(fit)
+ggcoxzph(test.ph)
+```
+
+The fit lines are supposed to be horizontal for the proportionality assumption to be satisfied.
+They are all nearly horizontal.
+On the other hand, all the significance tests can detect a departure from horizontal because we have so much data.
+We interpret this that we have enough data to do time-dependent analysis (see comment on this above) but that the hazard proportionality is fairly good.
+To err on the conservative side, we use the results to make qualitative conclusions and are careful to not over-interpret the actual hazard levels.
+
+So now the CPH model:
 ```{r fig.width = 5, fig.height = 10}

### 2020-08-26 code/tbf-analyses/tbf_analyses.py
added paper information

@@ -1,14 +1,13 @@
-#### This code was used for time-between-failure analysis in the following paper (pending acceptance):
-## Authors list (TBD, hidden for review), ``GPU Lifetimes on Titan Supercomputer: Survival Analysis and Reliability,'' 
-## The International Conference for High Performance Computing, Networking, Storage, and Analysis (SC20),
-## Atlanta, GA, Nov. 15-20, 2020.
+#### This code was used for time-between-failure (TBF) analysis in the following paper:
+## George Ostrouchov, Don Maxwell, Rizwan A. Ashraf, Christian Engelmann, Mallikarjun Shankar, and James H. Rogers. 2020. GPU Lifetimes on Titan Supercomputer: Survival Analysis and Reliability. In Proceedings of the International Conference for High Performance Computing, Networking, Storage and Analysis (SC '20). Association for Computing Machinery, New York, NY, USA.
 ## ...
 #### If using this code or this work, please cite the above paper in your work.
 
 
 #### This code reads in a csv file with GPU failure data and does the following:
-## 1. GPU-wise mean-time-between-failure analyses (see Fig-6 in paper).
+## 1. GPU-wise mean-time-between-failure (MTBF) analyses (see Fig-6 in paper).
 ## 2. System-wide mean-time-between-failure analyses over lifetime (see Fig-7 through 9 in paper).
+## Note: currently, system-wide MTBF lifetime analysis is done over quarters (Jan-Mar, Apr-Jun, ...). It can be easily modified to be done over months or years. Moreover, GPU-wise MTBFs can be easily calculated based on locations in the machine (cages, columns, rows).  
 ####
 
 #### last modified: 2020-06-04 ##################################################################
@@ -32,19 +31,27 @@ def epoch(timestring):
 def convertToTime(timestring):  
     return time.strptime(timestring, "%Y-%m-%d %H:%M:%S")
 
-#### Parse failure data, while recording failure times ##########################################
+#### Parse failure data #########################################################################
[…]

### 2020-08-26 code/tbf-analyses/tbf_analyses.py
minor comment updates

@@ -1,5 +1,8 @@
 #### This code was used for time-between-failure (TBF) analysis in the following paper:
-## George Ostrouchov, Don Maxwell, Rizwan A. Ashraf, Christian Engelmann, Mallikarjun Shankar, and James H. Rogers. 2020. GPU Lifetimes on Titan Supercomputer: Survival Analysis and Reliability. In Proceedings of the International Conference for High Performance Computing, Networking, Storage and Analysis (SC '20). Association for Computing Machinery, New York, NY, USA.
+## George Ostrouchov, Don Maxwell, Rizwan A. Ashraf, Christian Engelmann, Mallikarjun Shankar, and James H. Rogers. 2020. 
+## GPU Lifetimes on Titan Supercomputer: Survival Analysis and Reliability. In Proceedings of the International Conference  
+## for High Performance Computing, Networking, Storage and Analysis (SC '20).  
+## Association for Computing Machinery, New York, NY, USA.
 ## ...
 #### If using this code or this work, please cite the above paper in your work.
 
@@ -7,7 +10,9 @@
 #### This code reads in a csv file with GPU failure data and does the following:
 ## 1. GPU-wise mean-time-between-failure (MTBF) analyses (see Fig-6 in paper).
 ## 2. System-wide mean-time-between-failure analyses over lifetime (see Fig-7 through 9 in paper).
-## Note: currently, system-wide MTBF lifetime analysis is done over quarters (Jan-Mar, Apr-Jun, ...). It can be easily modified to be done over months or years. Moreover, GPU-wise MTBFs can be easily calculated based on locations in the machine (cages, columns, rows).  
+## Note: currently, system-wide MTBF lifetime analysis is done over quarters (Jan-Mar, Apr-Jun, ...). 
+## The code can be easily modified to perform MTBF calculations over months or years. 
+## Moreover, GPU-wise MTBFs can be easily calculated based on locations in the machine (cages, columns).  
 ####
 
 #### last modified: 2020-06-04 ##################################################################

### 2020-08-27 code/tbf-analyses/tbf_analyses.py
added some references to the paper

@@ -16,6 +16,8 @@
 ####
 
 #### last modified: 2020-06-04 ##################################################################
+## author: Rizwan Ashraf
+## email: rizwan.ashraf@knights.ucf.edu
 
 #### package imports ############################################################################
 ## for data parsing 
@@ -80,7 +82,8 @@ bad_data_repeat = []
 DBE_count = 0
 OTB_count = 0
 
-### All data is parsed in the loop below.
+### All data is parsed in the loop below. 
+### For more info on data collection, preprocessing and cleaning, see paper, Sections III and IV.
 ## csv file: row[0] serial number, row[1] location, row[2] insert datetime, row[3] remove datetime, 
 ##           row[4] duration (seconds between insert and remove), 
 ##           row[5] indicates whether the GPU was seen after remove (true/false), 
@@ -927,6 +930,7 @@ plt.tight_layout()
 plt.savefig('../../figs/NumFailures_Quarterly_newOld.pdf', dpi=600)
 
 #### write bad serial numbers ################################################################
+## for more info: see paper, Section IV, page 4.
 MyFile=open('bad_serials.dat','w')
 MyFile.write('# no record for loc insert found for following GPU Serial Numbers:\n')
 for element in bad_data_serials_set:
